---
output:
  pdf_document: default
  html_document: default
---
# (PART) Models with Differential Equations {-} 

# Models of rates with data  {#intro-01}


## Rates of change in the world: a model is born

The focus of this textbook is understanding _rates of change_ and how you can apply them to model real-world phenomena. Additionally, this textbook focuses on _using_ equations with data, building both your competence and confidence to construct a mathematical model from data and a context. 

Perhaps you analyzed rates of change in calculus course when answering the following types of questions:


- If $y = xe^{-x}$, what is the derivative function $f'(x)$?
- What is the equation of the tangent line to $y=x^{3}-x$ at $a=1$?
- Where is the graph of $\sin(x)$ increasing at an increasing rate?
- What is the largest area that can be enclosed with 100 feet of fencing, with one side being along a wall?
- If you release a ball from the top of a skyscraper 500 meters above the ground, what is its speed when it impacts the ground?

The first three questions do not appear to be connected in a real-world context - but the last two questions *do* have some context.  For the fencing problem, perhaps a person raises chickens and wants to care for their well-being, with a rectangular pen more aesthetically pleasing than a circular pen.  In the last example the ball falling off the skyscraper assumes that acceleration of the ball is constant.

The context may reveal underlying assumptions or physical principles, which are the starting point to build a mathematical model. For the chicken coop problem the next step is to use the assumed geometry (rectangle) with the 100 feet of fencing to develop a function for the area as a function of the length of one of the sides of the pen. For the ball problem, the velocity (or the antiderivative of acceleration) can be found, from which the position function can be calculated through antidifferentiation.

Let's say we have observational data and several different (perhaps conflicting) assumptions about the context at hand, and these assumptions describe models that involve rates of change. Which model is the best one to approximate the data?  The short answer: it depends. To understand why, let's take a look at a problem in context.


## Modeling in context: the spread of a disease
Consider the data in Figure \@ref(fig:sierra-leone-01), which come from an [Ebola outbreak](https://www.cdc.gov/vhf/ebola/history/2014-2016-outbreak/index.html) in Sierra Leone in 2014. Ebola is a fatal disease so we can also consider the vertical axis in Figure \@ref(fig:sierra-leone-01) to represent total *infections* due to Ebola.

```{r sierra-leone-01,warning=FALSE,message=FALSE,echo=FALSE,fig.cap='An Ebola outbreak in Sierra Leone', fig.width=4,fig.height=3}
ebola <- read_csv("data/ebola.csv") %>%
  mutate(date = mdy(`WHO report date`), `WHO report date` = NULL) %>%
  select(-c(1, 3, 5)) %>%
  gather(
    key = nation, value = deaths,
    1:3
  ) %>%
  mutate(monitor_days = as.numeric(date) - min(as.numeric(date)) + 1)

ebola %>%
  filter(nation == "Deaths Sierra Leone") %>%
  ggplot() +
  geom_point(aes(x = monitor_days, y = deaths), size = 2) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 14),
    axis.title.x = element_text(size = 18),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.y = element_text(size = 18)
  ) +
  labs(x = "Monitoring Days", y = "Ebola deaths")
```

Constructing a model from disease dynamics is part of the field of [mathematical epidemiology](https://en.wikipedia.org/wiki/Mathematical_modelling_of_infectious_disease). How we construct a mathematical model of the spread of this outbreak largely depends on the assumptions underlying the dynamics of the disease, such as considering the rate of spread of Ebola. For our purposes here we focus on the population level (person to person) spread of Ebola.  Other types of models could focus on the immune response in a single person - perhaps with a goal to design effective types of treatments to reduce the severity of infection.

Here are three initial assumptions one can make regarding the spread of Ebola:

1. The infection rate is proportional to the number of people infected.
2. The infection rate is proportional to the number of people **not** infected.
3. The infection rate is proportional to the number of infected people coming into contact with those not infected.

Let's see what each of these mathematical models would look like if we wrote down a mathematical equation.  Since we are discussing *rates* of infection, this means we will need a *rate of change* or derivative.  Let's use the letter $I$ to represent the number of people that are infected.

### Model 1: Infection rate proportional to number infected.
The first assumption states that the infection rate is proportional to the number of people infected.  Translated into an equation this would be the following:

\begin{equation}
\frac{dI}{dt} = kI (\#eq:00infected)
\end{equation}

In Equation \@ref(eq:00infected) $k$ can be thought of as a proportionality constant, with units of time$^{-1}$ for consistency.  Equation \@ref(eq:00infected) is an example of a *differential equation*, which is just a mathematical equation with rates of change.

The *solution* to a differential equation is a function $I(t)$. ^[You may be used to working with *algebraic equations* (e.g. solve $x^{2}-4=0$ for $x$) rather than differential equations.  For algebraic equations the solution can be points (for our example, $x=\pm2$).]  When we ``solve'' a differential equation we determine the family of functions consistent with our rate equation.  There are a lot of techniques we can use to do that, and we will examine a few later.

Back to this proportionality constant $k$ - another term for it is a *parameter*.  We can always try to solve an equation without specifying the parameter - and then if we wanted to plot a solution the parameter would also be specified.  In some situations we may not be as concerned with the particular _value_ of the parameter but rather its influence on the long-term behavior of the system (this is one aspect of [bifurcation theory](https://en.wikipedia.org/wiki/Bifurcation_theory)).  Otherwise we can use the collected data shown above with the given model to determine the value for $k$.  This combination of a mathematical model with data is called *data assimilation* or *model-data fusion*.


Before we think about possible solutions to our differential equation, let's try to reason if the first model would be plausible.  The first model assumptions states that the rate of change (the amount of increase) gets larger with the more sick people there are.  While this may seem reasonable initially, it could grows quickly unreasonable when the pandemic spreads.  In the case of Ebola or any other infectious disease, stringent public health measures would be enacted if the number of people infected become too large^[The COVID-19 pandemic that began in 2020 is an example of the heroic efforts of public health officials.].  Following public health measures we would expect that the rate of infection would decrease and the number of deaths to slow.  So perhaps the second model might be a little more plausible.  At some point the number of people who are *not* sick will reach zero, making the rate of infection be zero (or no increase).

### Model 2:  Infection rate proportional to number NOT infected.
In this description notice how we are talking about people who are sick (which we have denoted as $I$) and people who are *not* sick.  This looks like we might need to introduce another variable for the ``not sick" people, which we will call $S$, or susceptible.  So the differential equation we would write down would be:

\begin{equation} 
\frac{dI}{dt} = kS  (\#eq:01notinfected)
\end{equation}

We are still using the parameter $k$ as with the previous model.  Also note we introduced the second variable $S$ is in Equation \@ref(eq:01notinfected).  Because we have introduced another variable $S$ we should also include a differential equation for how $S$ changes as well.  One way that we can do this is by considering our entire population as consisting of two groups of people: $S$ and $I$.  Infection brings someone over from $S$ to $I$, which we have in Figure \@ref(fig:initial-si):

```{tikz, initial-si,warning=FALSE,message=FALSE,echo=FALSE,fig.align="center",fig.cap="Schematic diagram for Model 1"}
\tikzstyle{vspecies}=[rectangle,minimum size=0.5cm,draw=black]
\begin{tikzpicture}[auto, outer sep=1pt, node distance=2cm]

\node [vspecies] (S) {S} ;
\node [vspecies, right of = S] (I) {I} ;
\draw [->,thick] (S) --  node {\small{$k$}} (I) ;

\end{tikzpicture}
```

There are three reasons why I like to use diagrams like Figure \@ref(fig:initial-si): (1) they help organize my thinking about a mathematical model (2) any assumed parameters ($k$) are listed, and (3) they help me to see that rates can be conserved.  If I enter into the box for $I$, then someone is leaving $S$.  So then the rate of change equation for $S$ is $\displaystyle \frac{dS}{dt} = -kS$. Together with the differentail equation for $I$ I have the following:

\begin{equation}
\begin{split}
\frac{dS}{dt} &= -kS \\ 
\frac{dI}{dt} &= kS
\end{split}
(\#eq:02coupled)
\end{equation}


Equation \@ref(eq:02coupled) is an example of a *coupled differential equation*.  In order to ``solve" the system we need to determine functions for $S$ and $I$.  This coupled set of equations looks a little clunky, but there is something interest going on if we add the rates $\displaystyle \frac{dS}{dt}$ and $\displaystyle \frac{dI}{dt}$ together.  Algebraically we have:

\begin{equation} 
\frac{dS}{dt} +  \frac{dI}{dt} = \frac{d(S+I)}{dt} = 0 (\#eq:single-02)
\end{equation}

Recall from calculus that if a rate of change equals zero then the function is constant.  In this case, the variable $S+I$ is constant, or we can also call $S+I=N$, the number of people in the population.  This means that $S=N-I$, so we can re-write Equation \@ref(eq:02coupled) with a single equation:

\begin{equation}
\frac{dI}{dt} = k(N-I) (\#eq:02single)
\end{equation}


This second model does have some limiting behavior to this model as well.  As the number of infected people reaches $N$ (the total population size), the values of $\displaystyle \frac{dI}{dt}$ approaches zero, meaning $I$ doesn't change.  There is one caveat to this - if there are no infected people around ($I=0$) *the disease can still be transmitted*, which might make not good biological sense. 

### Model 3: Infection rate proportional to infected meeting not infected.
The third model rectifies some of the shortcomings of the second model (which rectified the shortcomings of the first model).  The third model states that the rate of infection is due to those who are sick infecting those who are not sick.  This would sort of scenario would also make some sense, as it focuses on the *transmission* of the disease between susceptibles and infected people. So if nobody is sick ($I=0$) then the disease is not spread.  Likewise if there are no susceptibles ($S=0$), the disease is not spread as well.

In this case the diagram outlining the third model looks something like this:

```{tikz, logistic-scheme,warning=FALSE,message=FALSE,echo=FALSE,fig.align="center",fig.cap = "Schematic diagram for Model 3"}
\tikzstyle{vspecies}=[rectangle,minimum size=0.5cm,draw=black]
\begin{tikzpicture}[auto, outer sep=1pt, node distance=2cm]

\node [vspecies] (S) {S} ;
\node [vspecies, right of = S] (I) {I} ;
\draw [->,thick] (S) --  node {\small{$kS$}} (I) ;

\end{tikzpicture}
```

Notice how in Figure \@ref(fig:logistic-scheme) there is an additional $S$ associated with $k$ to show how the rate of infection depends on $S$.  The differential equations that describe the scenario outlined in Figure \@ref(fig:logistic-scheme) are the following:

\begin{align*}
\frac{dS}{dt} &= -kSI \\
\frac{dI}{dt} &= kSI
\end{align*}

Just like before for Model 2 we can combine the two equations to yield a single differential equation:

\begin{equation}
\frac{dI}{dt} = k\cdot I \cdot (N-I)
\end{equation}

Look's pretty similar to Model 2, doesn't it?  In this case notice the variable $I$ outside the expression, which this seems to be appropriate - if $I=0$, then there is no increase in infection.  If $I=N$ (the total population size) then there is no increase in the infection.


Let's compare these two different models graphically. For both models let's plot $\displaystyle \frac{dI}{dt}$ versus $I$, and just so we can plot let's $k=1$ and $N=10$ respectively.  Plots of these functions are shown in Figure \@ref(fig:threeRates-01).

```{r threeRates-01,message=FALSE,echo=FALSE,fig.cap='Comparing rates of change for three models',warning=FALSE,fig.width=4,fig.height=3}
x <- seq(0, 12, length.out = 100)
n <- 10

data.frame(x, m1 = x, m2 = (n - x), m3 = x * (n - x)) %>%
  gather(key = model, value = rate, m1, m2, m3) %>%
  ggplot(aes(x = x, y = rate, color = model)) +
  geom_line(size = 1) +
  xlab("I") +
  ylab("dI/dt") +
  scale_color_discrete(labels = c("Model 1", "Model 2", "Model 3"), name = NULL) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 14),
    axis.title.x = element_text(size = 18),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.y = element_text(size = 18)
  ) +
  geom_hline(yintercept = 0)
```

Figure \@ref(fig:threeRates-01) has a lot to unpack, but we can use some of our understanding of rates of change in calculus to compare the three models. Notice how the sign of $\displaystyle \frac{dI}{dt}$ is always positive for Model 1, indicating that the solution ($I$) is always increasing.  For Models 2 and 3, $\displaystyle \frac{dI}{dt}$ equals zero when $I=10$, which also is the value for $N$  After that case, $\displaystyle \frac{dI}{dt}$ turns negative, meaning that $I$ is decreasing.


In summary, examining the graphs of the rates can tell a lot about the *qualitative behavior* of a solution to a differential equation even without the solution.

## Model solutions
Let's return back to possible solutions (in this case formulas for $I(t)$) for our models.  Usually a differential equation also has a starting or an initial value (typically at $t=0$) that actualizes the solution.  When we state a differential equation with a starting value we have an **initial value problem**.  We will represent that initial value as $I(0)=I_{0}$, where could be considered another parameter.

With that assumption, we can (and will solve later!) the following solutions for these models:

\begin{align*}
\mbox{ Model 1 (Exponential): } & I(t) = I_{0}e^{kt} \\
\mbox{Model 2 (Saturating): } & I(t) = N-(N-I_{0})e^{-kt} \\
\mbox{Model 3 (Logistic): } & I(t) = \frac{N \cdot I_{0} }{I_{0}+(N-I_{0})e^{-kt}}
\end{align*}

Notice how I assigned the names to each model (Exponential, Saturating, and Logistic).  That may not mean much at the moment, but Figure \@ref(fig:three-soln) plots the three functions $I(t)$ together when $I_{0}=5$, $k=0.03$, and $N=4000$.

```{r three-soln,message=FALSE,echo=FALSE,fig.cap='Three models compared',fig.width=4,fig.height=3,message=FALSE,warning=FALSE}
x <- seq(0, 600, length.out = 100)
n <- 4000
k <- 0.03
i0 <- 5

data.frame(x, m1 = i0 * exp(k * x), m2 = n - (n - i0) * exp(-k * x), m3 = n * i0 / (i0 + (n - i0) * exp(-k * x))) %>%
  gather(key = model, value = rate, m1, m2, m3) %>%
  ggplot(aes(x = x, y = rate, color = model)) +
  geom_line(size = 2) +
  xlab("t") +
  ylab("I") +
  scale_color_discrete(labels = c("Model 1", "Model 2", "Model 3"), name = NULL) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 14),
    axis.title.x = element_text(size = 18),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.y = element_text(size = 18)
  ) +
  ylim(c(0, 4010))
```

Notice how in Figure \@ref(fig:three-soln) Model 1 increases quickly - it actually grows without bound off the chart!  Model 2 and Model 3 have saturating behavior, but it looks like Model 3 might be the one that actually captures the trend of the data.  



## Which model is best?
All three of these scenarios describe different modeling scenarios. With the saturating and logistic models (Models 2 and 3) we have some limiting behavior the possibility that the the rate of infecion slows.  Of the two models, which one is the _best_ one?  Here could be some possible criteria we could evaluate:

- Do the model outputs match the data?
- (For timeseries data) are the trends accurately represented?
- Is the model easy to use?
- How will model outputs compare with newly collected measurements?
- Regarding model complexity - how many equations do we have?
- Are the number of model parameters too few or to many?


We will address several of these critera later on in this textbook when we discuss *model selection* (Section \@ref(information-criteria-14)).  Model selection is one key part of the modeling hypothesis - where we investigate the implications of a particular model analyzed.  If we don't do this, we don't have an opportunity to test out what is plausible and what is believeable in our models.


## Start here
In summary, it turns out that even with some initial assumptions we can very quickly build up a mathematical model to explain data.  Even with these first steps we have a lot more to uncover:

- How would you determine the parameters $k$ and $N$ with the collected data?
- Are there other more complicated models?
- What techniques are used to determine the formulas $I(t)$?
- Are there other numerical techniques to approximate the solution $I(t)$?
- What happens to our solutions when the parameters $k$ and $N$ change?
- What happens to our solutions when the number of infected people change randomly for some reason?

We will study answers to these questions and more.  Let's get started!

\newpage 

## Exercises

```{exercise, label="plot-soln"} 
Solutions to an outbreak model of the flu are the following:
  \begin{align*}
\mbox{Saturating model: } & I(t) = 3000-(2990)e^{-.1t} \\
\mbox{Logistic model: } &  I(t) = \frac{30000 }{10+(2990)e^{-.15t}},
\end{align*}

where $t$ is in days. Use these two functions to answer the following questions:
  


a. Plot the saturating and logistic models when $0 \leq t \leq 100$.  
b. For both models, how would you describe the growth of the outbreak as $t$ increases?  How many people will be infected overall?  
c. Finally, for both models evaluate $\lim_{t \rightarrow \infty} I(t)$. How do these results compare to values found on your graph?

```
&nbsp;


```{r, label="liberia-01",warning=FALSE,message=FALSE,echo=FALSE,fig.cap='An Ebola outbreak in Liberia in 2014',fig.width=4,fig.height=3,fig.pos='center'}
ebola <- read_csv("data/ebola.csv") %>%
  mutate(date = mdy(`WHO report date`), `WHO report date` = NULL) %>%
  select(-c(1, 3, 5)) %>%
  gather(
    key = nation, value = deaths,
    1:3
  ) %>%
  mutate(monitor_days = as.numeric(date) - min(as.numeric(date)) + 1)

ebola %>%
  filter(nation == "Deaths Liberia") %>%
  ggplot() +
  geom_point(aes(x = monitor_days, y = deaths), size = 2) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 14),
    axis.title.x = element_text(size = 18),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.y = element_text(size = 18)
  ) +
  labs(x = "Monitoring Days", y = "Ebola deaths")
```

&nbsp;
```{exercise} 
Figure \@ref(fig:liberia-01) shows the Ebola outbreak for the country of Liberia in 2014.
If we were to apply the logistic model (Model 3) based on this graphic what would be your estimate for $N$?
```
&nbsp;


```{exercise}
The general solution for the saturating and the logistic models are:
    \begin{align*}
\mbox{Saturating model: } & I(t) = N-(N-I_{0})e^{-kt} \\
\mbox{Logistic model: } &  I(t) = \frac{N \cdot I_{0} }{I_{0}+(N-I_{0})e^{-kt}},
\end{align*}
where $I_{0}$ is the initial number of people infected and $N$ is the overall population size. Using the functions from Exercise \@ref(exr:plot-soln) for both models, what are $N$ and $I_{0}$?
```
&nbsp;

```{exercise}
The general solution for the saturating and the logistic models are:
    \begin{align*}
\mbox{Saturating model: } & I(t) = N-(N-I_{0})e^{-kt} \\
\mbox{Logistic model: } &  I(t) = \frac{N \cdot I_{0} }{I_{0}+(N-I_{0})e^{-kt}},
\end{align*}
where $I_{0}$ is the initial number of people infected and $N$ is the overall population size. For both models carefully evaluate the limits to show $\lim_{t \rightarrow \infty} I(t)=N$. How do your limiting values these compare to the steady-state values you found for Models 2 and 3 in Figure \@ref(fig:three-soln), where $N=4000$?
```
&nbsp;

 <!-- Sethi model for advertising -->
```{exercise}
A model that describes the growth of sales of a product in response to advertising is the following:
  $$\frac{dS}{dt} = .55\sqrt{1-S}-S, $$
  where $S$ is the product's share of the market (scaled between 0 and 1).  Use this information to answer the following questions:
  

a. Make a plot of the function $f(S)=.55\sqrt{1-S}-S$. for $0 \leq S \leq 1$.
b. Interpret your plot to predict when the market share will be increasing and decreasing.  At what value is $\displaystyle \frac{dS}{dt}=0$?  (This is called the *steady-state* value.).
c.  A second campaign is has the following differential equation:
  $$\frac{dS}{dt} = .2\sqrt{1-S}-S $$
d. What is the steady-state value and how does it compare to the previous one?

```
&nbsp;

```{exercise}
A more general form of the advertising model is

\begin{equation}
\frac{dS}{dt} = r\sqrt{1-S}-S,
\end{equation}

where $S$ is the product's share of the market (scaled between 0 and 1). The parameter $r$ is related to the effectiveness of the advertising (between 0 and 1).
  

a. Solve $\displaystyle \frac{dS}{dt} = r\sqrt{1-S}-S$ for the steady state value (where $\displaystyle \frac{dS}{dt}=0$).  Your final answer should be expressed as a function $S(r)$.
b. Make a plot of the steady state value as a function of $r$, where $0 \leq r \leq 1$.
c. Based on your plot, what can you conclude about the steady state value as the effectiveness of the advertising increases?

  
```
&nbsp;

```{exercise}
A common saying is "You are what you eat."  This saying is mostly true and can be related in a mathematical model! Here's how: an equation that relates a consumer's nutrient content (denoted as $y$) to the nutrient content of food (denoted as $x$) is given by:
\begin{equation}
 y = c x^{1/\theta},
\end{equation}
where $\theta \geq 1$ and $c$ are both constants is a constant. Units on $x$ and $y$ are expressed as a proportion of a given nutrient (such as nitrogen or carbon). Let's start with an example when $c=1$ and $\theta = 1$.  Our function then is $y=x$.  In this case the point $(0.05,0.05)$ would say that if an animal ate food that was 5% nitrogen, their body composition would be 5% as well.


a.  Now assume that $c=1$. How does the nutrient content of the consumer compare to the food when $\theta=2$?  Draw a sample curve and interpret it, contrasting it to when $\theta = 1$.
b. Now assume that $c=1$. How does the nutrient content of the consumer compare to the food when $\theta=5$?  Draw a sample curve and interpret it, contrasting this curve to the previous two.
c. What do you think will happen when $\theta \rightarrow \infty$? Draw some sample curves to help illustrate your findings.

```
&nbsp;

```{exercise}
A model for the outbreak of a cold virus assumes that the rate people get infected is proportional to infected people contacting susceptible people, as with Model 3 (the Logistic model).  However people who are infected can also recover and become susceptible again with rate $\alpha$.  Construct a diagram similar Model 3 for this scenario and also write down what you think the system of differential equations would be.
```
&nbsp;


```{exercise}
A model for the outbreak of the flu assumes that the rate people get infected is proportional to infected people contacting susceptible people, as in Model 3.  However people also account for recovering from the flu, denoted with the variable $R$.  Assume that the rate of recovery is proportional to the number of infected people with parameter $\beta$.  Construct a diagram similar to Figure \@ref(fig:logistic-scheme) for this scenario and also write down what you think the system of differential equations would be.
```
&nbsp;
 <!-- Van den Berg page 19 -->
```{exercise}
Organisms that live in a saline environment biochemically maintain the amount of salt in their blood stream.  An equation that represents the level of $S$ in the blood is the following:
  
$$\frac{dS}{dt} = I + p \cdot (W - S), $$

where the parameter $I$ represents the active uptake of salt, $p$ is the permeability of the skin, and $W$ is the salinity in the water. Use this information to answer the following questions:
  

a. What is that value of $S$ at *steady state*, or when $\displaystyle \frac{dS}{dt} = 0$?  Your final answer should be a function $S(I,p,W)$.
b. Witn the steady state solution, Use your what parameters ($I$, $p$, or $W$) cause the steady state value $S$ to increase?

```
&nbsp;


<!-- From LW -->
```{exercise}
The immigration rate of bird species (species per time) from a mainland to an offshore island is $I_{m} \cdot (1-S/P)$, where $I_{m}$ is the maximum immigration rate, $P$ is the size of the source pool of species on the mainland, and $S$ is the number of species already occupying the island.  Additionally the extinction rate is $E \cdot S / P$, where $E$ is the maximum extinction rate.  The growth rate of the number of species on the island is the immigration rate minus the extinction rate.


a. Make representative plots of the of the immigration and the extinction rates as a function of $S$.  You may set $I_{m}$, $P$, and $E$ all equal to 1.
b. Determine the number of species for which the net growth rate is zero, or the number of species is in equilibrium.  Express your answer as $S$ as a function of $I_{m}$, $P$, and $E$.
c. Suppose that two islands of the same size are at different distances from the mainland. Birds arrive from the source pool and they have the same extinction rate on each island.  However the maximum immigration rate is larger for the island farther away.  Which of the two islands will have the larger number of species at equilibrium?

```
&nbsp;

```{exercise}
This problem relates to animal size and volume. Assume that an animal assimilates nutrients at a rate $R$ proportional to its surface area.  Also assume that it uses nutrients at a rate proportional to its volume. You may assume that the size of the animal is implicitly a function of the nutrient intake and usage. Determine the size of the animal if its intake and use rates were in balance (meaning $R$ is set to zero), assuming the animal is the following shapes:


a. A sphere (assume size is measured with radius $r$) *Note:* first determine the geometric formulas for surface area and volume.
b. A cube (assume size is measured with length $l$)


*Hint:* For both of these problems your goal is to determine a numeric value of $r$ and $l$.
```



