# Euler's method
The focus of this section is on *approximation* of solutions to a differential equation via a numerical method.  Typically a first numerical methods one might learn to tackle this problem is *Euler's method*, which is so fundamental it was popularized in the movie [Hidden Figures.](https://www.youtube.com/watch?v=v-pbGAts_Fg)


The way we are going to do this is through expansion of the idea of a locally linear approximation to the tangent line.  Let's start with an example.

## A Linear Approximation
```{example,name="The flu bug"}
The rate of change of the flu through a population is given by the number of people infected $t$ days after infection is, $$\displaystyle \frac{dI}{dt} = 3e^{-.025t}. $$  Assuming that $I(0)=10$, what is a locally linear approximation to this infection?  Second, what is the approximation for $I(1)$?

```
&nbsp;

```{remark}
In order to solve this problem, we know that the locally linear approximation is to $I(t)$ at $t=0$ is $L(t) = I(0) + I'(0) \cdot (t-0)$.  Here, $I(0)=10$ and $I'(0)=3$, so $$L(t) = 10 +3t$$.  Using $L(t) \approx I(t)$, we have $L(1)=10 + 3 = 13$.
```

&nbsp;

Notice we used two pieces of information: the (given) value of the function at $t=0$ and the estimate of the derivative from the rate of change.

This example is helpful because we can compare this to the actual solution, which is $I(t) = 130-120e^{-.025t}$ (you should verify this is the case by differentation).  When we compare the two, we have the following table:

$t$ | Euler's Method | Actual Solution
------------- | ------------- | -------------
0 | 10 | 10
1 | 13 | 12.96


Not too bad, huh? Our approximation at $L(1)$ is an *overestimate*, mainly because the actual solution is concave down, but it isn't that far off.


Let's build this solution out a little more by computing the rate of change at $t=1$, assuming that thirteen people is a pretty close estimate of the infection at that point?  What we would do is to build another linear approximation using our derivative equation. So the locally linear approximation is to $I(t)$ at $t=1$ is 

$$ L(t) = I(1) + I'(1) \cdot (t-1) $$.  

Here, $I(1)=13$ and $I'(1)=2.92$, and $L(t) = 13 +2.92(t-1).$

Assuming that $L(t) \approx I(t)$, we can evaluate $L(t)$ at $t=2$ *as an approximation* for $I(2)$: have $L(2)=13 + 2.92 = 15.92$.  Comparing this to the actual solution at $t=2$, we have $I(2)=15.85$.  Again, not too bad of a solution.

We can continue to build out the solution from there.  Figure \@ref(fig:eulers_ver1) shows what we would have for a solution if we continued to build out this approach:


```{r eulers_ver1,warning=FALSE,message=FALSE,echo=FALSE,fig.cap='Approximation of a solution using local linearity', fig.width=4,fig.height=3}
tibble(t=c(0,1,2),
       i_approx=c(10,13,15.92),
       i_soln=c(10,12.96,15.85)) %>%
  pivot_longer(cols=c("i_approx","i_soln"),names_to="solution",values_to="value") %>%
  ggplot(aes(x=t,y=value,color=solution,shape=solution)) +
  geom_point(size=2) +
  geom_line() +
  theme(legend.position="bottom",
        legend.text=element_text(size=14),
        axis.title.x=element_text(size=18),
        axis.text.x=element_text(size=10),
        axis.text.y=element_text(size=10),
        axis.title.y=element_text(size=18)) +
  labs(x = "Time",y = "Solutions") +
  theme(legend.title=element_blank()) +
  scale_shape_discrete(labels=c("Approximation", "Solution")) +
  scale_color_discrete(labels=c("Approximation", "Solution"))
```

When you plot them they do look indistinguishable from each other by eye.  It looks like we are onto something here!

## Defining an Algorithm
Here would be an algorithm that would describe our process to determine a solution to a differential equation:

* Determine the locally linear approximation at a given point.
* Forecast out to another time value.
* Repeat the locally linear approximation.

If we continue on in this way, let's take a look at how our approximation would do after several days:

```{r warning=FALSE,message=FALSE,echo=FALSE}

dt <- 1
t <- seq(0,100,by=dt)
i_soln <- 130-120*exp(-.025*t)
didt = 3*exp(-0.025*t)  # We are multiplying by dt
i_approx <- array(0,dim=length(t))
i_approx[1] <- 10
for(i in 2:length(t)) {
  i_approx[i] <- i_approx[i-1] + didt[i-1]*dt
}
```


$t$ | Euler's Method | Actual Solution
------------- | ------------- | -------------
90 | `r round(i_approx[90],1)` | `r round(i_soln[90],1)`
95 |  `r round(i_approx[95],1)` | `r round(i_soln[95],1)`

Now it seems that our approximation isn't so accurate as time goes on. What if we updated the infection rate every half day?  I know this means that we would be doing additional work (more iterations), but taking smaller timesteps goes hand in hand with more accurate solutions.  Let's start out smaller with the first few timesteps:

$t$ | $I$ | $\displaystyle \frac{dI}{dt}$ | $\displaystyle \frac{dI}{dt} \cdot \Delta t$
------------- | ------------- | ------------- | -------------
0 | 10 | 3 | 1.5
0.5 | = 10 + 1.5 = 11.5 | 2.96 | 1.48
1 | = 11.5 + 1.48 = 12.98 | 2.92 | 1.46
1.5 | = 12.92 + 1.46 = 14.38 | 2.88 | 1.44
2 | = 14.38 + 1.44 = 15.82 |  | 


Notice how we have started to build up a way to organize how to compute the solution.  Each column is a ``step'' of the method, computing the solution at a new timestep based on our step size $\Delta t$.  The third column just computes the value of the derivative for a particular time and $I$, and then the fourth column is the *increment* size, or the amount we are forecasting the solution will grow by to the next timestep.  (There are other ways to think about this, but if you have a *rate of change * multiplied by a time increment this will give you an approximation to the net change in a function.)


This idea of *approximate, forecast, repeat* is the heart of many [numerical methods](https://en.wikipedia.org/wiki/Numerical_methods_for_ordinary_differential_equations) that approximate solutions to differential equations. We display the results from additional steps in Figure \@ref(fig:eulers_ver2).

```{r eulers_ver2,warning=FALSE,message=FALSE,echo=FALSE,fig.cap='Approximation of a solution using local linearity', fig.width=4,fig.height=3}

# Let's clean this up somewhat
t_end <- 200
# Solution
t_soln <- seq(0,t_end,by=0.05)
i_soln <- 130-120*exp(-.025*t_soln)

# First timesteps
dt <- 1
t_1 <- seq(0,t_end,by=dt)

didt = 3*exp(-0.025*t_1)  # We are multiplying by dt
i_approx <- array(0,dim=length(t_1))
i_approx[1] <- 10
for(i in 2:length(t_1)) {
  i_approx[i] <- i_approx[i-1] + didt[i-1]*dt
}

# Smaller timestep
dt <- 0.5
t_2 <- seq(0,t_end,by=dt)
didt = 3*exp(-0.025*t_2)  # We are multiplying by dt
i_approx_2 <- array(0,dim=length(t_2))
i_approx_2[1] <- 10
for(i in 2:length(t_2)) {
  i_approx_2[i] <- i_approx_2[i-1] + didt[i-1]*dt
}


solution <- tibble(t_soln,i_soln)
euler_1 <- tibble(t_1, i_approx)
euler_2 <- tibble(t_2, i_approx_2)

ggplot() +
  geom_line(data=solution, aes(x=t_soln,y=i_soln)) +
  geom_point(data=euler_1, aes(x=t_1,y=i_approx),color='blue',size=1.5) +
  geom_line(data=euler_1, aes(x=t_1,y=i_approx),color='blue') +
  geom_point(data=euler_2, aes(x=t_2,y=i_approx_2),color='red',size=1.5) +
  geom_line(data=euler_2, aes(x=t_2,y=i_approx_2),color='red') +
  labs(x="Time (days)",y="Infected people")
  

```

You may notice that the approximation in Figure \@ref(fig:eulers_ver2) compares very favorably to the actual solution function.  At the end, we have the following comparisons:


$t$ | Euler's Method ($\Delta t = 1$) | Euler's Method ($\Delta t = 1$) |Actual Solution
------------- | ------------- | ------------- | -------------
190 | `r round(i_approx[t_1==190],1)` | `r round(i_approx_2[t_2==190],1)` | `r round(i_soln[t_soln==190],1)`
195 |  `r round(i_approx[t_1==195],1)` | `r round(i_approx_2[t_2==195],1)` | `r round(i_soln[t_soln==195],1)`
200 |  `r round(i_approx[t_1==200],1)` | `r round(i_approx_2[t_2==200],1)` | `r round(i_soln[t_soln==200],1)`


There is a tradeoff here - the smaller stepsizes you have the more work it will take to compute you solution.  You may have seen a similar tradeoff in Calculus when you explored numerical integration and Riemann sums.

## Building an iterative method
Let's more carefully formulate Euler's method.  Consider the following differential equation:

Say for example we have the following equation that describes the rate of change of the spread of a disease (such as Ebola, as we covered in the first section):

$$ \frac{dI}{dt} = 0.003 I \cdot (4000-I)  $$


Let's call the function $f(I) = 0.03 I\cdot (4000-I)$. In order to numerically approximate the solution, we will need to recall some concepts from calculus.   One way that we can approximate the derivative is through a difference function:

$$ \frac{dI}{dt} = \lim_{\Delta t \rightarrow 0} \frac{I(t+\Delta t) - I(t)}{\Delta t} $$

 As long as we consider the quantity $\Delta t$ to be small (say for this problem 0.1 days if you would like to have units attached to this), we can approximate the derivative with difference function on the right hand side.  With this information, we have a reasonable way to organize the problem:

\begin{align*}
\frac{I(t+\Delta t) - I(t)}{\Delta t}  &= 0.003 I \cdot (4000-I) \\
I(t+\Delta t) - I(t)  &= 0.003 I \cdot (4000-I)  \cdot \Delta t \\
I(t+\Delta t) &= I(t) + 0.003 I \cdot (4000-I)  \cdot \Delta t
\end{align*} 


The last equation $I(t+\Delta t) = I(t) + 0.03 I \cdot (4000-I)  \cdot \Delta t  = f(I) \cdot \Delta t$ is a reasonable way to define an iterative system, especially if we have a spreadsheet program.  Here is some code in `R` that can define a for loop to do this iteratively and then plot the solution with `plotData`:

```{r,warning=FALSE,fig.cap='An iterative method', fig.width=4,fig.height=3}

# Define your timestep and time vector
dt <- 0.1
t <- seq(0,2,by=dt)

# Define the number of steps we take.  This is equal to 10 / dt (why?)
N <- length(t)
# Define the initial condition
i_approx <- 10

# Define a vector for your solution:the derivative equation
for(i in 2:N) {  # We start this at 2 because 
  didt <- .003 * i_approx[i-1] * (4000-i_approx[i-1])
  i_approx[i] <- i_approx[i-1] + didt*dt
}


# Define your data for the solution:
solution_data <- tibble(time =t,
                        infected = i_approx)

# Now plot your solution:
ggplot() +
  geom_line(data = solution_data, aes(x=time,y=infected)) 
  labs(x='Time',
         y='Infected')

```

## An iterative method for systems

In the last example we created the solution directly in R.  Here we are going to show how we can utilize Euler's method to easily solve a system of differential equations.  To recap, the most general form of a differential equation is: 

$$ \displaystyle \frac{d\vec{y}}{dt} = f(\vec{y},\vec{\alpha},t), $$
where $\vec{y}$ is the vector of state variables you want to solve for, and $\vec{\alpha}$ is your vector of parameters.

At a given initial condition, Euler's method applies locally linear approximations to forecast the solution forward $\Delta t$ time units:

$$ \vec{y}_{n+1} = y_{n} + f(\vec{y}_{n},\vec{\alpha},t_{n}) \cdot \Delta t $$



In order to make Euler's method work, you will need four things:

- The size ($\Delta t$) of your timesteps.
- The number of timesteps you wish to run Euler's method.  More timesteps means more computational time.  If $N$ is the number of timesteps, $\Delta t \cdot N$ is the total time.
- A function that we have for our dynamics (this is $\displaystyle \frac{d\vec{y}}{dt} = f(\vec{y},\vec{\alpha},t)$),
- The values of the vector of parameters $\vec{\alpha}$.

We will work you through this step by step, with example code.

## Euler's method applied to systems
Here is a sample code that shows the dynamics for the Lotka-Volterra equations, as studied in Chapter 3:

\begin{align*} 
\frac{dH}{dt} &= r H - bHL \\ 
\frac{dP}{dt} &= e  b  H L - dL
\end{align*}

We are going to use Euler's method to solve this differential equation.  The `MAT369Code` has a function that implements Euler's method, however we will need to define some things first.

### Set up your time values
First you need to set up the length of each timestep ($\delta T$) and the number of timesteps we compute the solution.  You do this by specifiying the variables `deltaT` and `timeSteps`:
```{r, fig.show='hold'}
library(MAT369Code)
deltaT <- 0.05    # timestep length
timeSteps <- 200   # must be a number greater than 1
```


### Set up your parameters
Next you should specify the values of your parameters $\vec{alpha}$.  The good news is that you can name the parameters and their values, and just refer to the shorthand name (not the numerical value) in subsequent code.

```{r, fig.show='hold'}
parameters <- c(r = 2, b = 0.5, e = 0.1, d = 1)   # parameters: a named vector
```

### Set up your initial conditions.
Now we need to specify the starting values (initial conditions) for our differential equation.  The code can easily accommodate multiple initial conditions (so you can test different initial conditions - I call this a "run").

```{r, fig.show='hold'}

initialCondition <- c(H=1, L=3)  # Be sure you have enough conditions as you do variables.

# To do multiple initial conditions, here is some sample code:
# initialCondition <- rbind(c(H=1, L=3),  # Separate out initial conditions with a separate row.
#                        c(H=5,L=6),
#                        c(H=10,L=9) )
```



### Set up your dynamics
You need to have a function for the dynamics of the differential equation.  The function `euler` looks for a function called `dynamics`.

The format of this function is structured, but easily modifiable. For simplicity, I refer to the left hand side of the dynamics as `dVariable` so for example `dH` signifies $\displaystyle \frac{dH}{dt}$.

```{r, fig.show='hold'}
# R function to calculate the value of the derivatives at each time value
# Use the names of the variables as defined in the vectors above

dynamics <- function(t, state, parameters){
  with(as.list(c(state, parameters)), {  # Do not edit this line
    dH = r*H - b*H*L  # <-- You may edit this line
    dL = e*b*H*L - d*L # <-- You may edit this line.
    return(list(c(dH, dL)))  # <-- If you have more equations you will need to list the dVariables
  })
}


```



### Putting it all together

Now we have everything we need!  Basically the next step is to run the command `euler`, which will print to the screen a plot of your equation. If you have entered everything above correctly, you may copy the following code to your working file.

If you have multiple initial conditions, you will get a stacked plot for each run.

```{r, fig.show='hold',fig.width=4,fig.height=3}
euler(deltaT,timeSteps,initialCondition,FUN=dynamics,parameters=parameters)
```




\newpage


## Exercises

```{exercise}
Verify that $I(t) = 130-120e^{-0.25t}$ is a solution to the differential equation $$\displaystyle \frac{dI}{dt} = 130-0.025I $$ with $I(0)=10$.
```

&nbsp;

```{exercise}
The following exercise will help you explore the relationships between stepsize, ending points, and number of steps needed.
\begin{enumerate}
\item If we wish to do an Euler's method solution with step size 1 second and ending at $T=5$ seconds, how many steps will we take?
\item If we wish to do an Euler's method solution with step size 0.5 seconds and ending at $T=5$ seconds, how many steps will we take?
\item If we wish to do an Euler's method solution with step size 0.1 seconds and ending at $T=5$ seconds, how many steps will we take?
\item If we wish to do an Euler's method solution with step size $\Delta t$ and go to ending value of $T$, what is an expression that relates the number steps $N$ as a function of $\Delta t$ and $T$?
\end{enumerate}
```


&nbsp;

```{exercise}
For each of the following differential equations, apply Euler's method to generate a numerical solution to the differential equation and plot your solution.  The stepsize ($\Delta t$) and number of iterations ($N$) are listed.

\begin{enumerate}
\item Differential equation: $\displaystyle \frac{dS}{dt} =3-S$. Set $\Delta t = 0.1$, $N = 50$. Initial conditions:  $S(0) = 0.5$, $S(0) = 5$.
\item Differential Equation: $\displaystyle \frac{dS}{dt} =\frac{1}{1-S}$. Set $\Delta t = 0.01$, $N = 30$. Initial conditions:  $S(0) = 0.5$, $S(0) = 2$.
\item Differential equation: $\displaystyle \frac{dS}{dt} = 0.8 \cdot S \cdot (10-S)$. Set $\Delta t = 0.1$, $N = 50$. Initial conditions:  $S(0) = 3$, $S(0) = 10$.
\end{enumerate}
```

&nbsp;

```{exercise}
Evaluate  $\displaystyle \lim_{S \rightarrow 1} \frac{1}{1-S}$ (this means evaluate the limit as $S \rightarrow 1^{-}$ and $S \rightarrow 1^{+}$).  What can you say about the rate of change of $S$ near $S \approx 1$ for the differential equation $\displaystyle \frac{dS}{dt} =\frac{1}{1-S}$?
```

&nbsp;

```{exercise}
Let's do some more work with Euler's method for $\displaystyle \frac{dS}{dt} = 0.8 \cdot S \cdot (10-S)$.  This time set $S(0)=15$, $\Delta t = 0.1$, $N = 10$. 
When you examine your solution, what is incorrect about the Euler's method solution based on your qualitative knowledge of the underlying dynamics? Now calculate Euler's method for the same differential equation for the following conditions: $S(0)=15$, $\Delta t = 0.01$, $N = 100$.  What has changed in your solution? 
```

&nbsp;

```{exercise}
Let's do some more work with Euler's method for $\displaystyle \frac{dS}{dt} =\frac{1}{1-S}$.  This time set $S(0)=1.5$, $\Delta t = 0.1$, $N = 10$ and also $S(0)=1.5$, $\Delta t = 0.01$, $N = 100$.  Between these two solutions, what has changed? Do you think it is numerically possible to calculate a reasonable solution for Euler's method near $S=1$?  (*note: this differential equation is an example of finite time blow up*)
```

&nbsp;

```{exercise}
In the lynx hare example we can also consider an alternative system where the growth of the hare is not exponential:
  
  \begin{align*} 
\frac{dH}{dt} &= \frac{r H}{1+kH} - bHL \\ 
\frac{dP}{dt} &= e  b  H L - dL
\end{align*}

\begin{enumerate}
\item Evaluate $\displaystyle \lim_{t \rightarrow \infty} \frac{r H}{1+kH}$.  Explain how this growth rate compares to $rH$.
\item Apply Euler's method to numerically solve this system of equations when $k=0.1$ and $k=1$.  How does this compare to the original system studied in this section?  In the second case set the number of timesteps to be 2000.  Comment on this alternative growth rate - does it make biological sense?

\end{enumerate}
```
&nbsp;

For the following differential equations use the code `euler` to generate a numerical solution to the differential equation and plot your solution.  The stepsize ($\Delta t$) and number of iterations ($N$) are listed.  After your Euler's method solution is determined, explain why you would expect the behavior in the solution that you see. 

```{exercise}
Consider the following:
\begin{itemize}
\item Differential equation: $\displaystyle \frac{dS}{dt} = r \cdot S \cdot (K-S) $.
\item Set $r=1.2$ and $K=3$.
\item Set $\Delta t = 0.1$, $N = 50$.
\end{itemize}

Plot a solution with the initial conditions:  $S(0) = 1$, $S(0) = 3$, $S(0) = 5$, all on the same graph.  What do you notice?  Make a hypothesis regarding the long term behavior of this system. Then plot a few more solution curves to verify your guess.  Be sure to include your additional solution curves in your write up.
```

&nbsp;

```{exercise}
Consider the following:
\begin{itemize}
\item Differential equation: $\displaystyle \frac{dS}{dt} =K-S $.
\item Set $K=2$.
\item Set $\Delta t = 0.1$, $N = 50$.
\end{itemize}

Plot a solution with the initial conditions:  $S(0) = 0$, $S(0) = 2$, $S(0) = 5$, all on the same graph.  What do you notice?  Make a hypothesis regarding the long term behavior of this system. Plot a few more solution curves to verify your guess.  Be sure to include your additional solution curves in your write up.
```


