# Probability and Likelihood Functions {#likelihood-09}

The problem we examined in the last chapter was the following:

> Determine the set of parameters $\vec{\alpha}$ that minimize the difference between data $\vec{y}$ and the output of the function $f(\vec{x}, \vec{\alpha})$ and measured error $\vec{\sigma}$.


We are going to examine the linear regression problem again using a smaller dataset. Ohe to the linear regression problem is through _likelihood functions_, which is a topic from probability and statistics.

This section will introduce likelihood functions but also discuss some interesting visualization techniques of multivariable functions and contour plots.  We will also see a technique to evaluate a continuous function.  We are starting to build out some R skills and techniques that you can apply in other context.  Let's get started!


## Linear regression, part 2
Assume we have the following (limited) number of points where we wish to fit a function of the form $y=bx$.   

| *x* | *y*  | 
|:------:|:-----:|
| 1 | 3 |
| 2 | 5 |
| 4 | 4 |
| 4 | 10 |


For this example we are forcing the intercept term $a$ to equal zero - for most cases you will just fit the linear equation. See Exercise \@ref(exr:full-linear). Figure \@ref(fig:quick-scatter-07) displays a quick scatterplot of these data:

```{r quick-scatter-07,warning=FALSE,message=FALSE,echo=FALSE, fig.width=3,fig.height=2,fig.cap="A scatterplot of a small dataset."}
data.frame(x =c(1,2,4,6),y =c(3,5,4,10) ) %>%
ggplot(aes(x=x, y=y)) +
 geom_point(size=2,color='red') +
  xlim(c(0,10)) +
  ylim(c(0,15))

```

The goal here is to work to determine the value of $b$ that is most *likely* (in other words, consistent) with the data. In order to do this, we need to take a quick excursion into probability distributions. Let's go!

## Probability
In order to understand likelihood functions, first I am going to review very essential information about probability and probability distributions. Probability is the association of a set of observable events to a quantitative scale between 0 to 1.  (Zero means that event is not possible, 1 means that it definitely can happen).  This definition could be refined somewhat [@devore_modern_2021].  You can consider discrete events (think counting or combinatorial problems) or continuous events.  Here we are only going to consider continuous events, specifically in this case the probability of a parameter obtaining a particular value.

Consider this graphic, which may be familiar to you as the normal distribution or the bell curve:

```{r,warning=FALSE,message=FALSE,echo=FALSE,fig.cap='The normal distribution', fig.width=4,fig.height=3}
ggplot() +
geom_area(data=data.frame(x_val=seq(-5,5,length=200),y_val=dnorm(seq(-5,5,length=200))),
          aes(x=x_val,y=y_val),alpha=.6, fill="#FF6666") +
  xlab("x") +
  ylab("f(x)")

```

We tend to think of the plot and the associated function $f(x)$ as something with input and output (such as $f(0)=$ `r round(dnorm(0),digits=4)`).  However because it is a probability density function, the *area* between two points gives yields the probability of an event to fall within two values:

```{r,warning=FALSE,message=FALSE,echo=FALSE,fig.cap='The area between two values, normally distributed', fig.width=4,fig.height=3}
ggplot() +
geom_area(data=data.frame(x_val=seq(-5,5,length=200),y_val=dnorm(seq(-5,5,length=200))),
          aes(x=x_val,y=y_val),alpha=.6, fill="#FF6666") +
geom_area(data=data.frame(x_val=seq(-0.1,0.1,length=20),y_val=dnorm(seq(-0.1,0.1,length=20))),
          aes(x=x_val,y=y_val),alpha=.8, fill="#000CCC") +
  xlab("x") +
  ylab("f(x)")
```

In this case, the shaded area tells us the probability that our measurement is between $x=-0.1$ and $x=0.1$.  The value of the area, or the probability is `r round(pnorm(0.1)-pnorm(-0.1),digits=5)`.  When you took calculus the area was expressed as a definite integral:  $\displaystyle \int_{-0.1}^{0.1} f(x) \; dx=$ `r round(pnorm(0.1)-pnorm(-0.1),digits=5)`, where $f(x)$ is the formula for the probability density function for the normal distribution.



The basic idea is that we can assign values to an outcomes as a way of displaying our belief (confidence) in the result. With this intuition we can summarize key facts about probability density functions:

- $f(x) \geq 0$ (this means that probability density functions are positive values)
- Area integrates to one (in probability, this means we have accounted for all of our outcomes)

The formula for the normal distribution is 
\begin{equation}
f(x)=\frac{1}{\sqrt{2 \pi} \sigma } e^{-(x-\mu)^{2}/(2 \sigma^{2})}
\end{equation}

Where $\mu$ is the mean and $\sigma$ is the standard deviation

### Other probability distributions
Beyond the normal distribution some of the more common ones we utilize in the parameter estimation are the following:

- **Uniform:**  For this distribution we must specify between a minimum value $a$ and maximum value $b$.

```{r,warning=FALSE,message=FALSE,echo=FALSE,fig.cap='The uniform distribution', fig.width=4,fig.height=3}
ggplot() +
geom_area(data=data.frame(x_val=seq(-5,5,length=200),y_val=dunif(seq(-5,5,length=200),min = -5,max=5)),
          aes(x=x_val,y=y_val),alpha=.6, fill="#FF6666") +
  xlab("x") +
  ylab("f(x)") +
  ylim(c(0,0.5)) +
  ggtitle('Uniform Probability Distribution')

```

The formula for the uniform distribution is 
\begin{equation}
f(x)=\frac{1}{b-a} \mbox{ for } a \leq x \leq b
\end{equation}


- **Exponential:** For this distribution we must specify between a rate parameter $\lambda$.


```{r,warning=FALSE,message=FALSE,echo=FALSE,fig.cap='The exponential distribution', fig.width=4,fig.height=3}
ggplot() +
geom_area(data=data.frame(x_val=seq(0,5,length=200),y_val=dexp(seq(0,5,length=200))),
          aes(x=x_val,y=y_val),alpha=.6, fill="#FF6666") +
  xlab("x") +
  ylab("f(x)") +
  ylim(c(0,0.5)) +
  ggtitle('Exponential Probability Distribution')

```

The formula for the exponential distribution is 
\begin{equation}
f(x)=\lambda e^{-\lambda x} \mbox{ for } x \geq 0 
\end{equation}

where $\lambda$ is the rate parameter

- **Lognormal:** This distirbution is for positive values, with mean $\mu$ and standard deviation $\sigma$.

```{r,warning=FALSE,message=FALSE,echo=FALSE,fig.cap='The lognormal distribution', fig.width=4,fig.height=3}
ggplot() +
geom_area(data=data.frame(x_val=seq(0,5,length=200),y_val=dlnorm(seq(0,5,length=200))),
          aes(x=x_val,y=y_val),alpha=.6, fill="#FF6666") +
  xlab("x") +
  ylab("f(x)") +
  ggtitle('Lognormal Probability Distribution')

```



The formula for the lognormal distribution is 
\begin{equation}
f(x)=\frac{1}{\sqrt{2 \pi} \sigma x } e^{-(\ln(x)-\mu)^{2}/(2 \sigma^{2})} \mbox{ for } x \geq 0 
\end{equation}


### Computing probabilities in R
Here is the good news with `R`: the commands to generate densities and cumulative distributions are already included!  There are a variety of implementations: both for the density, cumulative distribution, random number generation, and lognormal distributions from these.  For the moment, Table \ref{table:prob-dist} summarizes some common probability distributions in `R`.

\begin{table}
\begin{tabular}{c|c|c|c}
Distribution & Key Parameters & R command & Density Example \\ \hline
Normal & $\mu \rightarrow$ \texttt{mean}, $\sigma \rightarrow$ \texttt{sd} & \texttt{norm} 
& \texttt{dnorm(mu=0,sd=1,seq(-5,5,length=200))} \\
Uniform & $a \rightarrow$ \texttt{min}, $b \rightarrow$ \texttt{max} & \texttt{unif} & \texttt{dunif(seq(-5,5,length=200),min = -5,max=5)} \\
Exponential & $\lambda \rightarrow$ \texttt{rate} & \texttt{exp} & \texttt{dexp(seq(0,5,length=200))} \\
Normal & $\mu \rightarrow$ \texttt{meanlog}, $\sigma \rightarrow$ \texttt{sdlog} & \texttt{lnorm} 
& \texttt{dlnorm(seq(0,5,length=200))}
\end{tabular}
\caption{Summary of common probability distributions in \texttt{R}}\label{table:prob-dist}
\end{table}

To make the graphs of these density functions in `R` we use the prefix `d + ` the name (`norm`, `exp`) etc of the distribution we wish to specify, including any of the key parameters.  If we don't include any of the parameters then it will just use the defaults (which you can see by typing `?NAME` where `NAME` is the name of the command (i.e. `?dnorm`).


Here is some sample code to plot the lognormal distribution:

```{example}
Make a graph of the lognormal density function with $\mu=0$ and $\sigma=1$ from $0 \leq x \leq 5$.
```

```{remark}
For this case we are using the defaults of the lognormal distribution.  The code is the following, and plotted in Figure \@ref(fig:log-norm-plot).
```

```{r log-norm-plot,fig.cap='Code to plot the lognormal distribution', fig.width=4,fig.height=3}

x <- seq(0,5,length=200)
y <- dlnorm(x)  # Just use the mean defaults 

lognormal_data <- tibble(x,y)

ggplot() +
  geom_line(data = lognormal_data,
             aes(x=x,y=y)) +
  labs(x = 'x', y= 'Lognormal density')


```


To find the area between two values in a density function we use the prefix `p`.


```{example exp-area}
Use `R` to evaluate $\displaystyle \int_{1}^{2} e^{-x} \; dx$.
```

```{remark}
The function $e^{-x}$ is the exponential probability distribution with $\lambda=1$. For this example if we wanted to find the area between two values in the exponential density in the shaded graph we would type `pexp(2)-pexp(1)` at the `R` console, which would give the value of `r round(pexp(2)-pexp(1),digits=3)`.  A visual representation of this area is shown in Figure \@ref(fig:exp-area-plot).
```


```{r exp-area-plot,warning=FALSE,message=FALSE,echo=FALSE,fig.cap='The area for the exponential distribution', fig.width=4,fig.height=3}
ggplot() +
geom_area(data=data.frame(x_val=seq(0,5,length=200),y_val=dexp(seq(0,5,length=200))),
          aes(x=x_val,y=y_val),alpha=.6, fill="#FF6666") +
geom_area(data=data.frame(x_val=seq(1,2,length=20),y_val=dexp(seq(1,2,length=20))),
          aes(x=x_val,y=y_val),alpha=.8, fill="#000CCC") +
  xlab("x") +
  ylab("f(x)") +
  annotate("text", x = 1.5, y = 0.5, label = "Area of region: 0.233")


```




## Connecting to linear regression
Now that we have made that small excursion into probablity, let's start to return back to the linear regression problem.  Another way to phrase this the linear regression problem studied in the last chapter is to examine the probability distribution of the model-data residual $\epsilon$:

\begin{equation}
\epsilon_{i} = y_{i} - f(x_{i},\vec{\alpha} ).
\end{equation}

The approach with likelihood functions assumes a particular probability distribution on each residual.  One common assumption is that the residual *distribution* is normal with mean $\mu=0$ and standard deviation $\sigma$ (which could be specified as measurement error, etc).

\begin{equation}
L(\epsilon_{i}) = \frac{1}{\sqrt{2 \pi} \sigma} e^{-\epsilon_{i}^{2} / 2 \sigma^{2} },
\end{equation}

To extend this further across all measurement, we use the idea of *independent, identically distributed* measurements so the joint likelihood of **all** the residuals is the product of the likelihoods

\begin{equation}
L(\vec{\epsilon}) = \prod_{i=1}^{N} \frac{1}{\sqrt{2 \pi} \sigma} e^{-\epsilon_{i}^{2} / 2 \sigma^{2} },
\end{equation}



We are making progress here, but in order to fully characterize the solution we need to specify the parameters $\vec{\alpha}$.  A simple redefining of the likelihood function where we specify the measurements ($x$ and $y$) and parameters ($\vec{\alpha}$) is all we need:

\begin{equation}
L(\vec{\alpha} | \vec{x},\vec{y} )= \prod_{i=1}^{N}  \frac{1}{\sqrt{2 \pi} \sigma} \exp(-(y_{i} - f(x_{i},\vec{\alpha} ))^{2} / 2 \sigma^{2} )  
\end{equation}

Now we have a function where the best parameter estimate is the one that optimizes the likelihood.


To get back to our original linear regression problem.  As a reminder we wanted to fit the function $y=bx$ to the following set of points:

| *x* | *y*  | 
|:------:|:-----:|
| 1 | 3 |
| 2 | 5 |
| 4 | 4 |
| 4 | 10 |

The likelihood $L(\epsilon_{i}) ~ N(0,\sigma)$ characterizing these data are the following:

\begin{equation}
L(b) = \left( \frac{1}{\sqrt{2 \pi} \sigma}\right)^{4} e^{-\frac{(3-b)^{2}}{2\sigma}} \cdot e^{-\frac{(5-2b)^{2}}{2\sigma}}  \cdot e^{-\frac{(4-4b)^{2}}{2\sigma}}  \cdot e^{-\frac{(10-4b)^{2}}{2\sigma}} (\#eq:small-data-likely)
\end{equation}

For the purposes of our argument here, we will assume $\sigma=1$. Let's also make a plot of this likelihood function:

```{r small-likelihood-plot,warning=FALSE,message=FALSE,echo=FALSE,fig.cap='The likelihood function for the small dataset', fig.width=4,fig.height=3}
b=seq(0,5,0.01)
new_data<-data.frame(x =c(1,2,4,4),y =c(3,5,4,10) )
lb = map(.x=b,.f=~(1/(2*pi)^length(new_data$y))*exp(-sum((new_data$y-.x*new_data$x)^2)/4)) %>% as.numeric()

data.frame(b,lb) %>%
  ggplot(aes(x=b,y=lb)) +
  geom_line() +
  geom_vline(xintercept = b[which.max(lb)],color='red') +
  ylab('L(b)')

```


Note that although the $y$ values of the likelihood function are really small, the likelihood function is maximized at $b=1.86$.


Notice how in the plot of our likelihood function we had really small values of $L(b)$.  It is also common to use the log likelihood (Equation \@ref(eq:loglikely):

\begin{equation}
\begin{split}
\ln(L(\vec{\alpha} | \vec{x},\vec{y} )) &=  N \ln \left( \frac{1}{\sqrt{2 \pi} \sigma} \right) - \sum_{i=1}^{N} \frac{ (y_{i} - f(x_{i},\vec{\alpha} )^{2}}{ 2 \sigma^{2}} \\
 & = - \frac{N}{2} \ln (2) - \frac{N}{2} \ln(\pi) - N \ln( \sigma) - \sum_{i=1}^{N} \frac{ (y_{i} - f(x_{i},\vec{\alpha} )^{2}}{ 2 \sigma^{2}}
\end{split} (\#eq:loglikely)
\end{equation}

In the homework you will be working on how to transform the likelihood function $L(b)$ to the log-likelihood $\ln(L(b))$.

For our purposes here we are going to assume independent, identical, normally distributed errors with a mean 0 for the likelihood functions.  This is a commonly used assumption for this approach, but you should always think about the errors in more advanced applications.


## Plotting likelihood surfaces
Ok, we are going to examine a second example from @gause_experimental_1932 which modeled the growing of yeast in solution.  This classic paper examines the biological principal of *competitive exclusion*, how one species can out compete another one for resources. For our purposes here we are going to examine a model for one species growing without competition.

First let's make a quick plot of the function:
```{r, fig.show='hold', fig.width=4,fig.height=3}
### Make a quick ggplot of the data

ggplot() +
  geom_point(data = yeast,
             aes(x=time,y=volume),
             color='red',
             size=2) +
  labs(x = 'Time', y= 'Volume')

```

We are going to assume the population of yeast (represented with the measurement of volume) over time changes according to the equation:

\begin{equation}
\frac{dy}{dt} = -by \frac{(K-y)}{K},
\end{equation}

where $y$ is the population of the yeast and $b$ represents the growth rate and $K$ is the carrying capacity of the population. It can be shown that the solution to this differential equation is $\displaystyle y = \frac{K}{1+e^{a-bt}}$, where the additional parameter $a$ can be found through application of the initial condition $y_{0}$.  In @gause_experimental_1932 the value of $a$ was determined by solving the initial value problem $y(0)=0.45$. In Exercise \@ref(exr:solve-gause) you will show that $\displaystyle a = \ln \left( \frac{K}{0.45} - 1 \right)$.


Here we are going to explore the likelihood function to try to determine the best set of values for the two parameters $K$ and $b$ using the function `compute_likelihood`.

Inputs to the `compute_likelihood` function are the following:

- A function $y=f(x,\vec{\alpha})$
- A dataset $(x,y)$
- Ranges of your parameters $\vec{\alpha}$. 

The `compute_likelihood` function also has an optional input that allows you to specify if you want to compute the likelihood or the log likelihood.





We are going to plot the likelihood equation surface for the `yeast` dataset, contained in the data frame `yeast`.  Recall the function that we fit to determine is $\displaystyle y = \frac{K}{1+e^{a-bt}}$.


Next we will define the equation used to compare our model in the likelihood.  As with the functions `euler` or `systems` we need to define this function:
```{r, fig.show='hold'}
library(MAT369Code)

# Gause model equation
gause_model <- volume ~ k/(1+exp(log(k/0.45-1)-b*time))


# Identify the ranges of the parameters that we wish to investigate
kParam <- seq(5,20,length.out=100)
bParam <- seq(0,1,length.out=100)


# Allow for all the possible combinations of parameters
gause_parameters <- expand.grid(k=kParam,b=bParam)
 

gause_likelihood <- compute_likelihood(model = gause_model,
                                        data = yeast,
                                        parameters = gause_parameters,
                                        logLikely=FALSE) 

```

Ok, let's break this code down step by step:

- The line `gause_model <- volume ~ k/(1+exp(log(k/0.45-1)-b*time))` identifies the formula that relates the variables `time` to `volume`.
- We define the ranges (minimum and maximum values) for our parameters by defining a sequence.  Because we want to look at *all possible combinations* of these parameters we use the command `expand.grid`.
- The input `logLikely` to `compute_likelihood` reports back log likelihood values.

Some care is needed in defining the number of points that we want to evaluate - we will have $100^{2}$ different combinations of $k$ and $b$, which do take time to evaluate.


The output to `compute_likelihood` is a list - this is a flexible data structure.  You can think of this as a collection of items.  In this case, what gets returned are two data frames: `likelihood`, which is a data frame of likelihood values for each of the parameters and `opt_value`, which reports back the values of the parameters that optimize the likelihood function.  Note that the optimum value is *an approximation*, as it is just the optimum from the input values.  Let's take a look at the reported optimum values, which we can do with the syntax `LIST_NAME$VARIABLE_NAME`, where the dollar sign ($) helps identify which variable from the list you are investigating.

```{r, fig.show='hold'}

gause_likelihood$opt_value

```

It is also important to visualize this likelihood function.  For this dataset we have the two parameters $k$ and $b$, so the likelihood function will be a likelihood surface.  The code to generate the plot looks a little different:

```{r gause-likely-plot,fig.cap="Likelihood surface and contour lines for the Gause dataset."}

# Define the likelihood values
my_likelihood <- gause_likelihood$likelihood

# Make a contour plot
ggplot(data = my_likelihood) +
  geom_tile(aes(x = k,y = b,fill = l_hood)) + 
  stat_contour(aes(x = k, y = b, z = l_hood))

```

Similar to before, let's take this step by step:

- The command `my_likelihood` just puts the likelihood values in a data frame.
- The `ggplot` command is similar as before.
- We use `geom_tile` to visualize the likelihood surface.  There are three required inputs from the `my_likelihood` data frame: the `x` and `y` axis data values and the `fill` value, which represents the height of the likelihood function.
- The command  `stat_contour` draws the contour lines, or places where the likelihood function is the same.  Notice how we used `z = l_hood` rather than `fill` here.

I chose some broad parameter ranges at first, so let's make a likelihood plot, exploring parameters closer to the last optimum value:

```{r revised-gause-likelihood, fig.show='hold',fig.cap="Revised likelihood surface. The computed location of the optimum value is shown as a red point."}

# Gause model equation
gause_model <- volume ~ k/(1+exp(log(k/0.45-1)-b*time))


# Identify the (new) ranges of the parameters that we wish to investigate
kParam <- seq(11,14,length.out=100)
bParam <- seq(0.1,0.3,length.out=100)


# Allow for all the possible combinations of parameters
gause_parameters_rev <- expand.grid(k=kParam,b=bParam)
 

gause_likelihood_rev <- compute_likelihood(model = gause_model,
                                        data = yeast,
                                        parameters = gause_parameters_rev,
                                        logLikely=FALSE) 

# Report out the optimum values
opt_value_rev <- gause_likelihood_rev$opt_value

opt_value_rev


# Define the likelihood values
my_likelihood_rev <- gause_likelihood_rev$likelihood

# Make a contour plot
ggplot(data = my_likelihood_rev) +
  geom_tile(aes(x = k,y = b,fill = l_hood)) + 
  stat_contour(aes(x = k, y = b, z = l_hood)) +
  geom_point(data=opt_value_rev,aes(x=k,y=b),color='red')


```

The reported values for $k$ and $b$ may be close to what was reported from the first time.  Notice that I also added in the location of the optimum point with the code `geom_point(data=opt_value_rev,aes(x=k,y=b),color='red')`.


Finally we can use the optimized parameters to compare the function against the data:


```{r, fig.show='hold', fig.width=4,fig.height=3,fig.cap="Model and data comparison from maximum likelihood estimation."}

# Define the parameters and the times that you will evaluate the equation
my_params <- gause_likelihood_rev$opt_value
time=seq(0,60,length.out=100)

# Get the right hand side of your equations
new_eq <- gause_model %>%
  formula.tools::rhs()

# This collects the parameters and data into a list
in_list <- c(my_params,time) %>% as.list()

# The eval command evaluates your model
out_model <- eval(new_eq,envir=in_list)


# Now collect everything into a dataframe:
my_prediction <- tibble(time = time, volume = out_model)


ggplot() +
 geom_point(data = yeast,
            aes(x=time,y=volume),
            color='red',
            size=2) +
 geom_line(data = my_prediction,
           aes(x=time,y=volume)) +
 labs(x = 'Time', y= 'Volume')



```

All right, this code block has some new commands and techniques that need explaining.  Once we have the parameter estimates we need to compute the modeled values.

- First we define the `params` and the `time` we wish to evaluate with our model.
- We need to evaluate the right hand side of $\displaystyle y = \frac{K}{1+e^{a+bt}}$, so the definition of `new_eq` helps to do that, using the package `formula.tools`.
- The `%>%` is the `tidyverse` [https://r4ds.had.co.nz/pipes.html#pipes](pipe).  This is a very useful command to help make code more readable!
- `in_list <- c(params,my_time) %>% as.list()` collects the parameters and input times in one list to evaluate the model with `out_model <- eval(new_eq,envir=in_list)`
- In order to plot we make a data frame `my_prediction`

And the rest of the plotting commands you should be used to.

### Single / more than two parameter parameter
Are you able use the `compute_likelihood` function if you have one or more than two parameters?  Why yes you can!

For a single parameter likelihood function then you will just use `geom_line` rather than `geom_tile` to plot the surface.



\newpage

## Exercises

```{exercise solve-gause}
Algebraically solve the equation $\displaystyle 0.45 = \frac{K}{1+e^{a}}$ for $K$.
```

&nbsp;

```{exercise}
Evaluate $\displaystyle \int_{0}^{5} 2 e^{-2x} \; dx$ by hand.  Then Use `R` to compute the value of $\int_{0}^{5} 2 e^{-2x} \; dx$.  Does your computed answer match with what you found in `R`?
```

&nbsp;  


```{exercise}
Make a plot of the normal density distribution with $mu=2$ and $\sigma=0.1$ for $0 \leq x \leq 4$.  Then use `R` to compute the following integral: $\displaystyle f(x)=\frac{1}{\sqrt{2 \pi} 0.1} e^{-(x-2)^{2}/(2\cdot 0.1^{2})}$
```

&nbsp;



```{exercise}
Visualize the likelihood function for the `yeast` dataset, but in this case report out and visualize the loglikelihood. (This means that you are setting the option `logLikely = TRUE` in the `compute_likelihood` function.)  Compare the loglikelihood surface to Figure \@ref(fig:revised-gause-likelihood).
```

&nbsp;

```{exercise}
When we generated our plot of the likelihood function in Figure \@ref(fig:small-likelihood-plot) we assumed that $\sigma=1$ in Equation \@ref(eq:small-data-likely).  For this exercise you will explore what happens in Equation \@ref(eq:small-data-likely) as $\sigma$ increases or decreases.

\begin{enumerate}[label=\alph*.]
\item Use desmos to generate a plot of Equation \@ref(eq:small-data-likely), but let $\sigma$ be a slider.  What happens to the shape of the likelihood function as $\sigma$ increases?
\item How does the estimate of $b$ change as $\sigma$ changes?
\item The spread of the distribution (in terms of it being more peaked or less peaked)is a measure of uncertainty of a parameter estimate. How does the resulting parameter uncertainty change as $\sigma$ changes?
\end{enumerate}
```

&nbsp;


```{exercise}
Using Equation \@ref(eq:small-data-likely) with $\sigma = 1$:

  \begin{enumerate}[label=\alph*.]
  \item Apply the natural logarithm to both sides of this expression.
  \item Using properties of logarithms, show that the loglikelihood function $\ln(L(b)) =-2 \ln(2) - 2 \ln (\pi) -(3-b)^{2}-(5-2b)^{2}-(4-4b)^{2}-(10-4b)^{2}$.
\item Make a plot of the log likelihood function (in desmos or R).  Where is this function optimized?  Is it a maximum or a minimum value?
  \item Compare this likelihood estimate for $b$ to what was found in Figure \@ref(fig:small-likelihood-plot). 
  \end{enumerate}
```

&nbsp;


```{exercise full-linear}
Consider the linear model $y=a+bx$ for the following dataset:


| *x* | *y*  | 
|:------:|:-----:|
| 1 | 3 |
| 2 | 5 |
| 4 | 4 |
| 4 | 10 |

\begin{enumerate}[label=\alph*.]
\item With the function \texttt{compute\_likelihood}, generate a contour plot of both the likelihood and log-likelihood functions.
\item Make a scatterplot of these data with the equation $y=a+bx$ with your maximum likelihood parameter estimates.
\item Earlier when we fit $y=bx$ we found $b=1.86$. How does adding $a$ as a model parameter affect your estimate of $b$?
\end{enumerate}
```

&nbsp;



```{exercise}
For the function $\displaystyle P(t)=\frac{K}{1+e^{a+bt}}$, with $P(0)=P_{0}$, determine an expression for the parameter $a$ in terms of $K$, $b$, and $P_{0}$.
```

&nbsp;


```{exercise}
The values of returned by the maximum likelihood estimate were a little different from those reported in @gause_experimental_1932:


**Parameter** | **Maximum Likelihood Estimate**  | **@gause_experimental_1932** 
|:------:|:-----:|:-----:|
$K$ | 12.7 |  13.0 |
$b$ | 0.24242 |  0.21827 |
  
Make of plot of the function $\displaystyle y = \frac{K}{1+e^{a-bt}}$ with $\displaystyle a = \ln \left( \frac{K}{0.45} - 1 \right)$ for both parameter values, along with the `yeast` data.  to generate plots with the `yeast` data with the curves with parameters from both the Maximum Likelihood estimate and from @gause_experimental_1932.  Which approach does a better job representing the data?
```

&nbsp;


```{exercise}
An equation that relates a consumer's nutrient content (denoted as $y$) to the nutrient content of food (denoted as $x$) is given by: $\displaystyle y = c x^{1/\theta}$, where $\theta \geq 1$ and $c$ are both constants.

\begin{enumerate}[label=\alph*.]
\item Use the dataset `phosphrous` make a scatter plot with the variable `algae` on the horizontal axis, `daphnia` on the vertical axis.
\item Generate a contour plot for the likelihood function for these data. You may assume $1 \leq \theta \leq 20$ and $0 \leq c \leq 5$.  What are the values of $\theta$ and $c$ that optimize the likelihood? \emph{Hint:} for the dataset `phosphorous` be sure to use the variables $x=$\texttt{algae} and $y=$\texttt{daphnia}.
\item With your values of $c$ and $\theta$ add the function $W$ to your scatterplot and compare the fitted curve to the data.
\end{enumerate}
```

&nbsp;

```{exercise}
A dog's weight $W$ (pounds) changes over $D$ days according to the following function:
\begin{equation}
W =f(D,p_{1},p_{2})= \frac{p_{1}}{1+e^{2.462-p_{2}D}}
\end{equation}
where we have the parameters $p_{1}$ and $p_{2}$. The dataset `wilson` shows how the weight of a dog (named Wilson changes adapted from  [here](http://bscheng.com/2014/05/07/modeling-logistic-growth-data-in-r/).  

\begin{enumerate}[label=\alph*.]
\item Make a scatterplot with the `wilson` data.  What is the long term weight of the dog? 
\item Generate a contour plot for the likelihood function for these data.  What are the values of $p_{1}$ and $p_{2}$ that optimize the likelihood?  \emph{You may assume that $p_{1}$ and $p_{2}$ are both positive.}
\item With your values of $p_{1}$ and $p_{2}$ add the function $W$ to your scatterplot and compare the fitted curve to the data.
\end{enumerate}
```

&nbsp;

  <!-- See rScripts file for code -->
```{exercise}

Consider the following data which represents the temperature over the course of a day:
\begin{tabular}{lccccccccccccc}
\textbf{Hour} & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 &10 &11 &12  \\ \hline
\textbf{Temperature} ($^{\circ}$F) 
& 54 &53 &55& 54& 58& 58 &61& 63 &67& 66& 67& 69& 68 \\ \hline \hline
\textbf{Hour} &13 &14 &15 &16 &17 &18 &19 &20& 21& 22& 23& 24 & \\
\textbf{Temperature} ($^{\circ}$F) &
 68& 66& 67& 63& 60& 59& 57& 56& 53& 52& 54& 53 & \\ \hline
\end{tabular}

A function that describes these data is $\displaystyle T = A + B \sin \left( \frac{\pi}{12} \cdot H \right) + C \cos \left( \frac{\pi}{12} \cdot H \right)$, where $H$ is the hour and $T$ is the temperature.  Use the function `compute_likelihood` to determine maximum likelihood parameter estimates for $A$, $B$, and $C$.


```
