# (PART) Stability Analysis for Differential Equations {.unnumbered}  


# Systems of Linear Differential Equations {#linearsystems-15}

Here we delve into a deeper understanding of differential equations by examining long term stability of equilibrium solutions. As a first step, Chapter \@ref(linearsystems-15) focuses on *linear* systems of differential equations, such as Equation \@ref(eq:example-ch15):\index{differential equation!linear system}

\begin{equation}
\begin{split} 
\frac{dx}{dt} &= 2x \\ 
\frac{dy}{dt} &= x+y
\end{split} (\#eq:example-ch15)
\end{equation}

Equation \@ref(eq:example-ch15) is a linear system of differential equations \index{differential equation!linear system} because it does contain terms such as $y^{2}$ or $\sin(x)$ on the right hand side of the equation. This chapter focuses on visualizing the phase plane for linear systems and determining the equilibrium solutions. Let's get started!


## Linear systems of differential equations and matrix notation
Another way to express Equation \@ref(eq:example-ch15) is with matrix notation:\index{matrix notation}

\begin{equation}
\begin{split}
\begin{pmatrix} \frac{dx}{dt} \\ \frac{dy}{dt} \end{pmatrix} &=  \begin{pmatrix} 2x \\ x+y \end{pmatrix} \\
&=\begin{pmatrix} 2 & 0 \\ 1 &  1 \end{pmatrix} \begin{pmatrix} x \\ y \end{pmatrix}
\end{split} (\#eq:example-ch15a)
\end{equation}

(Note: we can also use the prime notation ($x'$ or $y'$) to signify  $\displaystyle \frac{dx}{dt}$ or $\displaystyle \frac{dy}{dt}$.) We can also represent Equation \@ref(eq:example-ch15a) in a compact vector notation: $\displaystyle \frac{ d \vec{x} }{dt} = A \vec{x}$, where for this example $\displaystyle \vec{A}=\begin{pmatrix} 2 & 0 \\ 1 &  1 \end{pmatrix}$.

Now let's generalize. A system of linear equations:

\begin{equation}
\begin{split}
\frac{dx}{dt} &= ax+by \\ 
\frac{dy}{dt} &= cx+dy
\end{split} 
\end{equation}

can be expressed in the following way:


\begin{equation}
\begin{pmatrix} \frac{dx}{dt} \\ \frac{dy}{dt} \end{pmatrix} =  \begin{pmatrix} ax+by \\ cx+dy \end{pmatrix} =  \begin{pmatrix} a & b \\ c &  d \end{pmatrix} \begin{pmatrix} x \\ y \end{pmatrix} (\#eq:example-ch15b)
\end{equation}

Equation \@ref(eq:example-ch15) is an example of a *coupled* system of equations, mainly due to the expression $\displaystyle \frac{dy}{dt}=x+y$. An example of an *uncoupled* system of equations would be $\displaystyle \frac{dx}{dt}=3x$ and $\displaystyle \frac{dy}{dt}=-2y$, which could be solved with separation of variables.\index{differential equation!coupled system}\index{differential equation!uncoupled system}


## Equilibrium solutions
Now let's discuss equilibrium solutions for Equation \@ref(eq:example-ch15). Recall equilibrium solutions are places where both $\displaystyle \frac{dx}{dt}=0 \mbox{ and } \frac{dy}{dt}=0$. Since $\displaystyle \frac{dx}{dt}=2x$, an equilibrium solution would be $x=0$.\index{equilibrium solution}  Substituting $x=0$ into $\displaystyle \frac{dy}{dt}=x+y$ also shows $y=0$ is the corresponding equilibrium solution for $y$. 

For a general linear system of differential equations, it might be helpful to imagine what we should _expect_ for an equilibrium solution. Think back to calculus - what types of functions have a derivative that equals zero?  (Hopefully constant functions comes to mind!) The equilibrium solution is then $x=0$ and $y=0$.

Here is an amazing fact: it turns out **any linear system of differential equations has the origin as its only equilibrium solution.**  One way to verify this fact is to examine the theory behind solutions for linear systems of equations in linear algebra. You might be wondering why there is all the fuss with equilibrium solutions - especially the origin ($x=0$ and $y=0$)^[Another name for the origin equilibrium solution is the _trivial equilibrium_. Can you see why it is trivial?]. So while equilibrium solutions are not a terribly interesting question at the moment, the _stability_ of solutions is. In order to understand what I mean by stability, let's re-examine how to generate phase planes from Chapter \@ref(coupled-06).

## The phase plane
The phase plane is helpful here to understand the stability of an equilibrium solution. Remember that the phase plane shows the motion of solutions, visualized as a vector. For the system we examined earlier let’s take a look at the phase plane. Here is some `R` code from the `demodelr` package to help you visualize the phase plane for Equation \@ref(eq:example-ch15), shown in Figure \@ref(fig:ex-ch15fig).^[It is okay to refer back to Chapter \@ref(coupled-06) for a refresher on how the `phaseplane` command works.]

```{r eval = FALSE}
# For a two variable system of differential equations
# we need to define dx/dt and dy/dt separately:


linear_eq <- c(
  dxdt ~ 2 * x,
  dydt ~ x + y
)

# Now we plot the solution.
phaseplane(
  system_eq = linear_eq,
  x_var = "x",
  y_var = "y"
)
```

```{r ex-ch15fig,fig.cap="phase plane for Equation \\@ref(eq:example-ch15).",echo=FALSE}
# For a two variable system of differential equations
# we need to define dx/dt and dy/dt separately:


linear_eq <- c(
  dxdt ~ 2 * x,
  dydt ~ x + y
)

# Now we plot the solution.
phaseplane(
  system_eq = linear_eq,
  x_var = "x",
  y_var = "y"
) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 14),
    axis.title.x = element_text(size = 14),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.y = element_text(size = 14)
  ) +
  scale_color_colorblind()
```

Notice how in the Figure \@ref(fig:ex-ch15fig) phaseline the arrows seem to spin out from the origin? We are going to discuss that below - but based on what we see we might expect the stability of the origin to be _unstable_. 

## Non-equilibrium solutions and their stability 
Even though we have already identified the equilibrium solution for Equation \@ref(eq:example-ch15), there are other non-equilibrium solutions. Here are two non-equilibrium solutions that you can verify:

- **Solution 1:** $\displaystyle s_{1}(t) = \begin{pmatrix} x \\ y \end{pmatrix} = \begin{pmatrix} 0 \\ e^{t} \end{pmatrix}$
- **Solution 2:** $\displaystyle s_{2}(t) = \begin{pmatrix} x \\ y \end{pmatrix} = \begin{pmatrix} e^{2t} \\ e^{2t} \end{pmatrix}$ 

These two non-equilibrium solutions can aid us in understanding the stability of the equilibrium solutions. All of the solutions contain terms that are exponential growth, indicating movement away from the equilibrium solution at $x=0$, $y=0$. This is even more evident when we plot the solutions with the phase plane by defining a new data frame and plotting $s_{1}(t)$ and $s_{2}(t)$ with `geom_path` (Figure \@ref(fig:ex-ch15fig-sol)):


```{r ex-ch15fig-sol,fig.cap="phase plane for Equation \\@ref(eq:example-ch15), with straight line solutions $s_{1}(t)$ and $s_{2}(t)$.",echo=FALSE,warning = FALSE, message = FALSE}


# Define the straight line solutions:
solution1 <- tibble(
  t = seq(0, 1, length.out = 20),
  x = 0,
  y = exp(t),
  soln = 1
)

solution2 <- tibble(
  t = seq(0, 2, length.out = 20),
  x = exp(2 * t),
  y = exp(2 * t),
  soln = 2
)

solution <- rbind(solution1, solution2)

# Our differential equation from before:
systems_eq <- c(
  dxdt ~ 2 * x,
  dydt ~ x + y
)

# Plot the solution, first saving to a variable:
phaseplane(systems_eq, "x", "y") +
  geom_path(data = solution, aes(x = x, y = y, color = as.factor(soln)), size = 2) +
  xlim(c(-4, 4)) + ylim(c(-4, 4)) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 14),
    axis.title.x = element_text(size = 14),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.y = element_text(size = 14)
  ) +
  scale_color_colorblind(labels = c(expression("s"[1] * "(t)"), expression("s"[2] * "(t)")), name = "Solutions:")
```

The arrows in the phase plane of Figure \@ref(fig:ex-ch15fig-sol) show how $s_{1}(t)$ and $s_{2}(t)$ move away from the solution. The phase plane suggests for Equation \@ref(eq:example-ch15) that the equilibrium solution at the origin is unstable because _both_ the arrows in both directions seem to be pointing away from the origin. 

Also notice that in Figure \@ref(fig:ex-ch15fig-sol) the solutions $s_{1}(t)$ and $s_{2}(t)$ in the $xy$ plane are straight lines!  It turns out that these straight line solutions are quite useful - we will study them in a later chapter.

We can also investigate stability _algebraically_ for each solution ($s_{1}(t)$ and $s_{2}(t)$). We will organize our solutions in vector format, factoring out the exponential functions in each of the expressions:

- Solution 1: $\displaystyle  \vec{s}_{1}(t) = \begin{pmatrix} 0 \\ e^{t} \end{pmatrix}= \begin{pmatrix} 0 \\ e^{t} \end{pmatrix} =e^{t}  \begin{pmatrix} 0 \\ 1  \end{pmatrix}$
- Solution 2:  $\displaystyle \vec{s}_{2}(t) = \begin{pmatrix} e^{2t} \\ e^{2t} \end{pmatrix}= \begin{pmatrix} e^{2t} \\ e^{2t} \end{pmatrix} = e^{2t}  \begin{pmatrix} 1 \\ 1\end{pmatrix}$ 

The vectors $\displaystyle  \begin{pmatrix} 0 \\ 1\end{pmatrix}$ and $\displaystyle  \begin{pmatrix} 1 \\ 1\end{pmatrix}$ are the lines $x=0$ and $y=x$, as shown in Figure \@ref(fig:ex-ch15fig-sol). By factoring out the exponential functions we can see the straight line solutions!  

To further investigate stability we investigate the long term behavior of the exponential functions that are multiplying the straight line solutions. Notice that $\displaystyle \lim_{t \rightarrow \infty}   e^{t}$ and $\displaystyle \lim_{t \rightarrow \infty}   e^{2t}$ *both* do not have a finite value^[In other words, $\lim_{t \rightarrow \infty}  e^{t}=\infty$ and $\lim_{t \rightarrow \infty}  e^{2t}=\infty$.], so we conclusively classify the equilibrium solution as “unstable".\index{equilibrium solution!unstable} Conversely, if *both* of the exponential functions had exponential decrease we would classify the equilibrium solution as "stable". \index{equilibrium solution!stable}

Finally, with these straight line solutions we can create other solutions as a linear combination of $s_{1}(t)$ and $s_{2}(t)$. For example, we can define another solution which we will call $s_{3}(t)$, where $\vec{s}_{3}(t)=\vec{s}_{1}(t) -0.1 \vec{s}_{2}(t)$:


```{r ex-ch15fig-sol-2,fig.cap="Phase plane for Equation \\@ref(eq:example-ch15), with straight line solutions $s_{1}(t)$, $s_{2}(t)$, and $s_{3}(t)$.",,warning = FALSE, message = FALSE,echo=FALSE}

# Define the straight line solutions:
solution1 <- tibble(
  t = seq(0, 2, length.out = 20),
  x = 0,
  y = exp(t),
  soln = 1
)

solution2 <- tibble(
  t = seq(0, 2, length.out = 20),
  x = exp(2 * t),
  y = exp(2 * t),
  soln = 2
)

solution3 <- solution2 %>%
  mutate(
    x = solution1$x - 0.1 * x,
    y = solution1$y - 0.1 * y,
    soln = 3
  )



solution <- rbind(solution1, solution2, solution3)

# Our differential equation from before:
systems_eq <- c(
  dxdt ~ 2 * x,
  dydt ~ x + y
)

# Plot the solution, first saving to a variable:
phaseplane(systems_eq, "x", "y") +
  geom_path(data = solution, aes(x = x, y = y, color = as.factor(soln)), size = 2) +
  xlim(c(-4, 4)) + ylim(c(-4, 4)) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 14),
    axis.title.x = element_text(size = 14),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.y = element_text(size = 14)
  ) +
  scale_color_colorblind(
    labels = c(
      expression("s"[1] * "(t)"),
      expression("s"[2] * "(t)"),
      expression("s"[3] * (t))
    ),
    name = "Solutions:"
  )
```


So to recap the following about straight line solutions to two-dimensional linear systems:

- Straight line solutions have the form $\displaystyle \vec{s}(t) = e^{\lambda t} \cdot \vec{v}$. Methods to determine $\lambda$ and $\vec{v}$ will be studied in later chapters.

- For a two-dimensional linear system, you generally will have two straight line solutions $\vec{s}_{1}$ and $\vec{s}_{2}$. This means you will have two different values of $\lambda$ ($\lambda_{1}$ and $\lambda_{2}$). The most general solution to the system of differential equations is the linear sum of the $s_{1}(t)$ and $s_{2}(t)$:
$\vec{x}(t) = c_{1} \cdot \vec{s}_{1}(t) + c_{2} \cdot \vec{s}_{2}(t)$.
- If both values of $\lambda$ are *greater* than 0, the equilibrium solution is *unstable*.
- If both values of $\lambda$ are *less* than 0, the equilibrium solution is *stable*.
- Geometrically these straight line solutions are lines that pass through the origin in the $xy$ plane.

In the exercises you will look at additional examples to understand the behavior of linear systems.


## Exercises

<!-- Used code coeff_cal in rScripts/15-linear-systems-code -->
```{exercise matrix}
Write the following systems of equations in matrix notation ($\displaystyle \frac{ d \vec{x} }{dt} = A \vec{x}$):

a. $\displaystyle \frac{dx}{dt} = 2x-6y, \;  \frac{dy}{dt} = x-2y$
b. $\displaystyle \frac{dx}{dt} = 9x-22y, \;  \frac{dy}{dt} = 3x-7y$
c. $\displaystyle \frac{dx}{dt} = 4x - 2y, \;  \frac{dy}{dt} = 2x - 2y$
d. $\displaystyle \frac{dx}{dt}= 4x-15y, \; \frac{dy}{dt}=2x-7y$
e. $\displaystyle \frac{dx}{dt} = 3x-18y, \;  \frac{dy}{dt} = x-5y$
f. $\displaystyle \frac{dx}{dt} = 5x-12y, \;  \frac{dy}{dt} = x-2y$

```


```{exercise}
Verify that $\displaystyle s_{1}(t) =  \begin{pmatrix} 0 \\ e^{t} \end{pmatrix}$ and  $\displaystyle s_{2}(t) = \begin{pmatrix} e^{2t} \\ e^{2t} \end{pmatrix}$ are both solutions for Equation \@ref(eq:example-ch15).

```


```{exercise}
Verify that $\displaystyle s_{3}(t) = \begin{pmatrix} x \\ y \end{pmatrix} = \begin{pmatrix} -0.1e^{2t} \\ e^{t} - 0.1e^{2t} \end{pmatrix}$ is a solution for Equation \@ref(eq:example-ch15).

```


```{exercise}
Generate a new solution for Equation \@ref(eq:example-ch15) that is a linear combination of $s_{1}(t)$ and $s_{2}(t)$ and plot your solution with the phase plane for Equation \@ref(eq:example-ch15).
```


```{exercise}
Verify that $x=0$, $y=0$, and $z=0$ are solutions to the differential equation


\begin{equation}
\begin{split}
\frac{dx}{dt} &= 5x-4y + z \\ 
\frac{dy}{dt} &= y - 9z \\
\frac{dz}{dt} &= 7x-z \\
\end{split}
\end{equation}

```



```{exercise}
Explain why we call $x=0$ and $y=0$ _equilibrium_ solutions to the general linear differential equation $\displaystyle \frac{ d \vec{x} }{dt} = A \vec{x}$. In other words, why is the word _equilibrium_ important?  (Hint: Think about what the solution curves for $x$ and $y$ would be in this case.)
```




```{exercise}
Generate a phase plane for each of the systems in Exercise \@ref(exr:matrix) and classify the stability of the equilibrium solution at $x=0$ and $y=0$ as stable, unstable, or uncertain, providing reasons for your conclusion.

```


```{exercise}
This problem considers the differential equation 

\begin{equation}
\begin{split}
\frac{dx}{dt} &= x+y \\ 
\frac{dy}{dt} &= y-x
\end{split}
\end{equation}


a. Use the command `phaseplane` to create a phase plane of this differential equation.
b. Using the option `plot_points`, change the number of arrows shown to 5 and 20 (2 different plots). What do you notice about the updated phase plane?
c. Change the viewing window (`x_window` and `y_window`) from the default to minus 10 to 10 in both axes. Now change the number of arrows shown to 5 and 20 (2 different plots). What do you notice about the updated phase plane?

```


<!-- Adapted from LW pg 164 -->
<!-- tr = -a det = 1 -->
```{exercise}
Consider the following differential equation:
  
\begin{equation}
\begin{split}
\frac{dx}{dt} &= -ax-y \\ 
\frac{dy}{dt} &= x
\end{split}
\end{equation}


a. Write this system in the form $\displaystyle \frac{d\vec{x}}{dt}=A \vec{x}$.
b. Let $a= -3, \; -1, \; 0, \; 1, \; 3$. Generate a phase plane for each of these values of $a$.
c. With each of your phase plane plots, characterize the behavior of the equilibrium solution as $a$ changes.

```








<!-- %\marginnote{Exercises a-c,e Based on Example 4.9 page 166 in \lw. d on 1c, pg 163 in lw}  -->



```{exercise}
Consider the following differential equation:

\begin{equation}
\begin{split}
  \frac{dx}{dt} &= -y \\ 
  \frac{dy}{dt} &= x
\end{split}
\end{equation}



a. Generate a phase plane diagram of this system. What do you notice?
b. Verify that $x(t)=A \cos(t)$ and $y(t)=A \sin(t)$ is a solution to this differential equation.
c. An equation of a circle of radius $R$ is $x^{2}+y^{2}=R^{2}$. Use implicit differentiation to differentiate this equation to get an expression for $\displaystyle \frac{dy}{dx}$. How does your solution compare to the ratio of $\displaystyle \frac{y'}{x'}$ from your differential equation? (Note: $\displaystyle y' = \frac{dy}{dt}$ and $\displaystyle x' = \frac{dx}{dt}$)
d. Verify that $x_{2}(t)=A \cos(t) + B \sin(t)$ and $y_{2}(t)=A \sin(t)-B \cos(t)$ also is a solution to the differential equation. Choose $A=1$ and $B=1$ to make a parametric plot of the solution. What do you notice in your parametric plot?

```
