# Systems of nonlinear equations {#nonlinear-16}

In Section \@ref(linearsystems-15) we discussed systems of linear equations.  For this section we focus on _non_-linear systems of equations.  Here is the good news: the processes and approaches that we will apply for nonlinear systems are similar to linear equations.  While we previously discussed coupled (nonlinear) systems of equations in Section \@ref(coupled-06), it helps to have a refresher with an example.

Consider the following non-linear system of equations:

\begin{equation}
\begin{split} 
\frac{dx}{dt} &= y-1 \\  
\frac{dy}{dt} &= x^{2}-1 
\end{split} (\#eq:ex1-ch16)
\end{equation}

The nullclines for this system are $y=1$ and $x=\pm1$, leading to two equilibrium solutions when we create the phaseplane:

```{r phase-ex1-16,fig.width=4,fig.height=3,fig.cap="Phaseplane for $x'=y-1$ and $y'=x^{2}-1$."  }
# Define the range we wish to evaluate this vector field

system_eq <- c(
  dx ~ y - 1,
  dy ~ x^2 - 1
)

phaseplane(system_eq, "x", "y", x_window = c(-2, 2), y_window = c(-2, 2))
```

Wow!  The phaseplane in Figure \@ref(fig:phase-ex1-16) looks really interesting.  In the upper left corner there is some swirling action, so let's zoom in somewhat:

```{r zoom-ch16,fig.width=4,fig.height=3,fig.cap="Zoomed in phaseplane for $x'=y-1$ and $y'=x^{2}-1$."  }
# Define the range we wish to evaluate this vector field

system_eq <- c(
  dx ~ y - 1,
  dy ~ x^2 - 1
)

phaseplane(system_eq, "x", "y", x_window = c(-1.5, -0.5), y_window = c(0.5, 1.5)) +
  geom_point(data = tibble(xint = c(-1), yint = c(1)), aes(x = xint, y = yint), color = "red", size = 4)
```

Something interesting is happening at the point $(-1,1)$ in Figure \@ref(fig:zoom-ch16).  Let's take a look if we evaluate this point in our differential equation:

\begin{equation}
\begin{split} 
\frac{dx}{dt} &= 1-1 = 0 \\ 
\frac{dy}{dt} &= (-1)^{2}-1  = 0
\end{split}
\end{equation}

Aha!  So the point $(-1,1)$ is an equilibrium solution.  In later sections we will discuss _why_ we are observing the behavior with the swirling arrows. For now, the key point from Figure \@ref(fig:zoom-ch16) is to recognize that _nonlinear systems_ have equilibrium solutions as well.

Something else is interesting in the upper right corner of Figure \@ref(fig:phase-ex1-16).  Let's zoom in near the point $(1,1)$:

```{r zoom-ch16a,fig.width=4,fig.height=3,fig.cap="Another zoomed in phaseplane for $x'=y-1$ and $y'=x^{2}-1$."  }
# Define the range we wish to evaluate this vector field
system_eq <- c(
  dx ~ y - 1,
  dy ~ x^2 - 1
)

phaseplane(system_eq, "x", "y", x_window = c(0.5, 1.5), y_window = c(0.5, 1.5)) +
  geom_point(data = tibble(xint = c(1), yint = c(1)), aes(x = xint, y = yint), color = "red", size = 4)
```

It seems like there is a _second_ equilibrium solution at the point $(1,1)$!  Let's confirm this:

\begin{equation}
\begin{split} 
\frac{dx}{dt} &= 1-1 = 0 \\ 
\frac{dy}{dt} &= (1)^{2}-1  = 0
\end{split}
\end{equation}

We learned something important about nonlinear systems compared to linear systems.  In Section \@ref(linearsystems-15) we learned that the origin is the sole equilibrium solution for a linear system of differential equations.  On the other hand, there may be _multiple_ equilibrium solutions for nonlinear systems of equations.

## Determining equilibrium solutions
To determine an equilbrium solutions for a system of differential equations we first need to find the intersection of different nullclines. We do this by setting each of the rate equations ($\displaystyle \frac{dx}{dt}$ or $\displaystyle \frac{dy}{dt}$) equal to zero. Equation \@ref(eq:ex1-ch16) has two nullclines:

\begin{equation}
\begin{split} 
\frac{dx}{dt} = 0 &\rightarrow y-1 = 0\\  
\frac{dy}{dt} = 0 & \rightarrow x^{2}-1 = 0
\end{split} (\#eq:ex1-ch16-null)
\end{equation}

So solving for both nullclines in Equation \@ref(eq:ex1-ch16-null) we have that $y=1$ or $x = \pm 1$.  You can visually see the phaseplane with the nullclines in Figure \@ref(fig:ex16-phaseplane).

```{r ex16-phaseplane,fig.width=4,fig.height=3,echo=FALSE,fig.cap="Phaseplane for $x'=y-1$ and $y'=x^{2}-1$."  }
# Define the range we wish to evaluate this vector field

system_eq <- c(
  dx ~ y - 1,
  dy ~ x^2 - 1
)

phaseplane(system_eq, "x", "y", x_window = c(-2, 2), y_window = c(-2, 2)) + # The values in quotes are the labels for the axes
  stat_function(fun = function(x) 1, geom = "line", aes(colour = "x' = 0"), size = 1) +
  geom_vline(data = tibble(xint = c(-1, 1)), aes(xintercept = xint, colour = "y' = 0"), size = 1) + geom_point(data = tibble(xint = c(-1, 1), yint = c(1, 1)), aes(x = xint, y = yint), color = "red", size = 4) +
  scale_colour_discrete(name = "Nullclines")
```

Remember: equilibrium solutions occur when two _different_ nullclines intersect.

Let's try another example:

\begin{equation}
\begin{split} 
\frac{dx}{dt} &= x-0.5yx \\  
\frac{dy}{dt} &= yx -y^{2} 
\end{split} (\#eq:ex2-ch16)
\end{equation}

Figure \@ref(fig:ex16-2-phaseplane) shows the phaseplane for this example.  Can you guess where an equilibrium solution would be?

```{r ex16-2-phaseplane,fig.width=4,fig.height=3,echo=FALSE,fig.cap="Phaseplane for $x'=x-0.5xy$ and $y'=yx=y^{2}$."  }
# Define the range we wish to evaluate this vector field
system_eq <- c(
  dx ~ x - 0.5 * y * x,
  dy ~ y * x - y^2
)

phaseplane(system_eq, "x", "y", x_window = c(0, 4), y_window = c(0, 4))
```

Let's find the equations for the nullclines:

\begin{equation}
\begin{split} 
\frac{dx}{dt} = 0 & \rightarrow x-0.5yx = 0 \\  
\frac{dy}{dt} = 0 & \rightarrow yx -y^{2} = 0
\end{split}
\end{equation}

The algebra is becoming a little more involved.  Factoring $x-0.5yx = 0$ we have $x \cdot (1 - 0.5 y) = 0$, so either $x=0$ or $y=2$.  Factoring the second equation we have $y \cdot (x - y) = 0$, so either $y=0$ or $x=y$.  Notice how this second nullcline is a function of $x$ and $y$.

For Equation \@ref(eq:ex2-ch16) the equilibrium solutions are $(0,0)$, $(2,2)$.  You may be tempted to think that $(0,2)$ is also an equilibrium solution - however - $x=0$ and $y=2$ are equations for the $x$ nullcline.  It is easy to forget, but equilibrium solutions are determined from the intersection of _distinct_ nullclines.


## Stability of an equilibrium solution
The idea of stability of an equilibrium solution for a nonlinear system is intuitively similar to that of a linear system: the equilibrium is stable when all the phaseplane arrows point towards the equilibrium solution.  For Equation \@ref(eq:ex2-ch16), the equilibrium solution at $(1,1)$ is _unstable_ because some of the arrows point towards the equilibrium solution, whereas others point away from it.  


For the moment, we won't go into more specifics of how to classify a stable versus unstable equilibrium solution. If you can recognize from the phaseplane equilibrium solutions that appear to be stable (versus those aren't), then you have established some good intuition that can be confirmed with additional analyses. We will study how to do these analyses in the next few sections.


\newpage

## Exercises
```{exercise}
Consider the following nonlinear system of equations:
\begin{equation}
\begin{split}
\frac{dx}{dt} &= x - .5xy  \\
\frac{dy}{dt} &= .5yx-y
\end{split}
\end{equation}


a. What are the equations for the nullclines for this differential equation?
b. What are the equilibrium solutions for this differential equation?
c. Generate a phaseplane that includes all equilibrium solutions.
d. Based on the phaseplane, evaluate the stability of the equilibrium solution.


```

&nbsp;
<!-- LW pg 157 158 #3 -->
```{exercise}
(Inspired by @logan_mathematical_2009) A population of fish has natural predators. A model that describes this interaction is the following:

\begin{equation}
\begin{split}
\frac{dN}{dt} &= N - .3NP  \\
\frac{dP}{dt} &= .5NP - P ,
\end{split}
\end{equation}



a. What are the equations for the nullclines for this differential equation?
b. What are the equilibrium solutions for this differential equation?
c. Generate a phaseplane that includes all the equilibrium solutions.
d. Based on the phaseplane, evaluate the stability of the equilibrium solution.

```
&nbsp;


```{exercise}
Consider the following system:

\begin{equation}
\begin{split}
\frac{dx}{dt} &= y^{2} \\
\frac{dy}{dt} &= -\frac{2}{3} x,
\end{split}
\end{equation}


a. What are the nullclines for this system of equations?
b. What is the equilibrium solution for this system of equations?
c. Generate a phaseplane that includes the equilibrium solution.
d. Based on the phaseplane, evaluate the stability of the equilibrium solution.

```

&nbsp;

```{exercise}
The *Van der Pol Equation* is a second-order differential equation used to study radio circuits: $x'' + \mu \cdot (x^{2}-1) x' + x = 0$, where $\mu$ is a parameter.


a. Let $y'=x$. Show that with this change of variables the Van de Pol equation can be written as a system: 
\begin{equation}
\begin{split} 
\frac{dx}{dt} &= y \\ 
\frac{dy}{dt} &= -x-\mu (x^{2}-1)y
\end{split}
\end{equation}
b. Verify that the only equilibrium solution is $(0,0)$.
\item Set $\mu=1$. Generate a phaseplane that includes the equilibrium solution.
c. Based on the phaseplane, evaluate the stability of the equilibrium solution.


```

&nbsp;
```{exercise}
Consider the following nonlinear system:

\begin{equation}
\begin{split}
\frac{dx}{dt} &= x-y \\
\frac{dy}{dt} &=-y + \frac{5x^2}{4+x^{2}},
\end{split}
\end{equation}


a. What are the two equations for the nullclines?
b. Using desmos (or some other graphing utility), graph the two nullclines simultaneous.  What are the intersection points?
c. Generate a phaseplane for this system that contains all the equilibrium solutions.
d. Let's say instead that $\displaystyle \frac{dx}{dt} = bx-y$, where $b$ is a parameter such that $0 \leq b \leq 2$. How many equilibrium solutions do you have as $b$ changes?


```
&nbsp;


```{exercise}
(Inspired by @logan_mathematical_2009) Let $C$ be the amount of carbon in a forest ecosystem, with $P$ be the rate of increase due to photosynthesis.  Herbivores $H$ consume carbon on the following predator-prey model:

\begin{equation}
\begin{split}
\frac{dC}{dt}&=- aC - bHC \\
\frac{dH}{dt} &= ebHC-dC
\end{split}
\end{equation}


a. What are the equations for the nullclines?
b. Determine the equilibrium solutions.

```




