# Systems of nonlinear equations

In the last section we discussed systems on linear equations.  Here we are going to focus on _non_-linear systems of equations.  Here is the good news: the processes and approaches that we will take are similar to linear equations.  We have already discussed coupled (nonlinear) systems of equations in Section \@ref(coupled-06), but it helps to have a refresher. Let's start right away with an example.

Consider the following non-linear system of equations:

\begin{equation}
\begin{split} 
\frac{dx}{dt} &= y-1 \\  (\#eq:ex1-ch16)
\frac{dy}{dt} &= x^{2}-1 
\end{split}
\end{equation}

Let's examine the phaseplane for this system:
The nullclines for this system are $y=1$ and $x=\pm1$, leading two two equilibrium solutions when we create the phaseplane:

```{r,fig.width=4,fig.height=3,fig.cap="Phaseplane for Equation \\ref{eq:ex1-ch16}."  }
# Define the range we wish to evaluate this vector field


system_eq <- c(dx ~ y-1,
               dy ~ x^2-1)

phaseplane(system_eq,'x','y',x_window = c(-2,2),y_window = c(-2,2))
  

```

Wow!  This phaseplane looks really interesting.  In the upper left there is some swirling action, so let's zoom in somewhat:

```{r zoom-ch16,fig.width=4,fig.height=3,fig.cap="Zoomed in phaseplane for Equation \\ref{eq:ex1-ch16}."  }
# Define the range we wish to evaluate this vector field


system_eq <- c(dx ~ y-1,
               dy ~ x^2-1)

phaseplane(system_eq,'x','y',x_window = c(-1.5,-0.5),y_window = c(0.5,1.5)) +
  geom_point(data=tibble(xint=c(-1),yint=c(1)),aes(x=xint,y=yint),color='red',size=4)
  

```

Something interesting is happening at the point $(-1,1)$ in Figure \@ref(fig:zoom-ch16).  Let's take a look if we evaluate this point in our differential equation:

\begin{equation}
\begin{split} 
\frac{dx}{dt} &= 1-1 = 0 \\ 
\frac{dy}{dt} &= (-1)^{2}-1  = 0
\end{split}
\end{equation}

Aha!  So the point $(-1,1)$ is an equilibrium solution.  In later sections we will discuss _why_ we are observing the behavior that we are, but for now the key point is to know that _nonlinear systems_ have equilibrium solutions as well.

Something else is interesting in the upper right as well.  Let's zoom in near the point $(1,1)$:

```{r zoom-ch16a,fig.width=4,fig.height=3,fig.cap="Zoomed in phaseplane for Equation \\ref{eq:ex1-ch16}."  }
# Define the range we wish to evaluate this vector field


system_eq <- c(dx ~ y-1,
               dy ~ x^2-1)

phaseplane(system_eq,'x','y',x_window = c(0.5,1.5),y_window = c(0.5,1.5)) +
  geom_point(data=tibble(xint=c(1),yint=c(1)),aes(x=xint,y=yint),color='red',size=4)
  

```

It seems like there is a _second_ equilibrium solution at the point $(1,1)$!  Let's confirm this:

\begin{equation}
\begin{split} 
\frac{dx}{dt} &= 1-1 = 0 \\ 
\frac{dy}{dt} &= (1)^{2}-1  = 0
\end{split}
\end{equation}

We learned something important about nonlinear systems compared to linear systems.  In Section \@ref(linearsystems-15) we learned that the origin is the sole equilibrium solutino for a linear system of differential equations.  On the other hand, there may be _multiple_ equilibrium solutions for nonlinear systems of equations.

## Determining equilibrium solutions
To determine an equilbrium solutions we need to find the intersection of different nullclines. We do this by setting each of the rate equations ($\displaystyle \frac{dx}{dt}$ or $\displaystyle \frac{dy}{dt}$) equal to zero. Equation \@ref(eq:ex1-ch16) has two nullclines:

\begin{equation}
\begin{split} 
\frac{dx}{dt} = 0 &\rightarrow y-1 = 0\\  (\#eq:ex1-ch16)
\frac{dy}{dt} = 0 & \rightarrow x^{2}-1 = 0
\end{split}
\end{equation}

So solving for both nullclines we have that $y=1$ or $x = \pm 1$.  You can visually see the phaseplane with the nullclines in Figure \@ref(fig:ex16-phaseplane).

```{r ex16-phaseplane,fig.width=4,fig.height=3,echo=FALSE,fig.cap="Phaseplane for Equation \\ref{eq:ex1-ch16}."  }
# Define the range we wish to evaluate this vector field


system_eq <- c(dx ~ y-1,
               dy ~ x^2-1)

phaseplane(system_eq,'x','y',x_window = c(-2,2),y_window = c(-2,2))+  # The values in quotes are the labels for the axes
stat_function(fun=function(x)1, geom="line", aes(colour="x' = 0"),size=1) +
  geom_vline(data=tibble(xint=c(-1,1)),aes(xintercept=xint,colour="y' = 0"),size=1) + geom_point(data=tibble(xint=c(-1,1),yint=c(1,1)),aes(x=xint,y=yint),color='red',size=4) +
  scale_colour_discrete(name='Nullclines')
  

```

Remember: equilibrium solutions occur when two _different_ nullclines intersect.

Let's another example:

\begin{equation}
\begin{split} 
\frac{dx}{dt} &= x-0.5yx \\  (\#eq:ex2-ch16)
\frac{dy}{dt} &= yx -y^{2} 
\end{split}
\end{equation}

Figure \@ref(fig:ex16-2-phaseplane) shows the phaseplane for this example.  Can you guess where an equilibrium solution would be?

```{r ex16-2-phaseplane,fig.width=4,fig.height=3,echo=FALSE,fig.cap="Phaseplane for Equation \\ref{eq:ex2-ch16}."  }
# Define the range we wish to evaluate this vector field


system_eq <- c(dx ~ x-0.5*y*x,
               dy ~ y*x-y^2)

phaseplane(system_eq,'x','y',x_window = c(0,4),y_window = c(0,4))
  

```

Let's find the equations for the nullclines:

\begin{equation}
\begin{split} 
\frac{dx}{dt} = 0 & \rightarrow x-0.5yx = 0 \\  
\frac{dy}{dt} = 0 & \rightarrow yx -y^{2} = 0
\end{split}
\end{equation}

The algebra is becoming a little more involved.  Factoring $x-0.5yx = 0$ we have $x \cdot (1 - 0.5 y) = 0$, so either $x=0$ or $y=2$.  Factoring the second equation we have $y \cdot (x - y) = 0$, so either $y=0$ or $x=y$.  Notice how this second nullcline is a function of $x$ and $y$.

For this case, the equilibrium solutions are $(0,0)$, $(2,2)$.  You may be tempted to think that $(0,2)$ is also an equilibrium solution - however - $x=0$ and $y=2$ are equations for the $x$ nullcline.  It is easy to forget, but equilibrium solutions are determined from the intersection of _distinct_ nullclines.


## Stability of an equilibrium solution
The idea of stability of an equilibrium solution for a nonlinear system is intuitively similar to that of a linear system: the equilibrium is stable when all the phaseplane arrows point towards the equilibrium solution.  For Equation \@ref(eq:ex1-ch16), the equilibrium solution at $(1,1)$ is _unstable_ because some of the arrows point towards the equilibrium solution, whereas others point away from it.  


For the moment, we won't go into more specifics of how to classify a stable versus unstable equilibrium solution. If you can recognize equilibrium solutions that appear to be stable (versus those that do not, then you are well on your way to analyze these systems.



\newpage

## Exercises
```{exercise}
Consider the following nonlinear system of equations:
\begin{equation}
\begin{split}
\frac{dx}{dt} &= x - .5xy  \\
\frac{dy}{dt} &= .5yx-y
\end{split}
\end{equation}

\begin{enumerate}[label=\alph*.]
\item What are the equations for the nullclines for this differential equation?
\item What are the equilibrium solutions for this differential equation?
\item Generate a phaseplane that includes all equilibrium solutions.
\item Based on the phaseplane, evaluate the stability of the equilibrium solution.
\end{enumerate}

```

&nbsp;
<!-- LW pg 157 158 #3 -->
```{exercise}
A population of fish has natural predators. A model that describes this interaction is the following:

\begin{equation}
\begin{split}
\frac{dN}{dt} &= N - .3NP  \\
\frac{dP}{dt} &= .5NP - P ,
\end{split}
\end{equation}


\begin{enumerate}[label=\alph*.]
\item What are the equations for the nullclines for this differential equation?
\item What are the equilibrium solutions for this differential equation?
\item Generate a phaseplane that includes all the equilibrium solutions.
\item Based on the phaseplane, evaluate the stability of the equilibrium solution.
\end{enumerate}
```
&nbsp;


```{exercise}
Consider the following system:

\begin{equation}
\begin{split}
\frac{dx}{dt} &= y^{2} \\
\frac{dy}{dt} &= -\frac{2}{3} x,
\end{split}
\end{equation}

\begin{enumerate}[label=\alph*.]
\item What are the nullclines for this system of equations?
\item What is the equilibrium solution for this system of equations?
\item Generate a phaseplane that includes the equilibrium solution.
\item Based on the phaseplane, evaluate the stability of the equilibrium solution.
\end{enumerate}
```

&nbsp;

```{exercise}
The *Van der Pol Equation* is a second-order differential equation used to study radio circuits: $x'' + \mu \cdot (x^{2}-1) x' + x = 0$, where $\mu$ is a parameter.

\begin{enumerate}[label=\alph*.]
\item Let $y'=x$. Show that with this change of variables the Van de Pol equation can be written as a system: 
\begin{equation}
\begin{split} 
\frac{dx}{dt} &= y \\ 
\frac{dy}{dt} &= -x-\mu (x^{2}-1)y
\end{split}
\end{equation}
\item Verify that the only equilibrium solution is $(0,0)$.
\item Set $\mu=1$. Generate a phaseplane that includes the equilibrium solution.
\item Based on the phaseplane, evaluate the stability of the equilibrium solution.
\end{enumerate}

```

&nbsp;
```{exercise}
Consider the following nonlinear system:

\begin{equation}
\begin{split}
\frac{dx}{dt} &= x-y \\
\frac{dy}{dt} &=-y + \frac{5x^2}{4+x^{2}},
\end{split}
\end{equation}

\begin{enumerate}[label=\alph*.]
\item What are the two equations for the nullclines?
\item Using desmos, or some other graphing utility, graph the two nullclines simultaneous.  What are the intersection points?
\item Generate a phaseplane for this system that contains all the equilibrium solutions.
\item Let's say instead that $\displaystyle \frac{dx}{dt} = bx-y$, where $b$ is a parameter such that $0 \leq b \leq 2$. How many equilibrium solutions do you have as $b$ changes?
\end{enumerate}

```
&nbsp;


```{exercise}
Let $C$ be the amount of carbon in a forest ecosystem, with $P$ be the rate of increase due to photosynthesis.  Herbivores $H$ consume carbon on the following predator-prey model:
    \begin{align*}
\frac{dC}{dt}&=- aC - bHC \\
\frac{dH}{dt} &= ebHC-dC
\end{align*}

\begin{enumerate}
\item What are the equations for the nullclines?
\item Determine the equilibrium solutions.
\end{enumerate}
```




