# Local Linearization and the Jacobian {#sec-jacobian-17}

Chapters [-@sec-linearsystems-15] and [-@sec-nonlinear-16] focused on systems of differential equations and using phase planes to determine a preliminary classification of any equilibrium solution. In this chapter we study local linearization and the associated Jacobian matrix.\index{local linearization}\index{Jacobian}  These tools are used to analyze stability of equilibrium solutions for a nonlinear system, thereby building a bridge between nonlinear and linear systems of equations. Let's get started!

::: {.callout-note}
This chapter uses following libraries: `ggplot2`, and `tibble` (all part of the `tidyverse`),`demodelr`, as well as the `stats` library (which should be automatically loaded when `R` begins). See @sec-packages-02 on how to install packages.  Prior to running any code in this section, include `library(tidyverse)` and `library(demodelr)`. 
:::


## Competing populations
Let's take a look at a familiar example from @sec-likelihood-09, specifically the experiments of growing different species of yeast together [@gause_experimental_1932]. @eq-yeast-comp-17 (adapted from the one presented in @gause_experimental_1932) represents the volume of two species of yeast (which we will call $Y$ and $N$) growing in the same solution:

$$
\begin{split}
\frac{dY}{dt} &= .2 Y \left( \frac{13-Y-2N}{13} \right)  \\  
\frac{dN}{dt} &= .06 N \left( \frac{6-N-.4Y}{6} \right)  \\ 
\end{split}
$$ {#eq-yeast-comp-17}

While @eq-yeast-comp-17 is a tricky model to consider, the terms $2N$ and $.4Y$ represent the effect that $Y$ and $N$ have on each other since they are competing for the same resource. If these terms weren't present, both $Y$ and $N$ would follow a logistic growth (verify this on your own).

@eq-yeast-comp-17 has 4 equilibrium solutions: $(Y,N)=(0,0)$, $(Y,N)=(13,0)$, $(Y,N)=(0,6)$, and $(Y,N)=(5,4)$ (@exr-yeast-ss-17). @fig-yeast-orig-17 shows the phase plane for this system along with the equilibrium solutions.^[I encourage you to use the `phaseplane` command to re-create @fig-yeast-orig-17. If you do, be sure to use the options `x_window` and `y_window` to set things correctly.]

```{r,fig.cap="Phase plane for @eq-yeast-comp-17, with equilibrium solutions shown as red points.",echo=FALSE}
#| label: fig-yeast-orig-17

# Define the range we wish to evaluate this vector field
Y_window <- c(0, 20)
N_window <- c(0, 10)

yeast_eq <- c(
  dY ~ .2 * Y * (13 - Y - 2 * N) / 13,
  dN ~ .06 * N * (6 - N - .4 * Y) / 6
)

# Define the points where the
# equilibrium solutions are located
eq_solns <- tibble(
  Y = c(0, 13, 0, 5),
  N = c(0, 0, 6, 4)
)

# Reminder: The values in quotes are the labels for the axes
# We plot the phase plane and add on the points.
phaseplane(yeast_eq, "Y", "N",
  x_window = Y_window,
  y_window = N_window
) +
  geom_point(
    data = eq_solns,
    aes(x = Y, y = N),
    color = "red",
    size = 2
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 14),
    axis.title.x = element_text(size = 14),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.y = element_text(size = 14)
  ) +
  scale_color_colorblind()
```

Let's take a closer look at the phase plane near the equilibrium solution $(Y,N)=(5,4)$ in @fig-yeast-closer:

```{r,fig.cap="A zoomed in view of @eq-yeast-comp-17 near the $(Y,N)=(5,4)$ equilibrium solution.",echo=FALSE,warning=FALSE}
#| label: fig-yeast-closer

# Define the range we wish to evaluate this vector field
Y_window <- c(4, 6)
N_window <- c(3, 5)

yeast_eq <- c(
  dY ~ .2 * Y * (13 - Y - 2 * N) / 13,
  dN ~ .06 * N * (6 - N - .4 * Y) / 6
)

# Define the points where the
# equilibrium solutions are located
eq_solns <- tibble(
  Y = c(0, 13, 0, 5),
  N = c(0, 0, 6, 4)
)

# Reminder: The values in quotes are the labels for the axes
# We plot the phase plane and add on the points.
phaseplane(yeast_eq, "Y", "N",
  x_window = Y_window,
  y_window = N_window
) +
  geom_point(
    data = eq_solns,
    aes(x = Y, y = N),
    color = "red",
    size = 2
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 14),
    axis.title.x = element_text(size = 14),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.y = element_text(size = 14)
  ) +
  scale_color_colorblind() +
  xlim(Y_window) +
  ylim(N_window)
```

The phase plane in @fig-yeast-closer _looks_ like this equilibrium solution is stable (the arrows seem to suggest a "swirling" into this equilibrium solution. However, another way to verify this is by applying a **locally linear approximation**. Better stated, because @eq-yeast-comp-17 is a system of differential equations, we will construct a **tangent plane approximation** around $Y=5$, $N=4$. Let's review how to do that next.

## Tangent plane approximations
For a multivariable function $f(x,y)$, the tangent plane approximation \index{tangent!plane}  at the point $x=a$, $y=b$ is given by @eq-tangent-plane-17:

$$
L(x,y) = f(a,b) + f_{x}(a,b) \cdot (x-a) + f_{y}(a,b) \cdot (y-b),
$$ {#eq-tangent-plane-17}

where $f_{x}$ is the partial derivative of $f(x,y)$ with respect to $x$ and $f_{y}$ is the partial derivative of $f(x,y)$ with respect to $y$.

We will apply @eq-tangent-plane-17 to @eq-yeast-comp-17 at the equilibrium solution at $(Y,N)=(5,4)$. Since we have *two* equations, we need to compute two tangent plane approximations (one for each equation). The right hand sides for @eq-yeast-comp-17 look complicated, but we can expand them to identify $f(Y,N)=.2Y - .01538 Y^{2}  - .03077 YN$ and $g(Y,N)= .06N-.01 N^{2} - .004YN$.

First consider $f(Y,N)$. Let's compute the partial derivatives for $f(Y,N)$ at the equilibrium solution:


$$
\begin{split}
f_{Y} = .2 - .03076Y - .03077N & \rightarrow f_{Y}(5,4)=-.0769 \\
f_{N} = - .03077 Y  & \rightarrow f_{N}(5,4)=-.1539
\end{split}
$$

We also know that $f(5,4)=0$. As a result, the tangent plane approximation for $f(Y,N)$ is given by @eq-tangent-plane-f-17:


$$
f(Y,N) \approx  -.0769 \cdot (Y-5) -.1539 \cdot (N-4)
$$ {#eq-tangent-plane-f-17}


Likewise if we consider $g(Y,N)= .06N-.01 N^{2} - .004YN$, we have:

$$
\begin{split}
g_{Y} &= -.004N  \rightarrow g_{Y}(5,4)=-.016 \\
g_{N} &= .06 -.02N-.004Y  \rightarrow g_{N}(5,4)=-.04
\end{split}
$$

We also know that $g(5,4)=0$. As a result, the tangent plane approximation for $g(Y,N)$ is given by @eq-tangent-plane-g-17:

$$
g(Y,N) \approx  -.016  \cdot (Y-5) -.04 \cdot (N-4) 
$$ {#eq-tangent-plane-g-17}

So, at the equilibrium solution $(Y,N)=(5,4)$, @eq-yeast-comp-17  behaves like the following system of equations (@eq-yeast-17-54):

$$
\begin{split}
\frac{dY}{dt} &= -.0769 \cdot (Y-5) -.1539 \cdot (N-4) \\
\frac{dN}{dt} &= -.016  \cdot (Y-5) -.04 \cdot (N-4)
\end{split}
$$ {#eq-yeast-17-54}



## The Jacobian matrix
@eq-yeast-17-54 looks like a linear system of equations; however, we first need to re-define our system with the change of variables $y = Y-5$, $n = N-4$, which essentially is a shift of the variables so the equilibrium solution is at the origin (@eq-yeast-17-54-shift): 

$$
\begin{split}
\frac{dy}{dt} &= -.0769 y -.1539 n \\
\frac{dn}{dt} &= -.016  y -.04 n
\end{split}
$$ {#eq-yeast-17-54-shift}

Then we can write @eq-yeast-17-54-shift in matrix form using the tools from @sec-linearsystems-15:

$$
\begin{pmatrix} y' \\ n' \end{pmatrix} =\begin{pmatrix} -.0769 & -.1539 \\ -.016 &  -.04 \end{pmatrix} \begin{pmatrix} y \\ n \end{pmatrix} 
$$ {#eq-yeast-17-jacobian-mat}

We define the matrix $\displaystyle J= \begin{pmatrix} -.0769 & -.1539 \\ -.016 &  -.04 \end{pmatrix}$ as the Jacobian matrix.\index{Jacobian}

More generally, let's say we have the following system of differential equations, with an equilibrium solution at $(x,y)=(a,b)$:

$$
\begin{split}
\frac{dx}{dt} &= f(x,y) \\
\frac{dy}{dt} &= g(x,y)
\end{split}
$$

The _Jacobian matrix_ at that equilibrium solution is:

$$
J_{(a,b)} =\begin{pmatrix} f_{x}(a,b) & f_{y}(a,b) \\ g_{x}(a,b) &  g_{y}(a,b) \end{pmatrix} 
$$ {#eq-jacobian-17}

The notation $J_{(a,b)}$ signifies the Jacobian matrix evaluated at the equilibrium solution $(x,y)=(a,b)$. The Jacobian matrix (@eq-jacobian-17) follows naturally from tangent plane approximations (@eq-tangent-plane-17). In later chapters we will use the Jacobian matrix to investigate stability of nonlinear systems.

The Jacobian matrix is part of the following system of linear equations (@eq-linear-jacobian-17):

$$
\begin{split}
\frac{dX}{dt} &= f_{x}(a,b) \cdot X + f_{y}(a,b) \cdot Y \\
\frac{dY}{dt} &= g_{x}(a,b) \cdot X + g_{y}(a,b) \cdot Y 
\end{split}
$$  {#eq-linear-jacobian-17}

You may recognize that @eq-yeast-17-54 was a specific example of @eq-linear-jacobian-17. The variables $X=x-a$ and $Y=y-b$ help translate the tangent plane equation into a linear system. Also notice how @eq-jacobian-17 does not include the terms $f(a,b)$ or $g(a,b)$ from @eq-tangent-plane-17. This is because of the fact that we are building our tangent plane approximation at an equilibrium solution, so $f(a,b)=g(a,b)=0$! 

One more additional note: when we visualize the phase plane for a Jacobian matrix, we center the window at the origin $(x,y)=(0,0)$ (rather than at the equilibrium solution $(a,b)$) because @eq-linear-jacobian-17 is a linear system, with an equilibrium solution at the origin. While we don't discuss it here, the Jacobian matrix also extends to higher order systems as well.


### An unstable equilibrium solution
Let's return to @eq-yeast-comp-17 and investigate the Jacobian at the origin $(Y,N)=(0,0)$. The Jacobian matrix at that solution is the following:

$$
J_{(0,0)} =\begin{pmatrix} .2 & 0 \\0 &  .06 \end{pmatrix}
$$ {#eq-yeast-00-17}

(You should verify this on your own). This Jacobian matrix leads to an uncoupled system of linear equations where $Y' = .2Y$ and $N'=.06N$. In this case, both $Y$ and $N$ are growing exponentially. This type of behavior supports the idea that both $Y$ and $N$ are growing away from the origin, as indicated in @fig-yeast-orig-17.

The Jacobian matrix helps confirm some of our intuition from examining the phase plane of a system of differential equations. At the heart of the Jacobian matrix is that idea that we can understand the dynamics of a nonlinear system through examining a closely related linear system. @sec-eigenvalues-18 will continue to build on this idea, giving you a tool to quantitatively analyze the stability of an equilibrium solution.

## {{< fa burst >}} Differentiation in R
Computing the Jacobian and partial derivatives in this section is mostly done "by hand".  Technology does exist for symbolic differentiation, especially of complex expression.

In `R` the command `D` paired with the function `expression` can compute partial derivatives.  Consider $\displaystyle f(Y,N)=r_{1}Y \cdot \frac{\left(K_{1} - Y - c_{1}N \right)}{K_{1}}$.  We can calculate $\displaystyle \frac{\partial f}{\partial Y}$ with the following code:

```{r,eval=FALSE}
D(expression(r1 * Y * (K1 - Y - c1*N) / K1), "Y")
```

The resulting expression can also be evaluated with any numeric values of the parameters or variables.  The process we use is:

1. Create a symbolic expression for the partial derivatives.
2. Substitute values of any inputs.
3. If the function is fully specified, then evaluate it numerically.

For example, to evaluate $\displaystyle \frac{\partial f}{\partial Y}$ when $r_{1} = 0.2$, $c_{1}=2$, and $K_{1}=13$, we will have the following:

```{r,eval=FALSE}

# 1: Create a symbolic expression and compute partial derivative with respect to Y:
fY <- D(expression(r1 * Y * (K1 - Y - c1*N) / K1), "Y")

# 2.1 Define inputs values for parameters and variables:
inputs <- list(r1 = 0.2, c1 = 2, K1=13, Y=5, N=4)

# 2.2: Substitute parameter values into the expression
fY_sub <- do.call(substitute, list(fY, inputs))

# 3. Evaluate the result
eval(fY_sub)


```

Notice how the evaluation step defines a list of inputs, and then uses the functions `do.call` and `substitute` to evaluate the expression.  The output `fY_sub` is still an evaluated expression.  Step 3 applies the function `eval` to numerically compute the result.  This last step only works if all parameters and variables are defined.

The downside to symbolic differentiation with `R` is that while sometimes the resulting expression may be correct - it isn't algebraically simplified. Consider for example $\displaystyle f(x,y)=y + \frac{5x^2}{4+x^{2}}$. The value of  $\displaystyle \frac{\partial f}{\partial x}$ is $\displaystyle \frac{40x}{(4+x^{2})^{2}}$ after algebraic simplification with the quotient rule. @exr-quotient-17 will have you explore this further.


Other programs that can do symbolic manipulation include:

- [Mathematica](https://www.wolfram.com/mathematica/?source=wa-header), which includes [Wolfram alpha](https://www.wolframalpha.com/)
- [sageMath](https://www.sagemath.org/)
- [Maple](https://www.maplesoft.com/products/maple/)
- [Maxima](https://maxima.sourceforge.io/)



## Exercises


::: {#exr-yeast-ss-17}
Using algebra, show that the 4 equilibrium solutions to @eq-yeast-comp-17 are $(Y,N)=(0,0)$, $(Y,N)=(13,0)$, $(Y,N)=(0,6)$, and $(Y,N)=(5,4)$ *Hint:* Perhaps first determine the nullclines for each solution.
:::

::: {#exr-yeast-comp-17}
Construct the Jacobian matrices for the equilibrium solutions $(Y,N)=(13,0)$ and $(Y,N)=(0,6)$ to @eq-yeast-comp-17.
:::

::: {#exr-quotient-17}
{{< fa burst >}} Let $\displaystyle f(x,y)=y + \frac{5x^2}{4+x^{2}}$. By hand, verify $\displaystyle \frac{\partial f}{\partial x} = \frac{40x}{(4+x^{2})^{2}}$. Then use `R` to determine $\displaystyle \frac{\partial f}{\partial x}$. Show the algebraic steps to simplify the output from `R` to the simplified form of $\displaystyle \frac{\partial f}{\partial x}$.
:::
 
::: {#exr-symb-r-17}
{{< fa burst >}} Use symbolic differentiation in `R` to verify that when $g(Y,N)= .06N-.01 N^{2} - .004YN$, then $g_{Y}= -.004N$ and $g_{N}=.06 -.02N-.004Y$
:::

::: {#exr-eval-eq-17}
{{< fa burst >}} Consider the following system of differential equations:

$$
\begin{split}
\frac{dY}{dt} &= r_{1}Y \cdot \frac{\left(K_{1} - Y - c_{1}N \right)}{K_{1}} \\
\frac{dN}{dt} &= r_{2}Y \cdot \frac{\left(K_{2} - N - c_{2}Y \right)}{K_{2}}
\end{split}
$$


a. Use `R` to evaluate that this system of differential equations has an equilibrium solution at $(Y,N)=(5,4)$ when $r_{1}=.2$, $r_{2}=.06$, $K_{1}=13$, $K_{2}=6$, $c_{1}=2$, and $c_{2}=.4$.
b. Let $\displaystyle f(Y,N)=r_{1}Y \cdot \frac{\left(K_{1} - Y - c_{1}N \right)}{K_{1}}$. Use symbolic differentiation in `R` to compute $f_{Y}$ and $f_{N}$.
c. Let $\displaystyle g(Y,N)=r_{2}Y \cdot \frac{\left(K_{2} - N - c_{2}Y \right)}{K_{2}}$. Use symbolic differentiation in `R` to compute $g_{Y}$ and $g_{N}$.
d. Use `R` to evaluate $f_{Y}$, $f_{N}$, $g_{Y}$, and $g_{N}$ at the equilibrium solution $(Y,N)=(5,4)$.
e. Using your previous answer, construct the Jacobian matrix at this equilibrium solution. Compare your answer to @eq-yeast-17-jacobian-mat. What do you notice?

:::
 
::: {#exr-verify-hl-17}
By solving directly, show that $(H,L)=(0,0)$ and $(4,3)$ are equilibrium solutions to the following system of equations:

$$
\begin{split}
\frac{dH}{dt} &= .3 H - .1 HL \\
\frac{dL}{dt} &=.05HL -.2L
\end{split}
$$
:::


::: {#exr-which-sys-17}
A system of two differential equations has a Jacobian matrix at the equilibrium solution $(0,0)$ as the following:
  
$$
J_{(0,0)}=\begin{pmatrix} 0 & -1 \\ 1 & 0 \end{pmatrix}
$$

What would be a system of differential equations that would produce that Jacobian matrix?

:::


::: {#exr-nonlin-sys-17}
Consider the following nonlinear system:

$$
\begin{split} 
\frac{dx}{dt} &= y-1 \\ 
\frac{dy}{dt} &= x^{2}-1
\end{split}
$$


a. Verify that this system has equilibrium solutions at $(-1,1)$ and $(1,1)$.
b. Determine the linear system associated with the tangent plane approximation at the equilibrium solution $(x,y)=(-1,1)$ and $(1,1)$ (two separate linear systems).
c. Construct the Jacobian matrix at the equilibrium solutions at $(-1,1)$ and $(1,1)$.
d. With the Jacobian matrix, visualize a phase plane at these equilbrium solutions to estimate stability of the equilibrium solution.


:::



<!-- Strogatz pg 259-260. LW pg 183 #11 - cell differentiation -->

::: {#exr-strogatz-lw-17}
(Inspired by @strogatz_nonlinear_2015) Consider the following nonlinear system:

$$
\begin{split}
\frac{dx}{dt} &= y-x \\
\frac{dy}{dt} &=-y + \frac{5x^2}{4+x^{2}}
\end{split}
$$


a. Verify that the point $(x,y)=(1,1)$ is an equilibrium solution.
b. Determine the linear system associated with the tangent plane approximation at the equilibrium solution $(x,y)=(1,1)$.
c. Construct the Jacobian matrix at this equilibrium solution.
d. With the Jacobian matrix, visualize a phase plane at that equilbrium solution to estimate stability of the equilibrium solution.


:::



::: {#exr-jac-nonlin-17}
Consider the following system:

$$
\begin{split}
\frac{dx}{dt} &= y^{2} \\
\frac{dy}{dt} &= - x
\end{split}
$$


a. Determine the equilibrium solution.
b. Visualize a phase plane of this system of differential equations.
c. Construct the Jacobian at the equilibrium solution.
d. Use the fact that $\displaystyle \frac{dy}{dt} / \frac{dx}{dt} = \frac{dy}{dx}$, which should yield a separable differential equation that will allow you to solve for a function $y(x)$. Plot several solutions of $y(x)$. How does that solution compare to the phase plane from the Jacobian matrix?

:::

::: {#exr-r-diff-nonlin-17}
{{< fa burst >}} When $\displaystyle f(Y,N)=k_{1}Y \cdot \frac{\left(K - Y - c_{1}N \right)}{K}$, use `R` to compute $\displaystyle \frac{\partial f}{\partial N}$.
:::

::: {#exr-vdp-17}
The *Van der Pol Equation* is a second-order differential equation used to study radio circuits. In @sec-nonlinear-16 you showed that the differential equations $x'' + \mu \cdot (x^{2}-1) x' + x = 0$, where $\mu$ is a parameter can be written as a system of equations:

$$
\begin{split} 
\frac{dx}{dt} &= y \\ 
\frac{dy}{dt} &= -x-\mu (x^{2}-1)y
\end{split}
$$

a. Determine the general Jacobian matrix $J_{(x,y)}$ for this system of equations.
b. The point $(0,0)$ is an equilibrium solution. Evaluate the Jacobian matrix at the point $(0,0)$. Your Jacobian matrix will depend on $\mu$.
c. Evaluate your Jacobian matrix at the $(0,0)$ equilibrium solution for different values of $\mu$ ranging from $-3$, $-1$, 0, 1, 3.
d. Make a phase plane for the Jacobian matrices at each of the values of $\mu$.
e. Based on the phase planes that you generate, evaluate the stability of the equilibrium solution as $\mu$ changes.


:::




<!-- Modified from LW pg 157 158 #3 -->

::: {#exr-lw-157-17}
(Inspired by @logan_mathematical_2009) A population of fish $F$  has natural predators $P$. A model that describes this interaction is the following:

$$
\begin{split}
\frac{dF}{dt} &= F - .3FP  \\
\frac{dP}{dt} &= .5FP - P
\end{split}
$$


a. What are the equilibrium solutions for this differential equation?
b. Construct a Jacobian matrix for each of the equilibrium solutions.
c. Based on the phase plane from the Jacobian matrices, evaluate the stability of the equilibrium solutions.

:::


::: {#exr-pastor-17}
(Inspired by @pastor_mathematical_2008) The amount of nutrients (such as carbon) in soil organic matter is represented by $N$, whereas the amount of inorganic nutrients in soil is represented by $I$. A system of differential equations that describes the turnover of inorganic and organic nutrients is the following:

$$
\begin{split}
\frac{dN}{dt} &= L + kdI - \mu N I - \delta N  \\
\frac{dI}{dt} &= \mu N I - k d I - \delta I
\end{split}
$$


a. Verify that $\displaystyle N = \frac{L}{\delta}, \; I = 0$ and $\displaystyle N = \frac{kd+\delta}{\mu}, \; I = \frac{L \mu - \delta k d - \delta^{2}}{\mu \delta}$ are equilibrium solutions for this system.
b. Use `R` to compute each entry of the Jacobian matrix.
c. Construct a Jacobian matrix for each of the equilibrium solutions.


:::



<!-- Adapted from LW pg 121 -->

::: {#exr-lw-121-kermack-17}
(Inspired by @logan_mathematical_2009 \& @kermack_contribution_1927) A model for the spread of a disease where people recover is given by the following differential equation:

$$
\begin{split}
\frac{dS}{dt} &= -\alpha SI \\
\frac{dI}{dt} &= \alpha SI - \gamma I \\
\frac{dR}{dt} &= \gamma I
\end{split}
$$

Assume this population has $N=1000$ people.

a. Determine the equilibrium solutions for this system of equations.
b. Construct the Jacobian for each of the equilibrium solutions.
c. Let $\alpha=0.001$ and $\gamma = 0.2$. With the Jacobian matrix, generate the phase plane (using the equations for $\displaystyle \frac{dS}{dt}$ and $\displaystyle \frac{dI}{dt}$ only) for all of the equilibrium solutions and classify their stability.


:::




```{tikz,label=fig-glucose,warning=FALSE,message=FALSE,echo=FALSE,fig.cap="Glucose transporter reaction schemes.",fig.width=4,fig.height=3}


\tikzstyle{vspecies}=[rectangle,minimum size=0.5cm,draw=black,node distance = 3 cm]
\begin{tikzpicture}[auto, outer sep=1pt, node distance=1.5cm]

%% keener pg 65 - need to adapt somewhat with se
\node [vspecies] (A) {$c_{e}$} ;
\node [vspecies, right of = A] (B) {$p_{e}$} ;
\node [vspecies, below of = A] (C) {$c_{i}$} ;
\node [vspecies, below of = B] (D) {$p_{i}$} ;
\node [right of = D,node distance = 1.5cm] (model) {inside};
\node [right of = B,node distance = 1.5cm] (model2) {outside};

\draw [->] ([yshift=3pt]A.east) --  node[above] {\small{$k_+ S_{e}$}} ([yshift=3pt]B.west) ;
\draw [->] ([yshift=-3pt]B.west) --  node[below] {\small{$k_-$}} ([yshift=-3pt]A.east) ;

\draw [->] ([yshift=3pt]C.east) --  node[above] {\small{$k_+ S_{i}$}} ([yshift=3pt]D.west) ;
\draw [->] ([yshift=-3pt]D.west) --  node[below] {\small{$k_-$}} ([yshift=-3pt]C.east) ;

\draw [->] ([xshift=-3pt]A.south) --  node[left] {\small{$k$}} ([xshift=-3pt]C.north) ;
\draw [->] ([xshift=3pt]C.north) --  node[right] {\small{$k$}} ([xshift=3pt]A.south) ;

\draw [->] ([xshift=-3pt]B.south) --  node[left] {\small{$k$}} ([xshift=-3pt]D.north) ;
\draw [->] ([xshift=3pt]D.north) --  node[right] {\small{$k$}} ([xshift=3pt]B.south) ;



\end{tikzpicture}

```



<!-- Problem adapted from Keener pg 64 and 65 2.4.1 glucose transport -->

::: {#exr-keener-64-17}
(Inspired by @keener_mathematical_2009) The chemical glucose is transported across the cell membrane using carrier proteins. These proteins can have different states (open or closed) that can be bound to a glucose substrate. The schematic for this reaction is shown in @fig-glucose. The system of differential equations describing this reaction is:
  
$$
\begin{split}
\frac{dp_{i}}{dt} &= k p_{e} - k p_{i} + k_{+} s_{i}c_{i}-k_{i}p_{i} \\
\frac{dp_{e}}{dt} &= k p_{i} - k p_{e} + k_{+} s_{e}c_{e}-k_{-}p_{e} \\
\frac{dc_{i}}{dt} &= k c_{e} - k c_{i} + k_{-} p_{i}-k_{+}s_{i}c_{i} \\
\frac{dc_{e}}{dt} &= k c_{i} - k c_{e} + k_{-} p_{e}-k_{+}s_{e}c_{e}
\end{split}
$$


a. We can reduce this to a system of three equations. First show that $\displaystyle \frac{dp_{i}}{dt} + \frac{dp_{e}}{dt}+ \frac{dc_{i}}{dt} + \frac{dc_{e}}{dt} = 0$. Given that $p_{i}+p_{e}+c_{i}+c_{e}=C_{0}$, where $C_{0}$ is constant, use this equation to eliminate $p_{i}$ and write down a system of three equations.
b. Determine the equilibrium solutions for this new system of three equations.
c. Use `R` to compute each entry of the Jacobian matrix.
c. Construct the Jacobian matrix for each of these equilibrium solutions.

:::
