# Local Linearization and the Jacobian {#jacobian-17}

In the previous few sections we focused on systems of equations and the types of behaviors we should expect when analyzing the stability of the equilibrium solution.  In this section we will extend that concept to nonlinear systems of equations.  We will analyze stability of a nonlinear system through the concept of *local linearization*.

## A first example
Let's take a look at the following phase plane to this non-linear system:

\begin{equation}
\begin{split} 
\frac{dx}{dt} &= y-1 \\  (\#eq:ex1-ch17)
\frac{dy}{dt} &= x^{2}-1 
\end{split}
\end{equation}

The nullclines for this system are $y=1$ and $x=\pm1$, leading two two equilibrium solutions when we create the phaseplane:

```{r ex1-phaseplane,fig.width=4,fig.height=3,echo=FALSE,fig.cap="Phaseplane for Equation \\ref{eq:ex1-ch17}."  }
# Define the range we wish to evaluate this vector field


system_eq <- c(dx ~ y-1,
               dy ~ x^2-1)

phaseplane(system_eq,'x','y',x_window = c(-2,2),y_window = c(-2,2))+  # The values in quotes are the labels for the axes
stat_function(fun=function(x)1, geom="line", aes(colour="x' = 0"),size=1) +
  geom_vline(data=tibble(xint=c(-1,1)),aes(xintercept=xint,colour="y' = 0"),size=1) + geom_point(data=tibble(xint=c(-1,1),yint=c(1,1)),aes(x=xint,y=yint),color='red',size=4) +
  scale_colour_discrete(name='Nullclines')
  

```

From Figure \@ref(fig:ex1-phaseplane), arrows on the phaseplane by the equilibrium solution at $(x,y)=(-1,1)$ seems to spiral inwards, whereas the second equilibrium solution at $(x,y)=(1,1)$ has behavior that flows in and out depending on the direction you approach.  Beyond this visualization we can define additional tools to help characterize the stability of this system near the equilibrium solutions. 

This section is focused on understanding the behavior near the equilibrium solutions, and in particular applying concepts of _local linearization_ along with our understanding of linear systems.  Let's get started!


## The lynx hare revisited
Let's take a look at another familiar example.  Consider the following nonlinear system of equations from the Lynx-Hare model, with $H$ and $L$ measured in thousands of animals:

\begin{equation}
\begin{split}
\frac{dH}{dt} &= .3 H - .1 HL \\  (\#eq:lynx-hare-17)
\frac{dL}{dt} &=.05HL -.2L
\end{split}
\end{equation}


You can show that the steady states for Equation \@ref(eq:lynx-hare-17) are $(H,L)=(0,0)$ and $(4,3)$.  The phase plane diagram for this system is the following:

```{r,fig.width=4,fig.height=3}
# Define the range we wish to evaluate this vector field
H_window <- c(0,5)
L_window <- c(0,5)

system_eq <- c(dH ~ .3*H - .1*H*L,
               dL ~ .05*H*L -.2*L)

# Reminder: The values in quotes are the labels for the axes
phaseplane(system_eq,'H','L',x_window = H_window, y_window = L_window)  


```

Let's take a closer look at the phase plane near the first equilibrium solution:

```{r,label="lynx-hare-closer",fig.width=4,fig.height=3,fig.cap="A zoomed in view of the lynx-hare system."}
# Define the range we wish to evaluate this vector field
H_window <- c(0,0.5)
L_window <- c(0,0.5)

system_eq <- c(dH ~ .3*H - .1*H*L,
               dL ~ .05*H*L -.2*L)

# Reminder: The values in quotes are the labels for the axes
phaseplane(system_eq,'H','L',x_window = H_window, y_window = L_window) 


```

While the phaseplane in Figure \@ref(fig:lynx-hare-closer) _looks_ like this equilibrium solution is unstable, verifying this with another approach would be useful.  To do so, we are going to do a **locally linear approximation** or **tangent plane approximation** around $H=0$, $L=0$, which we discuss next.

## Tangent plane approximations
To extend the notion of a linear approximation, we apply the **tangent plane approximation**  at the point $x=a$, $y=b$:

\begin{equation}
L(x,y) = f(a,b) + f_{x}(a,b) \cdot (x-a) + f_{y}(a,b) \cdot (y-b),
\end{equation}

where $f_{x}$ is the partial derivative of $f(x,y)$ with respect to $x$ and $f_{y}$ is the partial derivative of $f(x,y)$ with respect to $y$.  For simplicity let's work with the equilibrium solution at $(0,0)$. If we say that $f(H,L)=.3 H - .1 HL$, we know $f(0,0)=0$.  Furthermore the locally linear approximation is:


\begin{equation}
\begin{split}
f_{H} = .3 - .1L & \rightarrow f_{H}(0,0)=.3 \\
f_{L} = -.1H & \rightarrow f_{L}(0,0)=0
\end{split}
\end{equation}

With this information we are able to take this and then write down the locally linear approximation for $f(H,L)$:
\begin{equation}
f(H,L) \approx 0 + .3 \cdot (H-0) - 0 \cdot (L-0) = .3H
\end{equation}
.05*H*L -.2*L
Likewise if we consider $g(H,L)= .05HL -.2L$, then we have:

\begin{equation}
\begin{split}
g_{H} = .05L-.2L & \rightarrow g_{H}(0,0)=0 \\
g_{L} = .05H-.2 & \rightarrow f_{L}(0,0)=-.2
\end{split}
\end{equation}


With these results, our locally linear approximation for $g(H,L)$ is:

\begin{equation}
g(H,L) \approx 0 + 0  \cdot (H-0) -.2 \cdot (L-0) = 0 - .2L.
\end{equation}

So, at the equilibrium solution, the system behaves like the following system of equations:

\begin{align}
\frac{dH}{dt} &= .3H\\
\frac{dL}{dt} &= - .2L
\end{align}


Notice how this the system is entirely decoupled linear system of equations, so we have for an initial condition $(H_{0},L_{0})$ that $H(t)=H_{0}e^{.3t}$ and $L(t)=L_{0}e^{-.2t}$.

## The Jacobian matrix
If we look at the last example before the solution we can write our system in matrix form:

\begin{equation}
\begin{pmatrix} H' \\ L' \end{pmatrix} =\begin{pmatrix} .3 & 0 \\ 0 &  -.2 \end{pmatrix} \begin{pmatrix} H \\ L \end{pmatrix}
\end{equation}

We define the matrix $\displaystyle J= \begin{pmatrix} .3 & 0 \\ 0 &  -.2 \end{pmatrix}$ as the Jacobian matrix.  This is a key matrix we will use to investigate stability of nonlinear systems.  While in the Lynx-Hare example we could easily solve the uncoupled system of equations, it may not be this case all the time. In that case we will need to understand a little bit more about analyzing them - which we will revisit the idea of straight line solutions and eigenvalues.

## Predator prey with logistic growth
Let's take a look at another model developed from the lynx-hare system.  We assumed that the hare grow exponentially (notice the term $rH$ in their equation.)  However we can modify their growth rate to be a logistic growth function with carrying capacity $K$:

\begin{align}
\frac{dH}{dt} &= r H \left( 1- \frac{H}{K} \right) - b HL \\
\frac{dL}{dt} &=ebHL -dL
\end{align}

Through a rescaling of this system $\displaystyle x=\frac{H}{K}$, $\displaystyle y=\frac{L}{r/b}$ and $T = r t$ we can rewrite this system as:

\begin{align}
\frac{dx}{d T} &= x(1-x) - xy \\
\frac{dy}{d T} &=\frac{ebK}{r}xy -\frac{d}{r}y
\end{align}

In order to analyze the Jacobian matrix we will need to compute several partial derivatives:

\begin{align}
\frac{\partial}{\partial x} \left( f(x,y) \right) &= \frac{\partial}{\partial x} \left( x(1-x) - xy \right) = 1-2x-y \\
\frac{\partial}{\partial y} \left( f(x,y) \right) &= \frac{\partial}{\partial y} \left( x(1-x) - xy \right) = -x \\
\frac{\partial}{\partial x} \left( g(x,y) \right) &= \frac{\partial}{\partial x} \left( \frac{ebK}{r}xy -\frac{d}{r}y \right) = \frac{ebK}{r}y  \\
\frac{\partial}{\partial y} \left( g(x,y) \right) &= \frac{\partial}{\partial y} \left( \frac{ebK}{r}xy -\frac{d}{r}y \right) = \frac{ebK}{r}x -\frac{d}{r}
\end{align}

So we can construct the Jacobian matrix, ready for evaluation at any $(x,y)$ equilibrium solution:

\begin{equation}
J_{(x,y)} = \begin{pmatrix} 1-2x-y & -x \\ \frac{ebK}{r}y & \frac{ebK}{r}x -\frac{d}{r} \end{pmatrix}
\end{equation}

Sometimes computing the Jacobian matrix is a good first step so then you are ready to compute the equilibrium solutions.  In the exercises you will determine equilibrium solutions and visualize the Jacobian matrix.

## Concluding thoughts
To summarize, let's say we have the following system of equations

\begin{align}
\frac{dx}{dt} &= f(x,y) \\
\frac{dy}{dt} &= g(x,y)
\end{align}

Assuming we have an equibrium solution at $(x,y)=(a,b)$, the Jacobian matrix at that solution is:

\begin{equation}
J_{(a,b)} =\begin{pmatrix} f_{x}(a,b) & f_{y}(a,b) \\ g_{x}(a,b) &  g_{y}(a,b) \end{pmatrix}
\end{equation}

The Jacobian matrix also extends to higher order systems as well.

\newpage

## Exercises
```{exercise}
Consider the following nonlinear system:

\begin{equation}
\begin{split} 
\frac{dx}{dt} &= y-1 \\ 
\frac{dy}{dt} &= x^{2}-1 
\end{split}
\end{equation}

\begin{enumerate}[label=\alph*.]
\item Verify that this system has equilibrium solutions at $(-1,1)$ and $(1,1)$.
\item Construct the Jacobian matrix at the equilibrium solutions at $(-1,1)$ and $(1,1)$.
\item With the Jacobian matrix, visualize a phaseplane at these equilbrium solutions to determine stability.
\end{enumerate}

```
&nbsp;

```{exercise}
By solving directly, show that $(H,L)=(0,0)$ and $(4,3)$ are equilibrium solutions to the following system of equations:

\begin{equation}
\begin{split}
\frac{dH}{dt} &= .3 H - .1 HL \\
\frac{dL}{dt} &=.05HL -.2L
\end{split}
\end{equation}
```
&nbsp;


```{exercise}
Consider the following nonlinear system:

\begin{equation}
\begin{split}
\frac{dx}{dt} &= x-y \\
\frac{dy}{dt} &=-y + \frac{5x^2}{4+x^{2}},
\end{split}
\end{equation}

\begin{enumerate}[label=\alph*.]
\item Verify that the point $(x,y)=(1,1)$ is an equilibrium solution.
\item Construct the Jacobian matrix at this equilibrium solution.
\item With the Jacobian matrix, visualize a phaseplane at that equilbrium solution to determine stability.
\end{enumerate}

```
&nbsp;
```{exercise}
Consider the nonlinear system

\begin{equation}
\begin{split}
\frac{dx}{dt} &= 2x+3y+xy \\
\frac{dy}{dt} &= -x + y - 2xy^{3},
\end{split}
\end{equation}

\begin{enumerate}[label=\alph*.]
\item Verify that the point (0,0) is an equilibrium solution.
\item Determine the Jacobian matrix at this equilbrium solution.
\item With the Jacobian matrix, visualize a phaseplane at that equilbrium solution to determine stability.
\end{enumerate}
```
&nbsp;

```{exercise}
Consider the following system:

\begin{equation}
\begin{split}
\frac{dx}{dt} &= y^{2} \\
\frac{dy}{dt} &= -\frac{2}{3} x,
\end{split}
\end{equation}

\begin{enumerate}[label=\alph*.]
\item Determine the equilibrium solution.
\item Construct the Jacobian at the equilibrium solution.
\item With the Jacobian matrix, visualize a phaseplane at that equilbrium solution to determine stability.
\item Use the fact that that $\displaystyle \frac{dy}{dx} / \frac{dy}{dt} = \frac{dy}{dx}$, which should yield a separable differential equation that will allow you to solve $y(x)$.  How does that solved differential equation compare to what you found by analyzing the Jacobian matrix?
\end{enumerate}
```

&nbsp;


```{exercise}
Consider the lynx-hare system with parameters $r$, $b$, $e$, and $d$:
  
\begin{equation}
\begin{split}
\frac{dH}{dt} &= r H - b HL \\
\frac{dL}{dt} &=ebHL -dL
\end{split}
\end{equation}

\begin{enumerate}[label=\alph*.]
\item Verify that the point $\displaystyle \left( \frac{d}{eb}, \frac{r}{b} \right)$ is an equilibrium solution.
\item Construct the Jacobian at the equilibrium solution. (Your final answer will include the parameters.)
\end{enumerate}
```

&nbsp;

```{exercise}
This problem revisits the modified lynx-hare system:
\begin{equation}
\begin{split}
\frac{dH}{dt} &= r H \left( 1- \frac{H}{K} \right) - b HL \\
\frac{dL}{dt} &=ebHL -dL
\end{split}
\end{equation}

\begin{enumerate}[label=\alph*.]
\item Verify that by rescaling of this system $\displaystyle x=\frac{H}{K}$, $\displaystyle y=\frac{L}{r/b}$ and $T = r t$ we can rewrite this system as:

\begin{equation}
\begin{split}
\frac{dx}{d T} &= x(1-x) - xy \\
\frac{dy}{d T} &=\frac{ebK}{r}xy -\frac{d}{r}y
\end{split}
\end{equation}

\item Determine the three equilibrium solutions of this rescaled system.
\item Evaluate the Jacobian at each of the equilibrium solutions.
\item With your Jacobian, select reasonable values of the parameters to generate a phaseplane diagram and classify the stability of each equilibrium solution.
\item How does this revised system compare to the stability of the regular lynx-hare model?
  \end{enumerate}
```

&nbsp;



<!-- Adapted from LW pg 121 -->
```{exercise}
A model for the spread of a disease where people recover is given by the following differential equation:

\begin{equation}
\begin{split}
\frac{dS}{dt} &= -\alpha SI \\
\frac{dI}{dt} &= \alpha SI - \gamma I \\
\frac{dR}{dt} &= \gamma I,
\end{split}
\end{equation}

\begin{enumerate}[label=\alph*.]
\item Determine the equilibrium solutions for this system of equations.
\item Construct the Jacobian for each of the equilibrium solutions.
\item Let $\alpha=0.001$ and $\gamma = 0.2$. With the Jacobian matrix, generate the phase plane (using the equations for $\displaystyle \frac{dS}{dt}$ and $\displaystyle \frac{dI}{dt}$ only) for all of the equilibrium solutions and classify their stability.
\end{enumerate}

```

&nbsp;

<!-- LW pg 157 158 #3 -->
```{exercise}
A population of fish used as food is an example of a renewable resource. This population decreases due to harvesting the fish.  As long as the rate of harvest is smaller than the replacement rate, theoretically the population would be considered a renewable resource.  However fish do have natural predators, which may also be harvested at the same time.  A model that describes this interaction is

\begin{equation}
\begin{split}
\frac{dN}{dt} &= rN - cNP - \rho E N \\
\frac{dP}{dt} &= bcNP - mP - \sigma EP,
\end{split}
\end{equation}

where $E$ is the fishing effort and $\rho$ and $\sigma$ are the catchability coefficients for the prey and predator.

\begin{enumerate}[label=\alph*.]
\item Determine all the equilibrium solutions for this model.
\item Construct the Jacobian matrix for each of the the equilibrium solutions.
\item Select different parameter values.  With the Jacobian, construct a phase plane diagram for each of the equilibrium solutions.
\item In the absence of fishing $E=0$.  How do these equilibrium solutions compare to the ones from the previous part?
\end{enumerate}
```
&nbsp;

```{r glucose,engine='tikz',warning=FALSE,message=FALSE,echo=FALSE,fig.cap="Glucose transporter reaction schemes."}

\begin{center}

\tikzstyle{vspecies}=[rectangle,minimum size=0.5cm,draw=black,node distance = 3 cm]
\begin{tikzpicture}[auto, outer sep=1pt, node distance=1.5cm]

%% keener pg 65 - need to adapt somewhat with se
\node [vspecies] (A) {$C_{e}$} ;
\node [vspecies, right of = A] (B) {$P_{e}$} ;
\node [vspecies, below of = A] (C) {$C_{i}$} ;
\node [vspecies, below of = B] (D) {$P_{i}$} ;
\node [right of = D,node distance = 1.5cm] (model) {inside};
\node [right of = B,node distance = 1.5cm] (model2) {outside};

\draw [->] ([yshift=3pt]A.east) --  node[above] {\small{$k_+ S_{e}$}} ([yshift=3pt]B.west) ;
\draw [->] ([yshift=-3pt]B.west) --  node[below] {\small{$k_-$}} ([yshift=-3pt]A.east) ;

\draw [->] ([yshift=3pt]C.east) --  node[above] {\small{$k_+ S_{i}$}} ([yshift=3pt]D.west) ;
\draw [->] ([yshift=-3pt]D.west) --  node[below] {\small{$k_-$}} ([yshift=-3pt]C.east) ;

\draw [->] ([xshift=-3pt]A.south) --  node[left] {\small{$k$}} ([xshift=-3pt]C.north) ;
\draw [->] ([xshift=3pt]C.north) --  node[right] {\small{$k$}} ([xshift=3pt]A.south) ;

\draw [->] ([xshift=-3pt]B.south) --  node[left] {\small{$k$}} ([xshift=-3pt]D.north) ;
\draw [->] ([xshift=3pt]D.north) --  node[right] {\small{$k$}} ([xshift=3pt]B.south) ;



\end{tikzpicture}
\end{center}

```

&nbsp;

<!-- Problem adapted from Keener pg 64 and 65 2.4.1 glucose transport -->
```{exercise}
The chemical glucose is transported across the cell membrane using a carrier proteins.  These proteins can have different states (open or closed) that can be bound to a glucose substrate.  The schematic for this reaction is shown in Figure \@ref(fig:glucose).  The system of differential equations describing this reaction is:
  
\begin{equation}
\begin{split}
\frac{dp_{i}}{dt} &= k p_{e} - k p_{i} + k_{+} s_{i}c_{i}-k_{i}p_{i} \\
\frac{dp_{e}}{dt} &= k p_{i} - k p_{e} + k_{+} s_{e}c_{e}-k_{-}p_{e} \\
\frac{dc_{i}}{dt} &= k c_{e} - k c_{i} + k_{-} p_{i}-k_{+}s_{i}c_{i} \\
\frac{dc_{e}}{dt} &= k c_{i} - k c_{e} + k_{-} p_{e}-k_{+}s_{e}c_{e}
\end{split}
\end{equation}

\begin{enumerate}[label=\alph*.]
\item We can reduce this to a system of three equations.  First show that $\displaystyle \frac{dp_{i}}{dt} + \frac{dp_{e}}{dt}+ \frac{dc_{i}}{dt} + \frac{dc_{e}}{dt} = 0$.  Given that $p_{i}+p_{e}+c_{i}+c_{e}=C_{0}$, where $C_{0}$ is constant, use this equation to eliminate $p_{i}$ and write down a system of three equations.
\item Determine the equilibrium solutions for this new system of three equations.
\item Construct the Jacobian matrix for each of these equilibrium solutions.
\end{enumerate}
```
