# Qualitative Stability Analysis


So far we have seen how we can use local linearization near an equilibrium solution to classify its stability.  We analyze stability by examining the eigenvalues. Computing eigenvalues can be tricky - but for a two-dimensional system some interesting results occur.It turns out that we can analyze stability without solving a quadratic equation, but first we will need to understand some key relationships.


## The general case
Consider the following two dimensional linear system, where $a$, $b$, $c$, and $d$ can be any number:

\begin{equation}
\begin{pmatrix} \frac{dx}{dt} \\ \frac{dy}{dt} \end{pmatrix} =  \begin{pmatrix} ax+by \\ cx+dy \end{pmatrix} =  \begin{pmatrix} a & b \\ c &  d \end{pmatrix} \begin{pmatrix} x \\ y \end{pmatrix}
\end{equation}

Eigenvalues are found by solving $\displaystyle \det (A - \lambda I ) =0$.  So we have the following:

\begin{equation}
\det \begin{pmatrix} a - \lambda & b \\ c & d-\lambda \end{pmatrix} = (a-\lambda)(d-\lambda) - bc
\end{equation}

If we factor out this expression we obtain what is called the characteristic equation:

\begin{equation}
\lambda^{2} - (a+d) \lambda + ad - bc = 0
\end{equation}

What is cool about this equation is that the roots can be expressed as functions of the entries of the matrix $A$.  In fact, the term $a+d$ is the sum of the diagonal entries, which is known as the **trace** of a matrix, or $\mbox{tr}(A)$.  And you may recognize that $ad-bc$ is $\det(A)$.  So our characteristic equation is $\lambda^{2} - \mbox{tr}(A)\lambda + \det(A)$.

```{example}
Write down the characteristic equation for the system $\vec{x}'=Ax$ where $\displaystyle A=  \begin{pmatrix} -1 & 1 \\ 0 &  3 \end{pmatrix}$ and classify the stability of the equilibrium solution.
```
&nbsp;
```{remark}
We can see that $\det(A)= -1(3) - 0(1) = -3$ and tr$(A)=2$, so our characteristic equation is $\lambda^{2}-2\lambda -3$.  If we solve $\lambda^{2}-2\lambda -3=0$ we have $(\lambda-3)(\lambda+1)=0$, so our eigenvalues are $\lambda=3$ and $\lambda=-1$, so the equilibrium solution is a saddle node.
```

This might be a computationally easier way to compute the eigenvalues, especially if you have a known parameter. We can also exploit this relationship even more, which may require some circularity in the argument.


Let's say we have two eigenvalues $\lambda_{1}$ and $\lambda_{2}$.  We make no assumptions on if they are real or imaginary or equal.  But if they are eigenvalues, then they are roots of the characteristic polynomial.  This means that $(\lambda-\lambda_{1})(\lambda-\lambda_{2})=0$.  If we factor out this equation we have $\lambda^{2}-(\lambda_{1}+\lambda_{2}) \lambda + \lambda_{1} \lambda_{2}=0$.  Hmmm.  If we compare our two equations we have:

\begin{equation}
\lambda^{2}-(\lambda_{1}+\lambda_{2}) \lambda + \lambda_{1} \lambda_{2} = \lambda^{2} - \mbox{tr}(A)\lambda + \det(A)
\end{equation}

This uncovers some neat relationships - in particular tr$(A)=(\lambda_{1}+\lambda_{2})$ and $\det(A)=\lambda_{1}+\lambda_{2}$.  Why should we bother with this?  Well this provides an alternative pathway to understand stability through the trace and determinant, in particular we have the following correspondence between the signs of the eigenvalues and the trace and determinant:

Sign of $\lambda_{1}$ | Sign of $\lambda_{2}$  | Tendency of solution | Sign of tr$(A)$ | Sign of $\det(A)$
-------------| ------------- | ------------- | -------------| ------------- |
    Positive | Positive | Source | Positive | Positive
    Negative |  Negative | Sink | Negative | Positive
   Positive  | Negative | Saddle | ? | Negative
    Negative |  Positive | Saddle | ? | Negative
   
For the moment we will only consider real non-zero values of the eigenvalues - more specialized cases will occur later.  But carefully at the table:

- If the determinant is *negative*, then the equilibrium solution is a *saddle*.
- If the determinant is *positive* and the trace is *negative*, then the equilibrium solution is a *sink*
- If the determinant and trace are both *positive*, then the equilibrium solution is a *source*.

This is a pretty quick and easy way to investigate stability!  We can also look back to the first example - notice how the determinant was negative, confirming that is a saddle.

So far we can represent the stability of solutions in what is called the trace-determinant plane, with the trace on the horizontal axis:

```{r,engine='tikz',warning=FALSE,message=FALSE,echo=FALSE}
\begin{center}
\begin{tikzpicture}
%\draw[help lines, color=gray!30, dashed] (-4.9,-4.9) grid (4.9,4.9);
\draw[<->,ultra thick] (-5,0)--(5,0) node[right]{tr$(A)$};
\draw[<->,ultra thick] (0,-1)--(0,3) node[above]{det$(A)$};
%\draw[thick,scale=0.5,domain=-5:5,smooth,variable=\x,blue] plot ({\x},{0.3*\x*\x});
\node at (0,-1.5) {Saddle node};
\node at (-3,1.5) {Sink};
\node at (3,1.5) {Source};
\end{tikzpicture}
\end{center}
```


It also turns out that we can extend this diagram to the imaginary and nonzero cases for the roots of the characteristic equation.  Using the quadratic formula we can solve directly for the eigenvalues as a function of the trace and determinant:

\begin{equation}\label{eq:roots}
\lambda_{1,2}= \frac{\mbox{tr}(A)}{2} \pm \frac{\sqrt{ (\mbox{tr}(A))^2-4 \det(A)}}{2}
\end{equation}

This seems like a more complicated expression, but it can be shown to be consistent with our above work.  Moving to our quadratic equation allows us to see what happens when we have imaginary eigenvalues - which can be a stable or unstable spiral depending on its location in the trace-determinant plane.

```{r,engine='tikz',warning=FALSE,message=FALSE,echo=FALSE}
\begin{center}
\begin{tikzpicture}
%\draw[help lines, color=gray!30, dashed] (-4.9,-4.9) grid (4.9,4.9);
\draw[<->,ultra thick] (-5,0)--(5,0) node[right]{tr$(A)$};
\draw[<->,ultra thick] (0,-1)--(0,3) node[above]{det$(A)$};
\draw[thick,scale=0.5,domain=-5:5,smooth,variable=\x,blue] plot ({\x},{0.3*\x*\x});
\node at (0,-1.5) {Saddle node};
\node at (-3,1.5) {Sink};
\node at (3,1.5) {Source};
\node[fill=white] at (0,1.5) {Center};
\node[align=center] at (1,2.3) {Spiral \\ source};
\node[align=center] at (-1,2.3) {Spiral \\ sink};
\end{tikzpicture}
\end{center}
```

If the discriminant (the part inside the square root, which is $(\mbox{tr}(A))^2-4 \det(A)$) of the eigenvalue expression is negative then we have a spiral source or spiral sink depending on the positivity of the tr$(A)$.
Finally the *center* equilibrium occurs when the trace is exactly zero and the determinant is positive.  This graphic of the trace-determinant plane is a quick way to analyze stability of a solution without a lot of algebraic analysis.


## Sensitivity to parameters with the trace-determinant
Where the trace determinant plane is even more useful is with a matrix of parameter values.  Consider the following lynx - hare system:

\begin{align}
\frac{dH}{dt} &= r H - b HL \\
\frac{dL}{dt} &=ebHL -dL
\end{align}

We have studied this system several times.  We know that the steady states of this system are $(H,L)=(0,0)$ and $\displaystyle \left( \frac{d}{eb}, \frac{r}{b} \right)$ in this model.  Let's compute the Jacobian at each of these systems:

\begin{align}
J_{(0,0)} &=\begin{pmatrix} r & 0 \\ 0 & -d \end{pmatrix} \\
J_{(\frac{d}{eb}, \frac{r}{b})} &=\begin{pmatrix} r-\frac{d}{e} & -\frac{d}{e} \\ er & 0 \end{pmatrix}
\end{align}

Let's analyze the Jacobian at the origin.  We have tr$(J)=r-d$ and $\det(J)=-rd$.  Since $r$ and $d$ are both positive parameters the determinant will always be negative, so no matter what the origin will be a saddle.

For the second equilibrium solution we have tr$\displaystyle (J)=r-\frac{d}{e}$ and $\det(J)=d$.  The determinant is positive, however we can see that the stability of this equilibrium solution will be dependent on the trace.  If the trace is positive, or $\displaystyle r > \frac{d}{e}$ than the equilibrium solution will be unstable.  If $\displaystyle r < \frac{d}{e}$ it will be stable.  You can derive even stronger boundaries between a spiral source or spiral sink as well.

## Higher dimensional stability
The trace-determinant plane is a really use approach to analyze stability without computing the eigenvalues directly.  It is also useful when you have a parameter in your Jacobian - the trace-determinant will allow you to characterize the stability of solutions as a function of a parameter.  This leads to what is known as a bifurcation diagram, which we will study in the next section.

The conditions between the trace and determinant really describe the notion of eigenvalues as a function of the entries of a matrix.  It may be natural to ask if similar conditions exist for larger dimensioned matrices.  It turns out yes ... to a point.  The [Routh-Hurwitz stability criterion](https://en.wikipedia.org/wiki/Routh%E2%80%93Hurwitz_stability_criterion) is one such approach, but it gets tricky for larger matrices.



\newpage
## Exercises

<!-- I think this is from LW -->
```{exercise}
Compute the trace and determinant for each of these linear systems.  Use the trace-determinant condition to classify the stability of the equilibrium solutions. Verify your results are consistent with the eigenvalue calculations.
\begin{enumerate}
\item $\displaystyle \frac{dx}{dt} = -x, \;  \frac{dy}{dt} = -2y$
\item $\displaystyle \frac{dx}{dt} = 3x+y, \;  \frac{dy}{dt} = 2x+4y$
\item $\displaystyle \frac{dx}{dt} = 8x-11y, \;  \frac{dy}{dt} = 6x-9y$
\item $\displaystyle \frac{dx}{dt}= 3x-y, \; \frac{dy}{dt}=3y$
\item $\displaystyle \frac{dx}{dt} = -2x-3y, \;  \frac{dy}{dt} = 3x-2y$
\end{enumerate}
```
&nbsp;
```{exercise}
For the lynx-hare system equilibrium $\displaystyle \left( \frac{d}{eb}, \frac{r}{b} \right)$, determine conditions on the parameters where the system will be a spiral source and sprial sink.
```

&nbsp;
```{exercise}
Algebraically show $\displaystyle \det(A-\lambda I) = \lambda^{2} - (a+d) \lambda + ad - bc$.
```
&nbsp;
```{exercise}
Apply the quadratic formula to $\lambda^{2} - \mbox{tr}(A)\lambda + \det(A)=0$ obtain Equation \ref{eq:roots}
```

&nbsp;

<!-- (x-(a+bi))(x-(a-bi)) = x^2 +(a+bi)(a-bi) -x(a+bi)-x(a-bi) = x^2 -2a x + (a^2+b^2)  -->
```{exercise}
Algebraically show that for two complex roots $\lambda_{1,2} = a \pm bi$ with $a$ and $b$ not equal to zero that tr$(A)=2a$ and $\det(A)=a^{2}+b^{2}$.
```

&nbsp;
<!-- LW pg 164 -->
```{exercise}
Consider the linear system of equations:
    \begin{align*}
\frac{dx}{dt}&=-ax-y \\
\frac{dy}{dt} &= -x-ay
\end{align*}

Apply the relationships between the trace and determinant to classify the stability of the equilibrium solution.

```
&nbsp;
```{exercise}
Explain why if the trace of a matrix equals zero and the determinant is positive the equilibrium solution is a center equilibrium.
```
&nbsp;
<!-- LW pg 183 -->
```{exercise}
The following equation can be applied to study cell differentiation:
  \begin{align*}
\frac{dx}{dt}&=y-x \\
\frac{dy}{dt} &= -y + \frac{5x^{2}}{4+x^{2}}
\end{align*}

\begin{enumerate}
\item Determine the equilibrium solutions.
\item Evaluate the Jacobian at each of the equilibrium solutions.
\item Evaluate the stability of each equilibrium solution using relationships between the trace and determinant.
\item Draw some sample solution curves using the command \texttt{systems}.
\end{enumerate}
```
&nbsp;
<!-- %LW #12 pg 183 -->
```{exercise}
 Consider the following predator-prey model:
\begin{align*}
x' &= \frac{2}{3} x \cdot \left(1- \frac{x}{4} \right) - \frac{xy}{1+x} \\
y' &= ry \cdot \left(1 - \frac{y}{x} \right),
\end{align*}
Notice how the carrying capacity of the predator depends on the prey population.
\begin{enumerate}
\item Determine the equilibrium solutions for this differential equation.
\item Evaluate the stability of the equilibrium solutions. \emph{To analyze stability you may also use the relationships between the eigenvalues and the trace and determinant.}
\end{enumerate}
```
&nbsp;
<!-- %LW #12 pg 183 -->

```{exercise}
All of the following systems have an equilibrium solution at the origin (0,0).  Compute the Jacobian of these solutions and apply the trace and determinant conditions to analyze the stability.

\begin{enumerate}
\item \begin{align*}
x' &=x+ \mu y \\
y' &= \mu x -y,
\end{align*}
\item \begin{align*}
x' &=x+y\\
y' &= \mu x +y,
\end{align*}
\item \begin{align*}
x' &= y \\
y' &=x^{2}-x+\mu y,
\end{align*}
\end{enumerate}
```
&nbsp;
<!-- adapted LW pg 183 -->
```{exercise}
Let $C$ be the amount of carbon in a forest ecosystem, with $P$ be the rate of increase due to photosynthesis.  Herbivores $H$ consume carbon on the following predator-prey model:
    \begin{align*}
\frac{dC}{dt}&=P - aC - bHC \\
\frac{dH}{dt} &= ebHC-dC
\end{align*}

\begin{enumerate}
\item Explain the various terms in this model and their biological meaning.
\item Determine the equilibrium solutions.
\item Evaluate the Jacobian at each of the equilibrium solutions.
\item Evaluate the stability of each equilibrium solution using relationships between the trace and determinant.
\end{enumerate}
```
