# Stochastic Differential Equations

In this section we will begin to combine our knowledge of random walks and apply them to stochastic differential equations.  Hopefully a lot of the groundwork has been laid to understand the structure of a stochastic differential equation.  Finally we will examine a numerical method based on Euler's method that allows us to simulate stochastic differential equations.


## The stochastic logistic model

In the last section we started to write down the format of a stochastic differential equation, which we will use the logistic equation for context:

\begin{equation}
dx = rx \left(1 - \frac{x}{K} \right) \; dt + \mbox{ Noise } \; dt
\end{equation}

The first term ($\displaystyle rx \left(1 - \frac{x}{K} \right) \; dt$) is called the "deterministic part" of the equation.  The second term ($\mbox{ Noise } \; dt$) is the "stochastic part".  We are going to model that with the white noise (Weiner process) $dW(t)=\mbox{ Noise } \; dt$, allowing us to write this more formally as:

\begin{equation}
dx = rx \left(1 - \frac{x}{K} \right) \; dt + dW(t)
\end{equation}


Here the term $dW(t)$ is short hand to the differential equation $\displaystyle \frac{dW}{dt} = \mbox{ Noise }$, where $W(t)$ is the solution to a Weiner process (random walk white noise) as studied previously. A good way to think of this stochastic differential equation is that a small change in the variable $x$ (represented by the term $dx$) is computed in two ways:

\begin{align*}
\mbox{Deterministic part: } & rx \left(1 - \frac{x}{K} \right) \; dt \\
\mbox{Stochastic part: } & dW(t)
\end{align*}

We know that the deterministic part of this equation has a stable steady state at $x=K$.  However how does the stochastic part of this differential equation change the solution trajectory?  Rather than focus on an exact solution technique we are going to focus on how to apply a numerical method to simulate solution trajectories and then take the ensemble average each of the time points.


## The Euler-Maruyama Method

The Euler-Maruyama method is a variation of Euler's method that accounts for stochasticity and implements the random walk.  We are going to build this up step by step.

First we consider $dx = x_{n+1}-x_{n}$, where $n$ is the current step of Euler's method.  Likewise $dW(t) = W_{n+1} - W_{n}$.  This terms represents one step of the random walk.  Given what we know about random walks we will approximate $W_{n+1}-W_{n} = \sqrt{\Delta t} Z_{1}$, where $Z_{1}$ is a random number from the standard unit normal distribution and $\Delta t$ is the timestep length.  We do this in `R` as `rnorm(1)`.  When we put these together we have the following iterative method:

- Given $\Delta t$ and starting value $x_{0}$.
- Then compute the next step: $\displaystyle x_{1} = x_{0} + rx_{0} \left(1 - \frac{x_{0}}{K} \right) \; \Delta t + \sqrt{\Delta t} Z_{1}$, where $Z_{1}$ is a random number from the standard unit normal distribution. (`rnorm(1)` in `R`)
- Repeat to step $n$: $\displaystyle x_{1} = x_{0} + rx_{0} \left(1 - \frac{x_{0}}{K} \right) \; \Delta t + \sqrt{\Delta t} Z_{1}$

That is it!  We can run this through as many steps as we want. I have some defined code that will apply the simulation, but just like `euler` or `systems` there are some things that need to be set first. In order to apply the Euler-Maruyama method you will need to define six things:
        
- The size ($\Delta t$) of your timesteps.
- The number of timesteps you wish to run Euler's method.  More timesteps means more computational time.  If $N$ is the number of timesteps, $\Delta t \cdot N$ is the total time.
- A function that we have for our deterministic dynamics.  For our example this equals $\displaystyle rx \left(1 - \frac{x}{K} \right)$.
- A function that we have for our stochastic dynamics for our example this equals 1.
- The values of the vector of parameters $\vec{\alpha}$.  For the logistic differential equation we will take $r=0.8$ and $K=100$.
- The standard deviation ($\sigma$) of our normal distribution and random walk.  Typically this is set to 1, but can be varied if needed.
- The number of simulations for our stochastic differential equation.
      
Sample code for this stochastic differential equation is shown below.  What is produced are two plots: a spaghetti plot showing the trajectory of each simulation, and an ensemble average across all the realizations.
      
### Set up your time values

First you need to set up the length of each timestep ($\delta T$) and the number of timesteps we compute the solution.  You do this by specifiying the variables `deltaT` and `timeSteps`:
```{r, fig.show='hold'}
  deltaT <- .05    # timestep length
  timeSteps <- 200   # must be a number greater than 1
```
      
      
### Set up your parameters
Next you should specify the values of your parameters $\vec{alpha}$.  The good news is that you can name the parameters and their values, and just refer to the shorthand name (not the numerical value) in subsequent code.
      
```{r, fig.show='hold'}
    parameters <- c(r=0.8, K=100)   # parameters: a named vector
```
      
### Set up your initial conditions.
Now we need to specify the starting values (initial conditions) for our differential equation.  *Note:* Where this method differs from `euler` or `systems` is that we only provide *one* set of initial conditions.
      
```{r, fig.show='hold'}
      
initialCondition <- c(x=3)  # Be sure you have enough conditions as you do variables.
      
```
      
### Set up your simulation parameters.
Now we need to specify the stochastic parameters - the number of simulations and the standard deviation of the random walk:
      
```{r, fig.show='hold'}
      
  nSimulations <- 500
  sigma <- 1
      
```

If these variables are not specified, the default is 1 for both.        
      
### Set up your dynamics
Next we need to code the deterministic and stochastic dynamics:
```{r, fig.show='hold'}
      # R function to calculate the value of the derivatives at each time value
      # Use the names of the variables as defined in the vectors above
      
  deterministicDynamics <- function(t, state, parameters){
      with(as.list(c(state, parameters)), {  # Do not edit this line
      dx = r*x*(1-x/K)  # <-- You may edit this line
      return(list(c(dx)))  # <-- If you have more equations you will need to list the dVariables
    })
  }
      
  stochasticDynamics <- function(t, state, parameters){
      with(as.list(c(state, parameters)), {  # Do not edit this line
      dx = 1  # <-- You may edit this line
      return(list(c(dx)))  # <-- If you have more equations you will need to list the dVariables
      })
    }
     
```

### Putting it all together
      
Now we have everything we need!  Basically the next step is to run the command `eulerStochastics`, which will print to the screen a plot of your equation. If you have entered everything above correctly, you may copy the following code to your working file.
      

      
```{r,fig.width=4,fig.height=3}
eulerStochastic(deltaT,timeSteps,initialCondition,deterministicDynamics,
                stochasticDynamics,parameters,nSimulations,sigma)
```
      
Notice the two outputs.  The first is the spaghetti plot of the outputs of all the simulations of the stochastic process.  The second plot is the ensemble average of the results.  Overall adding in this stochastic noise modified how quickly the population rose, but the average result follows the standard solution to the logistic curve.

## Adding stochasticity to parameters

Now that we have seen an example in adding stochasticity to the logistic equation, we can also parameters of the differential equation to be stochastic.  For example, let's say that the growth rate $r$ in the logistic equation is subject to stochastic effects.  How we would implement this by replacing $r$ with $r + \mbox{ Noise }$:

\begin{equation}
dx = (r + \mbox{ Noise } )x \left(1 - \frac{x}{K} \right) \; dt,
\end{equation}

Now what we do is separate out the terms that are multiplied by "Noise" - they will form the stochastic part of the differential equation.  The terms that aren't multipled by "Noise" form the deterministic part of the differential equation:

\begin{equation}
dx = r  x \left(1 - \frac{x}{K} \right) \; dt + x \left(1 - \frac{x}{K} \right) \mbox{ Noise } \; dt,
\end{equation}

So we have the following, writing $\mbox{ Noise } \; dt = dW(t)$:
\begin{align*}
\mbox{Deterministic part: } & rx \left(1 - \frac{x}{K} \right) \; dt \\
\mbox{Stochastic part: } & x \left(1 - \frac{x}{K} \right) dW(t)
\end{align*}

There are a few things to notice here - the deterministic part of the differential equation is what we would expect without noise added.  Also notice how the stochastic part of the differential equation changed.  Let's take a look at the simulations with the Euler-Maruyama method:

```{r,fig.width=4,fig.height=3}
  deltaT <- .05    # timestep length
  timeSteps <- 200   # must be a number greater than 1
  parameters <- c(r=0.8, K=100)   # parameters: a named vector
  initialCondition <- c(x=3)  # Be sure you have enough conditions as you do variables.
  nSimulations <- 500
  sigma <- 1

  deterministicDynamics <- function(t, state, parameters){
      with(as.list(c(state, parameters)), {  # Do not edit this line
      dx = r*x*(1-x/K)  # <-- You may edit this line
      return(list(c(dx)))  # <-- If you have more equations you will need to list the dVariables
    })
  }
      
  stochasticDynamics <- function(t, state, parameters){
      with(as.list(c(state, parameters)), {  # Do not edit this line
      dx = x*(1-x/K)  # <-- You may edit this line
      return(list(c(dx)))  # <-- If you have more equations you will need to list the dVariables
      })
    }

eulerStochastic(deltaT,timeSteps,initialCondition,deterministicDynamics,
                stochasticDynamics,parameters,nSimulations,sigma)
  
```
Notice how the dynamics are different - there is a lot more variability in how quickly the solution rises to the steady state of $K=100$.

## Concluding thoughts
If you start with a known differential equation and want to add stochasticity the process to identify the stochastic and deterministic parts are the following:

- Replace whatever parameter with a + Noise term.
- Collect terms multiplied by Noise - they will farm the stochastic part of the differential equation.
- The deterministic part of the differential equation should be your original differential equation.

The most general form of the stochastic differential equation is: $\displaystyle d\vec{y} = f(\vec{y},\vec{\alpha},t) \; dt + g(\vec{y},\vec{\alpha},t) \; dW(t)$, where $\vec{y}$ is the vector of state variables you want to solve for, and $\vec{\alpha}$ is your vector of parameters, and $dW(t)$ is the stochastic noise from the random walk.
      
At a given initial condition, the Euler-Maruyama method applies locally linear approximations to forecast the solution forward $\Delta t$ time units: $\displaystyle \vec{y}_{n+1} = y_{n} + f(\vec{y}_{n},\vec{\alpha},t_{n}) \cdot \Delta t + g(\vec{y}_{n},\vec{\alpha},t_{n}) \cdot \sigma \cdot \mbox{rnorm(N)} \cdot \sqrt{\Delta t}$, where rnorm(N) is $N$ dimensional random variable from a normal distribution with mean 0.



\newpage

## Exercises

<!-- LW pg 341 #2 -->
```{exercise}
Consider the logistic differential equation: $\displaystyle \frac{dx}{dt} = r x \left( 1-\frac{x}{K} \right)$.  Assume there is stochasticity in the inverse carrying capacity $1/K$ (so this means you will consider $1/K + \mbox{ Noise }$).  Identify the deterministic and stochastic part of each of the differential equation.  Next assume that $x(0)=3$,  $r=0.8$, $K=100$, $\Delta t = 0.05$, and $\sigma=1$, with 200 timesteps. Do 500 simulations of this stochastic process. Contrast your results to when we added stochasticity to the parameter $r$.
```

&nbsp;


```{r sis-stochastics,engine='tikz',warning=FALSE,message=FALSE,echo=FALSE,fig.cap="The $SIS$ model"}

\begin{center}
\tikzstyle{vspecies}=[rectangle,minimum size=0.5cm,draw=black]
\begin{tikzpicture}[auto, outer sep=1pt, node distance=2cm]

\node [vspecies] (S) {$S$} ;
\node [vspecies, right of = S] (I) {$I$} ;
\draw [->] ([yshift=3pt]S.east) --  node[above] {\small{$b$}} ([yshift=3pt]I.west) ;
\draw [<-] ([yshift=-3pt]S.east) --  node[below] {\small{$r$}} ([yshift=-3pt]I.west) ;
\end{tikzpicture}
\end{center}


```


<!-- Adapted from LW pg 346 #4 -->
```{exercise}
An $SIS$ model is one where susceptibles $S$ become infected $I$, and then after recovering from an illness, become susceptible again.  The schematic representing this is shown in Figure \@ref(fig:sis-stochastics).  While you can write this as a system of differential equations, assuming the population size is constant $N$ we have the following differential equation:
  
  \begin{equation}
\frac{dI}{dt} = b(N-I) I - r I
\end{equation}

\begin{enumerate}
\item Determine the equilibrium solutions for this model and analyze the stability of the equilibrium solutions.
\item Assuming $N=1000$, $r=0.01$, and $b=0.005$, $I(0)=1$, simulate this differential equation over two weeks with $\Delta t = 0.1$.  Show the plot of your result.
\item Assume the transmission rate $b$ is stochastic.  Write down this stochastic differential equation.  Do 500 simulations of this stochastic process with $\sigma = 1$.  Contrast this result to the deterministic solution.
\item Assume the recovery rate $r$ is stochastic.  Write down this stochastic differential equation.  Do 500 simulations of this stochastic process with $\sigma = 1$.  Contrast this result to the previous results.
\end{enumerate}
```

&nbsp;

```{exercise}
Consider the following Lotka-Volterra (predator prey) model:

  \begin{align*} 
\frac{dV}{dt} &= r V - kVP \\
\frac{dP}{dt} &= e k V P - dP 
\end{align*}

Assume that the parameter $k$ is stochastic.  Write down the stochastic differential equation, identifying the deterministic and stochastic parts to this system of equations.  Apply the Euler-Maruyama method (using `eulerStochastic`) for 100 simulations with $\sigma=0.01$ with the following values of parameters and step sizes:

  \begin{itemize}
  \item Initial condition: $V(0)=1$, $P(0)=3$
  \item Parameters: $r = 2$, $k = 0.5$, $e = 0.1$, and $d = 1$.
  \item Set $\Delta t = 0.05$ and $N=200$.
  \end{itemize}
```
 
&nbsp; 
 
```{exercise}
Consider the following model for zombie population dynamics:

\begin{align*}  
\frac{dS}{dt} &=-\beta S Z - \delta S  \\
\frac{dZ}{dt} &= \beta S Z + \xi R - \alpha SZ \\
\frac{dR}{dt} &= \delta S+ \alpha SZ  - \xi R
\end{align*}

\begin{enumerate}
\item Let's assume the transmission rate $\beta$ is a stochastic parameter.  With this assumption, group each differential equation into two parts: terms not involving noise (the deterministic part) and terms that are multiplied by noise (the stochastic part)

\begin{itemize}
\item  Deterministic part for $ \displaystyle \frac{dS}{dt}$: \hskip 2in Stochastic part for $\displaystyle \frac{dS}{dt}$: \hfill 
\item[]
\item Deterministic part for for $ \displaystyle \frac{dZ}{dt}$: \hskip 2in Stochastic part for $\displaystyle\frac{dZ}{dt}$: \hfill
\item[]
\item Deterministic part for for $ \displaystyle \frac{dR}{dt}$: \hskip 2inStochastic part for $\displaystyle \frac{dR}{dt}$: \hfill
\end{itemize}


\item Apply the Euler-Maruyama method to do 500 simulations of this stochastic differential equation.  How does making $\beta$ stochastic affect the disease transmission?  For this, please assume the following values of the parameters and Euler-Maruyama method:

\begin{itemize}
\item $\sigma = 0.004$
\item $\Delta t= 0.5$.
\item Timesteps: 200.
\item $\beta = 0.0095$, $\delta = 0.0001$ ,$\xi =  0.1$, $\alpha = 0.005$.
\item Initial condition: $S(0)=499$, $Z(0)=1$, $R(0)=0$.
\end{itemize}

\item If we assume that the population is fixed at 500 individuals, what is interesting about your stochastic results?

\end{enumerate}
```
