# Diffusion and Brownian Motion {#sec-diffusion-24}

Studying random walks in @sec-random-walks-23 led to some surprising results, namely that for an unbiased random walk the mean displacement was zero but the variance increased proportional to the step number. In this chapter we will revisit the random walk problem from another perspective that further strengthens its connection to understanding diffusion. Let's get started!

::: {.callout-note}
This chapter uses following libraries: `ggplot2`, `tibble`, and `purrr` (all part of the `tidyverse`) and `demodelr`. See @sec-packages-02 on how to install packages.  Prior to running any code in this section, include `library(tidyverse)` and `library(demodelr)`. 
:::

## Random walk redux
The random walk derivation in @sec-random-walks-23 focused on the *position* of a particle on the random walk, based upon prescribed rules of moving to the left and the right.\index{random walk} To revisit this random walk we consider the *probability* (between 0 and 1) that a particle is at position $x$ in time $t$, denoted as $p(x,t)$.\index{probability distribution!stochastic differential equation solution} In other words, rather than focusing on where the particle *is*, we focus on the *chance* that the particle will be at a given spot.

A way to conceptualize a random walk is that any given position $x$, a particle can arrive to that position from either the left or the right (@fig-random-walk-24):


```{tikz,label=fig-random-walk-24,warning=FALSE,message=FALSE,echo=FALSE,fig.cap="Schematic diagram for the one-dimensional random walk."}
\begin{tikzpicture}
    \draw (-3,0) -- (3,0);
    \foreach \i in {-3,-2,...,3} % numbers on line
      \draw (\i,0.1) -- + (0,-0.2) node[below] (\i) {$\i$};
   % \foreach \i in {0.5, 0.7, 0.9}% points on line
    \fill[red] (0,0) circle (1 mm);
     \draw[red] (-1,0) circle (1 mm);
     \draw[red] (1,0) circle (1 mm);
   \node[align=center] at (-1.1,0.5) {$p(-1,t)$};
    \draw [<-,red,thick] (-.05,0.15) to [out=150,in=30] (-1,0.15);
    \node[align=center] at (1.1,0.5) {$p(1,t)$};
       \node[align=center] at (1.9,-1) {$\displaystyle p(0,t+\Delta t)=\frac{1}{2} p(-1,t)+\frac{1}{2} p(1,t)$};

    \draw [<-,red,thick] (0.05,0.15) to [out=30,in=150] (1,0.15);
\end{tikzpicture}
```

We can generalize @fig-random-walk-24 further where the particle moves in increments $\Delta x$, as defined in @eq-master-diff:

$$
p(x,t+\Delta t) = \frac{1}{2} p(x-\Delta x,t) + \frac{1}{2} p(x+\Delta x,t) 
$$ {#eq-master-diff}

To analyze @eq-master-diff we apply Taylor approximations on each side of @eq-master-diff.\index{Taylor approximation}  First let’s do a locally linear approximation for $p(x,t+\Delta t)$:

$$
p(x,t+\Delta t)  \approx p(x,t) + \Delta t \cdot p_{t},
$$

where we have dropped the shorthand $p_{t}(x,t)$ as $p_{t}$. On the right hand side of @eq-master-diff we will compute the 2nd degree (quadratic) Taylor polynomial:

$$
\begin{align*}
\frac{1}{2} p(x-\Delta x,t)  & \approx \frac{1}{2} p(x,t) -   \frac{1}{2} \Delta x \cdot p_{x} + \frac{1}{4} (\Delta x)^{2}\cdot  p_{xx} \\
\frac{1}{2} p(x+\Delta x,t)  & \approx \frac{1}{2} p(x,t) +   \frac{1}{2} \Delta x \cdot p_{x} + \frac{1}{4} (\Delta x)^{2} \cdot p_{xx}
\end{align*}
$$
With these approximations we can re-write @eq-master-diff as @eq-new-master-diff:

$$
\Delta t \cdot p_{t} = \frac{1}{2} (\Delta x)^{2} p_{xx} \rightarrow  p_{t} = \frac{1}{2} \frac{(\Delta x)^{2}}{\Delta t} \cdot p_{xx} 
$$ {#eq-new-master-diff}

@eq-new-master-diff is called a partial differential equation - what this means is that it is a differential equation with derivatives that depend on two variables ($x$ and $t$ (two derivatives).\index{partial differential equation} As studied in @sec-random-walks-23, @eq-new-master-diff is called the **diffusion equation**.\index{diffusion} In @eq-new-master-diff we can also define $\displaystyle D =  \frac{1}{2} \frac{(\Delta x)^{2}}{\Delta t}$ so $p_{t}=D \cdot p_{xx}$.

The solution to @eq-new-master-diff is given by @eq-diff-eq-soln.^[Determining an exact solution to the diffusion equation requires more study in techniques of partial differential equations (see @keener_biology_2021).]

$$
p(x,t) = \frac{1}{\sqrt{4 \pi Dt} } e^{-x^{2}/(4 D t)}
$$  {#eq-diff-eq-soln}



What @eq-diff-eq-soln represents is the probability that the particle is at the position $x$ at time $t$. @fig-diffusion-profile shows *profiles* for $p(x,t)$ when $D=0.5$ at different values of $t$.


```{r,echo=FALSE,fig.cap="Profiles of $p(x,t)$ (Equation \\@ref(eq:diff-eq-soln)) for different values of $t$ with $D = 0.5$."}
#| label: fig-diffusion-profile

# sigma = sqrt(2*D*t) set #D = 1/2
x <- seq(-10, 10, length.out = 100)
y1 <- dnorm(x, mean = 0, sd = 1) # t = 1
y2 <- dnorm(x, mean = 0, sd = sqrt(2)) # t = 2
y3 <- dnorm(x, mean = 0, sd = 4) # t = 16
y4 <- dnorm(x, mean = 0, sd = sqrt(32)) # t = 32

data.frame(x, y1, y2, y3, y4) |>
  gather(key = run, value = y, -1) |>
  ggplot(aes(x = x, y = y, color = as.factor(run))) +
  geom_line(size = 1) +
  theme(legend.position = "bottom") +
  ylab("p(x,t)") +
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 14),
    axis.title.x = element_text(size = 14),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.y = element_text(size = 14)
  ) +
  scale_color_colorblind(name = NULL, labels = c("t = 1", "t = 2", "t = 16", "t = 32"))
```

As you can see, as time increases the graph of $p(x,t)$ gets flatter - or more uniform. What this tells you is that the longer $t$ increases it is less likely to find the particle at the origin.

### Verifying the solution to the diffusion equation
Verifying that @eq-diff-eq-soln is the solution to @eq-new-master-diff is a good review of your multivariable calculus skills!  As a first step to verifying this solution, let's take the partial derivative with respect to $x$ and $t$.

First we will compute the partial derivative of $p$ with respect to the variable $x$ (represented as $p_{x}$):

$$
\begin{align*}
p_{x} &= \frac{\partial }{\partial x} \left( \frac{1}{\sqrt{4 \pi Dt} } e^{-x^{2}/(4 D t)} \right) \\
&= \frac{1}{\sqrt{4 \pi Dt} } e^{-x^{2}/(4 D t)} \cdot \frac{-2x}{4Dt}
\end{align*}
$$

Notice something interesting here:  $\displaystyle p_{x} = p(x,t) \cdot \left( \frac{-x}{2Dt} \right)$.

To compute the second derivative, we have the following expressions by applying the product rule:

$$
\begin{align*}
p_{xx} &= p_{x} \cdot \left( \frac{-x}{2Dt} \right) - p(x,t) \cdot \left( \frac{1}{2Dt} \right) \\
&= p(x,t) \cdot \left( \frac{-x}{2Dt} \right) \cdot \left( \frac{-x}{2Dt} \right)- p(x,t) \cdot \left( \frac{1}{2Dt} \right) \\
&= p(x,t) \left( \left( \frac{-x}{2Dt} \right)^{2} - \left( \frac{1}{2Dt} \right) \right) \\
&= p(x,t) \left( \frac{x^{2}-2Dt}{(2Dt)^{2}}\right).
\end{align*}
$$

So far so good. Now computing $p_{t}$ gets a little tricky because this derivative involves both the product rule with the chain rule in two places (the variable $t$ appears twice in the formula for $p(x,t)$).\index{chain rule}\index{product rule}  To aid in computing the derivative we identify two functions $\displaystyle f(t) = (4 \pi D t)^{-1/2}$ and $\displaystyle g(t) = -x^{2} \cdot (4Dt)^{-1}$. This changes $p(x,t)$ into $p(x,t) = f(t) \cdot e^{g(t)}$. In this way $p_{t} = f'(t) \cdot e^{g(t)} + f(t) \cdot e^{g(t)} \cdot g'(t)$. Now we can focus on computing the individual derivatives $f'(t)$ and $g'(t)$ (after simplification - be sure to verify these on your own!):

$$
\begin{align*}
f'(t) &= -\frac{1}{2} (4 \pi D t)^{-3/2} \cdot 4 \pi D = -2\pi D \; (4 \pi D t)^{-3/2}  \\
g'(t) &= x^{2}\; (4Dt)^{-2} 4D = \frac{x^{2}}{4Dt^{2}}
\end{align*}
$$

Assembling these results together, we have the following:

$$
\begin{align*}
p_{t} &= f'(t) \cdot e^{g(t)} + f(t) \cdot e^{g(t)} \cdot g'(t) \\
&= -2\pi D \; (4 \pi D t)^{-3/2} \cdot e^{-x^{2}/(4 D t)}  + \frac{1}{\sqrt{4 \pi Dt} } \cdot e^{-x^{2}/(4 D t)} \cdot \frac{x^{2}}{4Dt^{2}} \\
&= \frac{1}{\sqrt{4 \pi Dt} } \cdot e^{-x^{2}/(4 D t)} \left( -2 \pi D \; (4 \pi D t)^{-1} +  \frac{x^{2}}{4Dt^{2}} \right) \\
&= \frac{1}{\sqrt{4 \pi Dt} } \cdot e^{-x^{2}/(4 D t)} \left( -\frac{1}{2t} +  \frac{x^{2}}{4Dt^{2}} \right) \\
&= p(x,t)  \left( -\frac{1}{2t} +  \frac{x^{2}}{4Dt^{2}} \right)
\end{align*}
$$

Wow. Verifying that @eq-diff-eq-soln is a solution to the diffusion equation is getting complicated, but also notice that through algebraic simplification, $\displaystyle p_{t} = p(x,t)  \left(\frac{x^{2}-2Dt}{4Dt^{2}} \right)$. When we compare $p_{t}$ to $D p_{xx}$, they are equal!


 The connections between diffusion and probability are so strong. @eq-diff-eq-soln is related to the formula for a normal probability density function (@eq-normal-09 from @sec-likelihood-09)! In this case, the standard deviation in @eq-diff-eq-soln equals $\sqrt{2Dt}$ and is time dependent (see @exr-normal-compare)). Even though we approached the random walk differently here compared to @sec-random-walks-23, we also saw that the variance grew proportional to the time spent, so there is some consistency.

## Simulating Brownian motion
Another name for the process of a particle undergoing small random movements is *Brownian Motion*.\index{Brownian motion}  We can simulate Brownian motion similar to the random walk as discussed in @sec-random-walks-23. Brownian motion is connected to the diffusion equation (@eq-new-master-diff) and its solution (@eq-diff-eq-soln). These connections are helpful when simulating stochastic differential equations. To simulate Brownian motion we will also apply the workflow from @sec-stoch-sim-22 (Do once $\rightarrow$ Do several times $\rightarrow$ Summarize $\rightarrow$ Visualize).

### Do once
First we define a function called `brownian_motion` that will compute a sample path given the:

- number of steps to run the stochastic process;
- diffusion coefficient $D$;
- timestep $\Delta t$;

a sample path will be computed (see @fig-sample-brownian).

```{r, fig.cap="Sample trajectory for a realization of Brownian motion. The horizontal axis represents a step of the random walk."}
#| label: fig-sample-brownian

brownian_motion <- function(number_steps, D, deltaT) {
  # D: diffusion coefficient
  # deltaT: timestep length
  ### Set up vector of results
  x <- array(0, dim = number_steps)

  for (i in 2:number_steps) {
    x[i] <- x[i - 1] + sqrt(2 * D * deltaT) * rnorm(1)
  }

  out_x <- tibble(t = 0:(number_steps - 1), x)
  return(out_x)
}

# Run a sample trajectory and plot
try1 <- brownian_motion(100, 0.5, 0.1)

plot(try1, type = "l")
```

### Do several times
Once we have the function for Brownian motion defined we can then run this process several times and plot the spaghetti plot (try the following code out on your own):

```{r eval = FALSE}

number_steps <- 200 # Then number of steps in random walk
D <- 0.5 # The value of the diffusion coefficient
dt <- 0.1 # The timestep length

n_sims <- 500 # The number of simulations

# Compute solutions
brownian_motion_sim <- map(1:n_sims, ~ list()) |>
  set_names(paste0("sim", 1:n_sims)) |>
  map(~ brownian_motion(number_steps, D, dt)) |>
  map_dfr(~.x, .id = "simulation")

# Plot these all together
ggplot(data = brownian_motion_sim, aes(x = t, y = x)) +
  geom_line(aes(color = simulation)) +
  ggtitle("Random Walk") +
  guides(color = "none")
```

### Summarize and visualize
Finally, the 95% confidence interval is computed and plotted in @fig-ensemble-ave-24, using similar code from @sec-stoch-sim-22 to compute the ensemble average. Note that the horizontal axis is time so each step is scaled by `dt`.


```{r, warning=FALSE,message=FALSE,fig.cap="Ensemble average of 500 simulations for the random walk. Each step on the horizontal axis is scaled by `dt`.",echo=FALSE}
#| label: fig-ensemble-ave-24

number_steps <- 200 # Then number of steps in random walk
D <- 0.5 # The value of the diffusion coefficient
dt <- 0.1 # The timestep length

n_sims <- 500 # The number of simulations

# Compute solutions
brownian_motion_sim <- map(1:n_sims, ~ list()) |>
  set_names(paste0("sim", 1:n_sims)) |>
  map(~ brownian_motion(number_steps, D, dt)) |>
  map_dfr(~.x, .id = "simulation")

# Compute Quantiles and summarize
quantile_vals <- c(0.025, 0.5, 0.975)

brownian_motion_quantile <- brownian_motion_sim |>
  group_by(t) |>
  summarize(
    q_val = quantile(x, # x is the column to compute the quantiles
      probs = quantile_vals
    ),
    q_name = quantile_vals
  ) |>
  pivot_wider(
    names_from = "q_name", values_from = "q_val",
    names_glue = "q{q_name}"
  )

# Plot Ensemble Average
ggplot(data = brownian_motion_quantile) +
  geom_line(aes(x = t*dt, y = q0.5),
    color = "red", size = 1, inherit.aes = FALSE
  ) +
  geom_ribbon(aes(x = t*dt, ymin = q0.025, ymax = q0.975),
    alpha = 0.2, fill = "red", inherit.aes = FALSE
  ) +
  guides(color = "none") +
  labs(x = "Time", y = "Ensemble average") +
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 14),
    axis.title.x = element_text(size = 14),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.y = element_text(size = 14)
  )
```

I sure hope the results are very similar to ones generated in @sec-random-walks-23 (especially @fig-ensemble-ave-23) - this is no coincidence! With the ideas of a random walk developed here and in @sec-random-walks-23, we will now be able to understand and simulate how small changes in a variable or parameter affect the solutions to a differential equation. Looking ahead to [Chapters @sec-sdes-25 and @sec-simul-stoch-26], we will simulate stochastic processes using numerical methods (Euler's method in @sec-euler-04) with Brownian motion. Onward!



## Exercises

::: {#exr-direct-24}
Through direct computation, verify the following calculations:

a. When $\displaystyle f(t)=\frac{1}{\sqrt{4 \pi Dt} }$, then $\displaystyle f'(t)=-2\pi D (4 \pi D t)^{-3/2}$
b. When $\displaystyle g(t)=\frac{-x^{2}}{4Dt}$, then $\displaystyle g'(t)=\frac{x^{2}}{4Dt^{2}}$
c. Verify that $\displaystyle \left( -\frac{1}{2t} +  \frac{x^{2}}{4Dt^{2}} \right)= \left( \frac{x^{2}-2Dt}{(2Dt)^{2}}\right)$

:::

::: {#exr-normal-compare}
The equation for the normal distribution is $\displaystyle f(x)=\frac{1}{\sqrt{2 \pi} \sigma } e^{-(x-\mu)^{2}/(2 \sigma^{2})}$, with mean $\mu$ and variance $\sigma^{2}$. Examine the formula for the diffusion equation (@eq-diff-eq-soln) and compare it to the formula for the normal distribution. If @eq-diff-eq-soln represents a normal distribution, what do $\mu$ and $\sigma^{2}$ equal?
:::




::: {#exr-int-normal-24}
For this problem you will investigate $p(x,t)$ (@eq-diff-eq-soln) with $\displaystyle D=\frac{1}{2}$.

a. Evaluate $\displaystyle \int_{-1}^{1} p(x,10) \; dx$. Write a one sentence description of what this quantity represents.
b. Using desmos or some other numerical integrator, complete the following table:

| **Equation** | **Result**  |
|:------:|:-----:|
|$\displaystyle \int_{-1}^{1} p(x,10) \; dx=$ |  |
|$\displaystyle \int_{-1}^{1} p(x,5) \; dx=$ |  |
|$\displaystyle \int_{-1}^{1} p(x,2.5) \; dx=$ |  |
|$\displaystyle \int_{-1}^{1} p(x,1) \; dx=$ |  |
|$\displaystyle \int_{-1}^{1} p(x,0.1) \; dx=$ |  |
|$\displaystyle \int_{-1}^{1} p(x,0.01) \; dx=$ |  |
|$\displaystyle \int_{-1}^{1} p(x,0.001) \; dx=$ |  |

c. Based on the evidence from your table, what would you say is the value of $\displaystyle \lim_{t \rightarrow 0^{+}} \int_{-1}^{1} p(x,t) \; dx$?
d. Now make graphs of $p(x,t)$ at each of the values of $t$ in your table. What would you say is occuring in the graph as $\displaystyle \lim_{t \rightarrow 0^{+}} p(x,t)$? Does anything surprise you? (The results you computed here lead to the foundation of what is called the Dirac delta function.)

:::

::: {#exr-int-normal-reduc-24}
{{< fa burst >}} Numerical integration can also be done using `R` and the command `integrate`.  Consider the following code that defines a function `pxt`, which is a shorthand for $p(x,t)$ in @eq-diff-eq-soln:

```{r,eval=FALSE}
pxt = function(x, t, D = 1) {
	exp(-x^2 / (4*D*t)) / sqrt(pi*4*D*t)
}

integrate(pxt, lower = -1, upper=1, t = 1, D = 1/2)

```

Let $D=1/2$ and $t=1$ and use `R` to calculate the following integrals:

| **Equation** | **Result**  |
|:------:|:-----:|
|$\displaystyle \int_{-0.1}^{0.1} p(x,10) \; dx=$ |  |
|$\displaystyle \int_{-1}^{1} p(x,5) \; dx=$ |  |
|$\displaystyle \int_{-10}^{10} p(x,2.5) \; dx=$ |  |
|$\displaystyle \int_{-100}^{100} p(x,1) \; dx=$ |  |

What do you notice as the lower and upper limits of integration increase?

:::

::: {#exr-explain-normal-graph-24}
Consider the function $\displaystyle  p(x,t) = \frac{1}{\sqrt{4 \pi D t}}  e^{-x^{2}/(4 D t)}$. Let $x=1$.


a. Explain in your own words what the graph $p(1,t)$ represents as a function of $t$.
b. Graph several profiles of $p(1,t)$ when $D = 1$, $2$, and $0.1$. How does the value of $D$ affect the profile?

:::

::: {#exr-conf-int-24}
In statistics an approximation for the 95% confidence interval is twice the standard deviation. Confirm this by adding the curve $y=2\sqrt{2Dt}$ to the ensemble average plot in Figure 24.4. Recall that $D$ was equal to $0.5$ and $\Delta t = 0.1$, so the horizontal axis will need to be scaled appropriately.
:::

::: {#exr-pd-24}
Consider the function $\displaystyle  p(x,t) = \frac{1}{\sqrt{\pi t}} e^{-x^{2}/t}$:


a. Using your differentiation skills compute the partial derivatives $p_{t}$, $p_{x}$, and $p_{xx}$.
b. Verify $p(x,t)$ is consistent with the diffusion equation $\displaystyle p_{t}=\frac{1}{4} p_{xx}$.

:::

::: {#exr-variation-D-24}
Modify the code used to generate @fig-ensemble-ave-24 with $D=10, \; 1, \; 0.1, \; 0.01$. Generally speaking, what happens to the resulting ensemble average when $D$ is small or large? In which scenarios are stochastic effects more prevalent?
:::

::: {#exr-advec-24}
For the one-dimensional random walk we discussed where there was an equal chance of moving to the left or the right. Here is a variation on this problem.

Let's assume there is a chance $v$ that it moves to the left (position $x - \Delta x$), and therefore a chance is $1-v$ that the particle remains at position $x$. The basic equation that describes the particle's position at position $x$ and time $t + \Delta t$ is:

$$
p(x,t + \Delta t) = (1-v) \cdot p(x,t) + v \cdot p(x- \Delta x,t)
$$


Apply the techniques of local linearization in $x$ and $t$ to show that this random walk is used to derive the following partial differential equation, called the *advection equation*:

$$
p_{t} = - \left( v \cdot \frac{ \Delta x}{\Delta t} \right) \cdot p_{x}
$$

*Note: you only need to expand this equation to first order*
:::

::: {#exr-bivariate-24}
Complete @exr-random-2d-23 if you haven't already. If @eq-diff-eq-soln (a normal distribution) is the solution to the one-dimensional diffusion equation, what do you think the solution would be in the bivariate case?
:::
