# Solving Stochastic Differential Equations

Previously we found stochastic differential equations arose from modeling contexts. In this section we will examine stochastic differential equations in general. For this final section we will examine how we can characterize solutions to stochastic differential equations.  From our work with random walks and simulating stochastic differential equations, solutions to a stochastic differential equation have to be understood in terms of probability density functions, which can evolve in both time and any state variables.

This section leads us into some powerful mathematics that hopefully you will want to study at some point in the future!

## Meet the Fokker-Planck Equation
But the generic form to a stochastic differential equation is the following:

\begin{equation}
dx = a(x,t) \; dt + b(x,t) \; dW(t)
\end{equation}

The "solution" to this SDE will be a probability density function $f(x,t)$. 
Our goal is to have a function that describes the evolution of $f(x,t)$ in both time and space. Based on our work with birth-death processes, the probability density function $f(x,t)$ should have the following properties:

- $E[f(x,t)]$ is the deterministic solution to $\displaystyle \frac{dx}{dt} = a(x,t)$.
- Var$[f(x,t)]$ is proportional to $b(x,t)$.


We determine the probability density function through solution of the following differential eqution:

\begin{equation}
\frac{\partial f}{\partial t} = − \frac{\partial}{\partial x} \left(f(x,t) \cdot a(x,t) \right) + \frac{1}{2}\frac{\partial^{2} }{\partial x^{2}} \left(f(x,t) b^{2}(x,t) \right)
\end{equation}

We can write this equation in shorthand, dropping the dependence of $x$ and $t$ for $f(x,t)$, $a(x,t)$ and $b(x,t)$: $\displaystyle f_{t} = −  (f \cdot a)_{x} + \frac{1}{2} (f \cdot b^{2})_{xx}$.

```{example}
Consider the SDE $dx = dW(t)$ and apply the Fokker-Planck equation to characterize the solution $f(x,t)$.
```

In this case $a(x,t)=0$ and $b(x,t)=1$, so the Fokker-Planck Equation is:

\begin{equation*}
f_{t} = \frac{\sigma^{2}}{2} f_{xx}.
\end{equation*}

This equation should look familiar - it is the partial differential equation for diffusion.  We know that the solution is $\displaystyle f(x,t) = \frac{1}{\sqrt{2 \pi \sigma^{2} t}} e^{−x^{2}/(2 \sigma^{2} t)}$



The initial condition $f(x,0)=\delta(x)$, which is called the $\delta$ function, which is a special type of probability density function called the [Dirac delta Function](https://en.wikipedia.org/wiki/Dirac_delta_function).  The evolution of this solution is shown in the following figure:


```{r,echo=FALSE,fig.width=4,fig.height=3,fig.align='center'}
# sigma = sqrt(2*D*t) set #D = 1/2
x <- seq(-10,10,length.out=100)
y1 <- dnorm(x,mean=0,sd=1) # t = 1 
y2 <- dnorm(x,mean=0,sd=sqrt(2)) # t = 2
y3 <- dnorm(x,mean=0,sd=4) # t = 16
y4 <- dnorm(x,mean=0,sd=sqrt(32)) # t = 32 

data.frame(x,y1,y2,y3,y4) %>%
  gather(key=run,value=y,-1) %>%
  ggplot(aes(x=x,y=y,color=as.factor(run))) + geom_line(size=2) +
  theme(legend.position="bottom") +
  scale_color_discrete(name=NULL,labels=c("t = 1", "t = 2","t = 16","t = 32")) +
  ylab('p(x,t)')


```

```{example}
Consider the SDE $dx = r \; dt + \sigma \; dW(t)$, where r and $\sigma$ are constants. Apply the Fokker-Planck equation to characterize the solution $f(x,t)$.
```

In this case the Fokker-Planck equation is $\displaystyle f_{t} =−r f_{x}+ \frac{\sigma^{2}}{2} f_{xx}$.  This is an example of a *diffusion-advection equation*.   Amazingly this equation can be reduced to a diffusion equation through a change of variables and application of the multivariable change of variables. Let $z=x-rt$. Then we plan to write $f(x,t)=f(z)$.


```{r,engine='tikz',warning=FALSE,message=FALSE,echo=FALSE,fig.cap='Multivariable chain rule'}
\begin{center}

\begin{tikzpicture}[auto, outer sep=1pt, node distance=1.5cm]
\node (A) {$f$};
\node [below of = A, left of = A] (B) {$z$} ;
\node [below of = A, right of = A] (C) {$\tau$} ;
%\node [below of = C] (D) {$t$};
\node [below of = B, left of = B] (E) {$x$} ;
\node [below of = B, right of = B] (F) {$t$} ;

\draw [-] (A) -- (B);
\draw [-] (B) -- (E);
\draw [-] (B) -- (F);
\draw [-] (A) -- (C);
\draw [-] (C) -- (F);
%\draw [-] (A) -- node[left, yshift=10pt] {$\displaystyle \frac{\partial z}{\partial x}$} (B);

%\draw [-] (A) -- node[right, yshift=10pt] {$\displaystyle \frac{\partial z}{\partial t}$} (C);


\end{tikzpicture}

\end{center}
```


Our goal is to develop expressions for $f_{t}$ $f_{x}$, and $f_{xx}$ with this change of variables. Note that since $z=x-r \tau$ we have $z_{\tau} = -r$ and $z_{x} = 1$. Also since $\tau=t$, $\tau_{t}=1$.  We have:

\begin{align*}
\frac{\partial f}{\partial t} & = \frac{\partial f}{\partial \tau} \cdot \frac{ \partial \tau}{\partial t} + \frac{\partial f}{\partial z} \cdot \frac{ \partial z}{\partial t}  = \frac{\partial f}{\partial \tau} -r \frac{\partial f}{\partial z} \\
\frac{\partial f}{\partial x} & = \frac{\partial f}{\partial z} \cdot \frac{ \partial z}{\partial x} = \frac{\partial f}{\partial z}
\end{align*}


Next we need to compute $\displaystyle \frac{\partial^{2} f}{\partial z^{2}}$, or $f_{zz}$.  Since $f_{z}=f_{x}$, then $f_{zz} = f_{xx}$.  So if we re-write our equation from the variables $x$ and $t$ to $z$ and $\tau$ we have:

\begin{align*}
f_{t} &=−r f_{x}+ \frac{\sigma^{2}}{2} f_{xx} \\
f_{\tau} - r f_{z} &= -r f_{z} + \frac{\sigma^{2}}{2} f_{zz} \\
f_{\tau} &=  \frac{\sigma^{2}}{2} f_{zz}
\end{align*}

This is exciting because with this change of variables we can characterize the solution: $\displaystyle f(z, \tau) = \frac{1}{\sqrt{2 \pi \sigma^{2} \tau}} e^{−z^{2}/(2 \sigma^{2} \tau)}$, and then transform back into the original variables $x$ and $t$:

\begin{equation}
f(x, t) = \frac{1}{\sqrt{2 \pi \sigma^{2} t}} e^{−(x-rt)^{2}/(2 \sigma^{2} t)}
\end{equation}

Based on what we know of this distribution is that it should look like a normal distribution with mean $\mu = rt$ and variance $\sigma^{2} t$.  So what this tells us that the mean is shifting and growing more diffuse as time increases.  Let's take a look at some representative plots:


```{r,echo=FALSE,fig.width=4,fig.height=3,fig.align='center'}
# sigma = sqrt(2*D*t) set #D = 1/2
r <- 0.1
x <- seq(-10,10,length.out=100)
y1 <- dnorm(x,mean=r,sd=1) # t = 1 
y2 <- dnorm(x,mean=r*2,sd=sqrt(2)) # t = 2
y3 <- dnorm(x,mean=r*16,sd=4) # t = 16
y4 <- dnorm(x,mean=r*32,sd=sqrt(32)) # t = 32 

data.frame(x,y1,y2,y3,y4) %>%
  gather(key=run,value=y,-1) %>%
  ggplot(aes(x=x,y=y,color=as.factor(run))) + geom_line(size=2) +
  theme(legend.position="bottom") +
  scale_color_discrete(name=NULL,labels=c("t = 1", "t = 2","t = 16","t = 32")) +
  ylab('f(x,t)')


```

Notice the differences in this solution: as $t$ increases the solution shifts to the right and grows more diffuse. The differential equation is an example of a *diffusion-advection* equation, or a solution that drifts as time increases.

These are just a few examples of how to build a deeper understanding of stochastic processes and differential equations.  There is a lot of power in understanding the theoretical distribution for some test cases.  The additional study of stochastic differential equations extends this out further, with some interesting mathematics!


\newpage

## Exercises
<!-- Based off LW pg 346 #3 -->
```{exercise}
Let $R(t)$ denote the rainfall at a location at time $t$, which is a random process.  Assume that probability of the change in rainfall from day $t$ to day $t+\Delta t$ is the following:

\begin{center}

\begin{tabular}
{ c c}
change & probability \\ \hline \hline
$\Delta R = \rho$ & $\lambda \Delta t$ \\
$\Delta R = 0$ & $1- \lambda \Delta t$ \\ \hline
\end {tabular}

\end{center}

The stochastic differential equation generated by this process is $dR = \lambda \rho \; dt + \sqrt{\lambda \rho^{2}} \; dW(t)$.

\begin{enumerate}
\item What is the Fokker-Planck partial differential equation for the probability distribution $f(R,t)$?
\item What is a formula that solves the Fokker-Planck partial differential equation?
\item Make some representative plots of the solution as it evolves over time
\end{enumerate}
```

&nbsp;

<!-- Based off Example 7.24 pg 355 LW -->
```{exercise}
Consider the differential equation $\displaystyle x' = \lambda x - c \mu x^{2}$, which is similar to a logistic differential equation.  The *per capita* rate equation for this differential equation is $\displaystyle \frac{x'}{x} = \lambda - c \mu x$.

\begin{enumerate}
\item Assume there is noise to this per capita rate, i.e. $\displaystyle \frac{x'}{x} \rightarrow \displaystyle \frac{x'}{x} + \mbox{ Noise}$.  With this revised equation, what are the deterministic and stochastic parts?
\item What is the Fokker-Planck partial differential equation for the probability distribution $f(x,t)$?
\item The \emph{steady-state distribution} assumes that $f_{t} = 0$, so $f(x,t) \rightarrow f(x)$.  Through direct verification, show that $f(x)=Dx^{\alpha-1} e^{-\beta x}$ is a solution to the steady-state distribution, where $\alpha = 2 \lambda - 1$ and $\beta = 2c\mu$.
\item With $\lambda = c = \mu = D = 1$, make a graph of $f(x)$ and graph it below:
\end{enumerate}
```
&nbsp;

```{exercise}
Models of cell membranes take account for the energy needed for ions and other materials to cross the cell membrane, usually expressed as a membrane potential $U(x)$, where $x$ is the current position of a particle distance.  The probability $p$ of the particle being at position $x$ at time $t$  is given by the Fokker-Planck equation:
  
  \begin{equation}
v \frac{\partial p}{\partial t} = \frac{\partial}{\partial x} \left( U'(x) p \right) + k T \frac{\partial^{2} p}{\partial x^{2}},
\end{equation}

where $k$ is Boltzmann's constant and $T$ is the temperature.

\begin{enumerate}
\item Write the Fokker-Planck equation in steady state.
\item Show that a solution for the steady-state equation is $p(x)=C e^{-\frac{U(x)}{kT}}$
\item If the steady-state distribution is normal, what is function will $U(x)$ be?
\end{enumerate}
```
