# Simulating Stochastic Dynamics {#simul-stoch-26}

In Section \@ref(sdes-25) we built up a stochastic differential equation from adding in stochasticity (noise) to the parameter values.  Another approach is to assume instead the variables are stochastic.  Here is the good news: we can still simulate SDEs with stochastic variables, setting them up in a similar way to Section \@ref(sdes-25). We will generate realizations and then compute the ensemble averages.


## The stochastic logistic model redux
Let’s go back to the logistic population model but re-written in a specific way:

\begin{equation}
\frac{dx}{dt} = r x \left( 1 - \frac{x}{K} \right) = r x - \frac{rx^{2}}{K} (\#eq:log-sde-26)
\end{equation}

From Equation \@ref(eq:log-sde-26), we can obtain a change in the variable $x$ (denoted as $\Delta x$) over $\Delta t$ units by re-writing the differential equation in differential form:

\begin{equation}
\Delta x = r x \Delta t - \frac{rx^{2}}{K} \Delta t (\#eq:log-sde-diff-26)
\end{equation}

$\displaystyle \Delta x = r x \Delta t - \frac{rx^{2}}{K} \Delta t$.  Equation \@ref(eq:log-sde-diff-26) is separated into two terms - one that increases the variable $x$ (represented by $r x \Delta t$, same units as $x$) and one that decreases the variable (represented by $\displaystyle \frac{rx^{2}}{K} \Delta t$, same units as $x$). We will consider these changes as on a unit scale, organized via the following table:

Outcome | Probability | 
------------- | ------------- | 
$\Delta x = 1$ (population change by 1) | $r x \; \Delta t$ |
$\Delta x = -1$ (population change by -1) | $\displaystyle \frac{rx^{2}}{K} \; \Delta t$ |
$\Delta x = 0$ (no population change) | $\displaystyle 1 - rx \; \Delta t - \frac{rx^{2}}{K} \; \Delta t$ |

It also may be helpful to think of these changes on a random walk number line:

```{tikz,random-walk-stoch,warning=FALSE,message=FALSE,echo=FALSE,fig.width=4,fig.height=3}
\begin{tikzpicture}
    \draw (-3,0) -- (3,0);
    \foreach \i in {-3,-2,...,3} % numbers on line
      \draw (\i,0.1) -- + (0,-0.2) node[below] (\i) {$\i$};
   % \foreach \i in {0.5, 0.7, 0.9}% points on line
    \fill[red] (0,0) circle (1 mm);
   \node[align=center] at (-1,0.5) {$\displaystyle \frac{rx^{2}}{K} \Delta t$};
    \draw [->,red,thick] (0,0.15) to [out=150,in=30] (-1,0.15);
    \node[align=center] at (1,0.5) {$r x \Delta t$};
    \draw [->,red,thick] (0,0.15) to [out=30,in=150] (1,0.15);
  \end{tikzpicture}
```


It may seem odd to think of the different outcomes ($\Delta x$ equals 1, -1, or 1) as probabilities. Part of the reason why that formulation is useful is to apply concepts from probability theory. Let $Y$ be a random variable with a finite number of finite outcomes $\displaystyle y_{1},y_{2},\ldots ,y_{k}$ with probabilities $\displaystyle p_{1},p_{2},\ldots ,p_{k}$,  respectively, then the expected value $\mu$ of $Y$ is:

\begin{equation*}
\mu = E[Y] = \sum_{i=1}^{k} y_{i}\,p_{i}=y_{1}p_{1}+y_{2}p_{2}+\cdots +y_{k}p_{k}.
\end{equation*}

If we apply this definition to the random variable $\Delta x$ we have:

\begin{equation}
\begin{split}
\mu = E[\Delta x] &= (1) \cdot \mbox{Pr}(\Delta x = 1) + (- 1) \cdot \mbox{Pr}(\Delta x = -1)  + (0) \cdot \mbox{Pr}(\Delta x = 0) \\
&= (1) \cdot \left( r x \; \Delta t \right) + (-1)  \frac{rx^{2}}{K} \; \Delta t \\
&= r x \; \Delta t - \frac{rx^{2}}{K} \; \Delta t
\end{split} (\#eq:log-sde-mu-26)
\end{equation}

Notice that the right hand side of Equation \@ref(eq:log-sde-mu-26) is the same as the right hand side of the original differential equation (Equation \@ref(eq:log-sde-diff-26)))!

Next let's also calculate the variance of $\Delta x$, defined for a discrete random variable:

\begin{equation*}
 \sigma^{2} = E[(Y - \mu)^{2}] = \sigma^{2}= E[Y^{2}] - (E[Y] )^{2}.
\end{equation*}



\begin{equation}
\begin{split}
\sigma^{2} = E[(\Delta x)^{2}] - (E[\Delta x] )^{2} &= (1)^{2} \cdot \mbox{Pr}(\Delta x = 1) + (- 1)^{2} \cdot \mbox{Pr}(\Delta x = -1)  + (0)^{2} \cdot \mbox{Pr}(\Delta x = 0) - (E[\Delta x] )^{2} \\
&= (1) \cdot \left( r x \; \Delta t \right) + (1)  \frac{rx^{2}}{K} \; \Delta t  - \left( r x \; \Delta t - \frac{rx^{2}}{K} \; \Delta t \right)^{2}
\end{split}
\end{equation}

Along with the computation of the variance,  We are going to compute the variance to first order in $\Delta t$.  Because of that, we are going to assume that in $\sigma^{2}$ any terms involving $(\Delta t)^{2}$ are small, or in effect negligible.  While this is a huge simplifying assumption for the variance, but it is useful!

Doing these calcuations we have that $\displaystyle \sigma^{2} = \left( r x \; \Delta t \right) +  \frac{rx^{2}}{K} \; \Delta t$. Computing the mean and variance will help characterize the ensemble average.  In addition the following random walk properties can be applied:

- Since $\Delta x$ is the sum of many smaller changes we can apply the [central limit theorem](https://en.wikipedia.org/wiki/Central_limit_theorem) to characterize $\Delta x$ as a normal random variable.
- To first order, we assume that $\Delta x$ follows a probability distribution that is normal with mean $\mu$ and variance $\sigma^{2}$.
- The distribution for $\Delta x$ can be expressed as $\Delta x = \mu + \sigma Z$, where Z is random variable from a unit normal distribution (so in `R` we would use `rnorm(1)`).
- We can simulate $\Delta x$ as a Wiener process.
- Since $\Delta x = x_{n+1}-x_{n}$, then we have $x_{n+1} = x_{n} + \mu + \sigma Z$.


Notice how the last step provides a way to generate a solution trajectory to our differential equation.  Cool!  To simulate this stochastic process we will use the function `birth_death_stochastic`, which is set up in a similar way to `euler_stochastic` from Section \@ref(sdes-25).  I will append `_log` to denote "logistic" for each of the parts


```{r log-sde-26,fig.cap="One realization of the logistic differential equation with stochastic variables.",fig.width=4,fig.height=3}

# Identify the birth and death parts of the DE:
birth_rate_log <- c(dx ~ r*x)
death_rate_log <-  c(dx ~ r*x^2/K)

# Identify the initial condition and any parameters
init_log <- c(x=3)  # Be sure you have enough conditions as you do variables. 
parameters_log <- c(r=0.8, K=100)   # parameters: a named vector

# Identify how long we run the simulation
deltaT_log <- .05    # timestep length
time_steps_log <- 200   # must be a number greater than 1

# Identify the standard deviation of the stochastic noise
sigma_log <- 1

# Do one simulation of this differential equation
out_log <- birth_death_stochastic(birth_rate = birth_rate_log,
                             death_rate = death_rate_log,
                             init_cond = init_log,
                             parameters = parameters_log,
                             deltaT = deltaT_log,
                             n_steps = time_steps_log,
                             sigma = sigma_log)

# Plot out the solution
ggplot(data = out_log) +
  geom_line(aes(x=t,y=x))
```

Notice how the resulting spaghetti plot shows variation in the variables.

Making an ensemble average plot is also similar to how we computed them in Section \@ref(sdes-25):

```{r spaghetti-r-26,fig.cap="Several different realizations of the logistic SDE with stochasticity in the variables, along with the ensemble average plot.",fig.width=4,fig.height=3}
# Many solutions
n_sims <- 100  # The number of simulations

# Compute solutions
logistic_sim_r <- rerun(n_sims) %>%
  set_names(paste0("sim", 1:n_sims)) %>%
  map(~ birth_death_stochastic(birth_rate = birth_rate_log,
                             death_rate = death_rate_log,
                             init_cond = init_log,
                             parameters = parameters_log,
                             deltaT = deltaT_log,
                             n_steps = time_steps_log,
                             sigma = sigma_log)
) %>%
  map_dfr(~ .x, .id = "simulation")


# Plot these all up together
  ggplot(data = logistic_sim_r) +
  geom_line(aes(x=t, y=x, color = simulation)) +
  ggtitle("Spaghetti plot for the logistic SDE") +
  guides(color="none")

  
  ### Summarize the variables
summarized_logistic_sim_r <- logistic_sim_r %>%
  group_by(t) %>%  # All simulations will be grouped at the same timepoint.
  summarise(x = quantile(x, c(0.25, 0.5, 0.75)), q = c("q0.025", "q0.5", "q0.975")) %>%
  pivot_wider(names_from = q, values_from = x)


### Make the plot
ggplot(data = summarized_logistic_sim_r) +
  geom_line(aes(x = t, y = q0.5)) +
  geom_ribbon(aes(x=t,ymin=q0.025,ymax=q0.975),alpha=0.2) +
  ggtitle("Ensemble average plot for the logistic SDE")

```




Notice that as before the spaghetti and ensemble average plots are generated.  While there are some simulations that may always be zero, in the *ensemble* the median resembles the (deterministic) solution to the logistic differential equation.

The types of stochastic processes we are describing in this section are called "birth-death" processes. Here is another way to think about Equation \@ref(eq:log-sde-diff-26):

\begin{equation}
\begin{split}
r x \; \Delta t   &= \alpha(x)  \mbox{ (birth) }\\
\frac{rx^{2}}{K} \; \Delta t &= \delta(x) \mbox{ (death) }
\end{split} (\#eq:birth-death-log-26)
\end{equation}

In this way we think of $\alpha(x)$ in Equation \@ref(eq:birth-death-log-26) as a "birth process" and $\delta(x)$ as a "death process". When we computed the mean $\mu$ and variance $\sigma^{2}$ for Equation \@ref(eq:log-sde-diff-26) we had $\mu = \alpha(x)-\delta(x)$ and $\sigma^{2}=\alpha(x)+\delta(x)$ (you should double check to make sure this is the case!). Here is an interesting fact: $\mu = \alpha(x)-\delta(x)$ and $\sigma^{2}=\alpha(x)+\delta(x)$ holds up for *any* differential equation where we have identified a birth or death process.



## A stochastic system of equations
Let's examine a stochastic system of differential equations.  Here we will return to the lynx-hare model:

\begin{equation}
\begin{split}
\frac{dH}{dt} &= r H - b HL \\
\frac{dL}{dt} &=ebHL -dL
\end{split}
\end{equation}

In this case we still split *each equation* into the birth ($\alpha$) and death ($\delta$) parts:
\begin{align*}
\alpha & =
\begin{cases}
\frac{dH}{dt}: & r H \\
\frac{dL}{dt}: & ebHL\\
\end{cases} \\
\delta & =
\begin{cases}
\frac{dH}{dt}: & bHL \\
\frac{dL}{dt}: & dL\\
\end{cases} 
\end{align*}


To simulate this stochastic process, the setup of the code is similar to previous ways we solved systems of differential equations.  Since we have a system of equations to compute the expected value and variance requires additional knowledge of matrix algebra, but that is already included in the function `birth_death_stochastic`

```{r stoch-pred-prey-26,fig.width=4,fig.height=5,fig.cap="One realization of the stochastic predator-prey model.",fig.width=4,fig.height=3}
# Identify the birth and death parts of the DE:
birth_rate_pred <- c(dH ~ r*H, dL ~ e*b*H*L)
death_rate_pred <-  c(dH ~ b*H*L, dL ~ d*L)

# Identify the initial condition and any parameters
init_pred <- c(H=1, L=3)  # Be sure you have enough conditions as you do variables

# Identify the parameters
parameters_pred <- c(r = 2, b = 0.5, e = 0.1, d = 1)

# Identify how long we run the simulation
deltaT_pred <- .05    # timestep length
time_steps_pred <- 200   # must be a number greater than 1

# Identify the standard deviation of the stochastic noise
sigma_pred <- .1


# Do one simulation of this differential equation
out_pred <- birth_death_stochastic(birth_rate = birth_rate_pred,
                                       death_rate = death_rate_pred,
                                       init_cond = init_pred,
                                       parameters = parameters_pred,
                                       deltaT = deltaT_pred,
                                       n_steps = time_steps_pred,
                                       sigma = sigma_pred)

# Visualize the solution
ggplot(data = out_pred) +
  geom_line(aes(x=t,y=H),color='red') +
  geom_line(aes(x=t,y=L),color='blue') +
labs(x='Time',
     y='Lynx (red) or Hares (blue)')

```

Excellent!  Notice how Figure \@ref(fig:stoch-pred-prey-26) has stochasticity in the variables for *both* *H* and *L*. The next step would be to do many simulations and then compute the ensemble average.  First let's compute the individual simulations:

```{r}
# Many solutions
n_sims <- 100  # The number of simulations

# Compute solutions
pred_sim <- rerun(n_sims) %>%
  set_names(paste0("sim", 1:n_sims)) %>%
  map(~ birth_death_stochastic(birth_rate = birth_rate_pred,
                                       death_rate = death_rate_pred,
                                       init_cond = init_pred,
                                       parameters = parameters_pred,
                                       deltaT = deltaT_pred,
                                       n_steps = time_steps_pred,
                                       sigma = sigma_pred)
) %>%
  map_dfr(~ .x, .id = "simulation")

```

The above code may take a while to complete - but this is certainly shorter than computing these all by hand!  In order to make the spaghetti plot, we will plot the variables $H$ and $L$ separately.

```{r spaghetti-pred-26,fig.cap="Several different realizations of the stochastic predator-prey system.",fig.width=4,fig.height=3}

# Plot the hares first:
ggplot(data = pred_sim) +
  geom_line(aes(x=t,y=H,color=simulation)) +
  ggtitle("Spaghetti plot for hares in the predator-prey SDE") +
  guides(color="none")

# Then plot the lynx:
ggplot(data = pred_sim) +
  geom_line(aes(x=t,y=L,color=simulation)) +
  ggtitle("Spaghetti plot for lynx in the predator-prey SDE") +
  guides(color="none")

```

For the ensemble average plots, we will first utilize the `pivot_longer` command, group and then summarize:

```{r}
 ### Summarize the variables
summarized_pred_sim <- pred_sim %>%
  pivot_longer(cols=c("H","L")) %>%
  group_by(name,t) %>%  # All simulations will be grouped at the same timepoint.
  summarise(value = quantile(value, c(0.25, 0.5, 0.75)), q = c("q0.025", "q0.5", "q0.975")) %>%
  pivot_wider(names_from = q, values_from = value)

# Let's take a look at the resulting data frame
glimpse(summarized_pred_sim)
```

Let's break this down:

- The data frame `pred_sim` has columns `simulation`, `t`, `H`, and `L`. We want to group *each* variable by the time `t`. In order to do that efficiently we need to pivot the data frame longer.  By applying the command `pivot_longer(cols=c("H","L"))` we will still have 4 columns, but this time they are called `simulation`, `t`, `name`, and `value`.  The column `name` is a categorical variable that is either `H` or `L` (the columns we made longer), and the column `value` has the numerical values for both at each time.
- We then group by the column `name` first, and then by `time`.
- The remainder of the code is similar to how we computed the ensemble average above.

The tricky part is that resulting columns of `summarized_pred_sim` are `name`, `t`, and the quantile values. This may seem a challenge to plot, but we can easily make a small multiples plot with the command `facet_grid`:

```{r ensemble-pred-26,fig.cap="Ensemble average plot of the predator-prey system.",fig.width=4,fig.height=3}
ggplot(data = summarized_pred_sim) +
  geom_line(aes(x = t, y = q0.5)) +
  geom_ribbon(aes(x=t,ymin=q0.025,ymax=q0.975),alpha=0.2) +
  facet_grid(. ~ name)
```

The last command `facet_grid(. ~ name)` splits the plot up into different panels (or facets) by the column `name`.  So easy to make!

If you are feeling a little overwhelmed by all these new plotting commands - don't worry. You can easily modify these examples for a different systme of differential equations. You've got this!




## Generalizing the approach.
As mentioned previously, another way to think about the logistic differential equation as what as known as a "birth-death process".

Let’s call the part of the differential equation that contributes to a positive rate as a "birth process" and parts that contribute to a negative rate as a "death process".  If we have the differential equation

\begin{equation*}
\frac{dx}{dt} = \alpha(x)-\delta(x)
\end{equation*}

Then we would simulate the birth death process with $\mu = \alpha(x)-\delta(x)$ and $\sigma^{2} = \alpha(x)+\delta(x)$.

For a multivariable system of equations the process is the same, however because we have a system of equations the calculations for the expected value and variance require more knowledge of matrix algebra which is beyond the scope here.  The provided code does take this into account.




\newpage

## Exercises

```{exercise}
Return back to one simulation of the logistic differential equation. (Figure \@ref(fig:log-sde-26)). For this plot we set $\sigma = 1$. What happens to the resulting spaghetti and ensemble plots when $\sigma =0, \;  0.01, \; .1, \; 10$?
```
&nbsp;

```{exercise}
Return back to the example of one simulation of the stochastic predator prey  (Figure \@ref(fig:stoch-pred-prey-26)).  For these plots we set $\sigma = .1$. What happens to the resulting spaghetti and ensemble plots when $\sigma =0, \;  0.01, \; 1, \; 10$?
```
&nbsp;
```{exercise}
For the predatory-prey simulation of stochastic variables we used `group_by(name,t)` to begin summarizing our variables. Modify the code so you apply `group_by(t,name)` and generate the ensemble plot. Do you receive a similar result?
```
&nbsp;

```{exercise}
For the logistic differential equation consider the following splitting of $\alpha(x)$ and $\delta(x)$:

\begin{equation}
\begin{split}
\alpha(x) &= rx + \frac{rx^{2}}{2K} \\
\delta(x) &=  \frac{rx^{2}}{2K}
\end{split}
\end{equation}

Simulate this SDE using the same values of parameters for the logistic example and compare your results.

```

&nbsp;


<!-- LW pg 346 #3 -->
```{exercise}
(Inspired by @logan_mathematical_2009) Let $R(t)$ denote the rainfall at a location at time $t$, which is a random process.  Assume that probability of the change in rainfall from day $t$ to day $t+\Delta t$ is the following:

  | change | probability  | 
|:------:|:-----:|
| $\Delta R = \rho$ | $\lambda \Delta t$ |
| $\Delta R = 0$ | $1- \lambda \Delta t$ |



a. With this information, compute E[$\Delta R$] and variance of $\Delta R$.
b. Simulate this stochastic process.  Use $R(0)=0$ and run 500 simulations of this stochastic process.  Set $\lambda \rho = 18$ and $\sqrt{ \lambda \rho^{2}}=16$.

```

&nbsp;

```{exercise}
 Consider the following model for zombie population dynamics [@smith_mathematical_2014]:

\begin{equation}
\begin{split}
\frac{dS}{dt} &=-\beta S Z - \delta S  \\
\frac{dZ}{dt} &= \beta S Z + \xi R - \alpha SZ \\
\frac{dR}{dt} &= \delta S+ \alpha SZ  - \xi R
\end{split}
\end{equation}


a. Determine the birth and death terms for the zombie model so that you could encode this as a birth/death process:

- Birth part for $\displaystyle \frac{dS}{dt}$:
- Death part for $\displaystyle \frac{dS}{dt}$:
- Birth part for $\displaystyle \frac{dZ}{dt}$:
- Death part for $\displaystyle\frac{dZ}{dt}$:
- Birth part for $\displaystyle \frac{dR}{dt}$:
- Death part for $\displaystyle \frac{dR}{dt}$:


b. Use `birth_death_stochastic` to perform 500 simulations of this stochastic differential equation. Please assume the following values of the parameters and stochastic method:


- $\sigma = 0.004$
- $\Delta t= 0.5$.
- Timesteps: 200.
- $\beta = 0.0095$, $\delta = 0.0001$ ,$\xi =  0.1$, $\alpha = 0.005$.
- Initial condition: $S(0)=499$, $Z(0)=1$, $R(0)=0$.


```

&nbsp;

```{tikz,sis-bd,warning=FALSE,message=FALSE,echo=FALSE,fig.cap="The $SIS$ model"}

\tikzstyle{vspecies}=[rectangle,minimum size=0.5cm,draw=black]
\begin{tikzpicture}[auto, outer sep=1pt, node distance=2cm]

\node [vspecies] (S) {$S$} ;
\node [vspecies, right of = S] (I) {$I$} ;
\draw [->] ([yshift=3pt]S.east) --  node[above] {\small{$b$}} ([yshift=3pt]I.west) ;
\draw [<-] ([yshift=-3pt]S.east) --  node[below] {\small{$r$}} ([yshift=-3pt]I.west) ;
\end{tikzpicture}



```
&nbsp;
<!-- Ornstein uhlenbeck process -->
```{exercise}
Consider the stochastic differential equation $\displaystyle dS = \left( 1 - S \right) + \sigma dW(t)$, where $\sigma$ controls the amount of stochastic noise.  For this stochastic differential equation what is $E[S]$ and Var$(S)$?
```
&nbsp;


<!-- Adapted from LW pg 346 #4 -->
```{exercise}
(Inspired by @logan_mathematical_2009) An $SIS$ model is one where susceptibles $S$ become infected $I$, and then after recovering from an illness, become susceptible again.  The schematic representing this is shown in Figure \@ref(fig:sis-bd).  While you can write this as a system of differential equations, assuming the population size is constant $N$ we have the following differential equation:

\begin{equation}
\frac{dI}{dt} = b(N-I) I - r I
\end{equation}


a. Identify $\alpha(I)$ and $\delta(I)$ for this model.
b. Assuming $N=1000$, $r=0.01$, and $b=0.005$, $I(0)=1$, simulate this differential equation over two weeks with $\Delta t = 0.1$.  Show the plot of your result.

```

&nbsp;

```{exercise}
Consider the equation
\begin{equation*}
\Delta x = \alpha(x) \; \Delta t - \delta(x) \; \Delta t
\end{equation*}

If we consider $\Delta x$ to be a random variable, show that the expected value $\mu$ equals $\alpha(x) \; \Delta t - \delta(x) \; \Delta t$ and the variance $\sigma^{2}$, to first order, equals $\alpha(x) \; \Delta t + \delta(x) \; \Delta t$.
```


